---
title: "Multiple Treatment Trials and ANOVA"
subtitle: "AGRON 5130"
author: "Caio dos Santos"
output: pdf_document
---

# Introduction

The goal of this assignment is to help you connect
the Completely Randomized Design (CRD) to the
Analysis of Variance (ANOVA).

In this module, we moved from comparing two
treatments to analyzing experiments with multiple
treatments. Instead of running several $t$-tests,
we learned how ANOVA partitions total variation
into:

* Variation due to treatments (signal)
* Variation due to random error (noise)

Recall the linear additive model:

$$
Y_{ij} = \mu + \tau_i + \epsilon_{ij}
$$

Throughout this assignment, you will:

* Write linear models
* Conduct ANOVA tests
* Interpret results
* Compute and summarize treatment means

Submit a knitted PDF or HTML file generated from
your R Markdown document to Canvas.

**Note:** When writing models or hypotheses, you do
not need perfect mathematical notation. Plain text
is fine as long as your meaning is clear.


# Tillage and Soil Organic Matter

The effect of tillage on soil organic matter was
studied in a Polish trial. The three treatments
were:

* Conventional tillage (CT)
* Reduced tillage (RT)
* Strip-tillage (ST)

Organic matter is measured in percent.

This dataset is simulated but inspired by:

Debska, B.; Jaskulska, I.; Jaskulski, D.
Method of Tillage with the Factor Determining the
Quality of Organic Matter. Agronomy 2020, 10, 1250.


## Question 1

1 point

Read in the dataset located at
`data/soil_om_tillage_debska.csv`

Use the `head()` function to review the data.

```{r}
deb <- read.csv('./data/soil_om_tillage_debska.csv')
head(deb)
```


## Question 2

1 point

Write the linear additive model that explains the
observed organic matter value for an experimental
unit in this experiment.

Be sure to identify:

- The overall mean
- The treatment effect
- The random error term

Y_ij = mu + tau_i + epsilon_ij

## Question 3

2 points

Use the analysis of variance to test whether the
tillage treatment explains significantly more of
the observed variance among plots than the error
(residual) effect.

Remember, first you use the aov() function

```{r}
fit1 <- aov(om ~ tillage, data = deb)
```

then, you extrac the anova table using the
summary() function

```{r}
summary(fit1)
```


## Question 4

1 point

Based on your ANOVA results, does the tillage
treatment account for significantly more variation
in soil organic matter than random error? Briefly
justify your answer.



## Question 5

3 points

Create a data table containing the tillage
treatment means using the aggregate() function

Display the resulting table.

```{r}
trt_means <- aggregate(deb['om'],
                       deb['tillage'],
                       mean)
trt_means
```


# Light Color and Kale Biomass

A controlled-environment agriculture study was
conducted to test the effect of white, purple,
and pink light on Chinese kale biomass.

Biomass is measured in grams per plant.

This dataset is simulated but inspired by:

Zhang, Y.; Ji, J.; Song, S.; Su, W.; Liu, H.
Growth, Nutritional Quality and Health-Promoting
Compounds in Chinese Kale Grown under Different
Ratios of Red:Blue LED Lights. Agronomy 2020, 10,
1248.


## Question 6

1 point

Load the dataset located at
`data/led_kale_zhang.csv`

Display the top six rows.

```{r}
kale <- read.csv('./data/led_kale_zhang.csv')
head(kale)
```


## Question 7

2 points

Use the analysis of variance to test whether light
color explains significantly more of the observed
variance in biomass than random error.

Remember, first you use the aov() function

```{r}
fit2 <- aov(biomass ~ light, data = kale)
```

then, you extrac the anova table using the
summary() function

```{r}
summary(fit2)
```


## Question 8

1 point

Based on the ANOVA results, what can you conclude
about the effect of light color on kale biomass?


## Question 9

3 points

Create a table of biomass means by light color
treatment means using the aggregate() function:

Display the resulting table.

```{r}
trt_means <- aggregate(kale['biomass'],
                       kale['light'],
                       mean)
trt_means
```


# Final Reflection

Remember that ANOVA does not tell us which means
differ, it tells us whether treatment variation
is large relative to background noise.

As you complete this assignment, think carefully
about:

* What source of variation you are testing
* How the F-statistic reflects signal vs. noise
* What your conclusion means biologically

