---
title: "Correlation and Simple Regression"
subtitle: "AGRON 5130 - Module 10"
author: "Caio dos Santos"
date: "2026-03-11"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    smooth: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Introduction

In this unit, we transition from categorical
comparisons to relationships between **quantitative
variables**.

We will cover two related tools:

- correlation (association)
- simple linear regression (prediction and causal
  modeling framework)

# Correlation vs Regression

Correlation answers:

> Do two variables move together?

Regression answers:

> How does Y change as X changes, and can we
predict Y from X?

Correlation does not establish causation.
Regression also does not prove causation by
itself, but it gives us a model framework when we
have a defensible direction from X to Y.

# Case Study: Cucumber Traits

```{r}
set.seed(20260314)
cucumber <- data.frame(
  branches = rpois(80, lambda = 6),
  leaves = NA,
  earlyfruit = NA,
  totalfruit = NA
)

cucumber$leaves <- 18 + 3.1 * cucumber$branches + rnorm(80, 0, 4)
cucumber$earlyfruit <- 2 + 0.7 * cucumber$branches + rnorm(80, 0, 1.5)
cucumber$totalfruit <- 8 + 1.2 * cucumber$branches + rnorm(80, 0, 2)

head(cucumber)
```

## Visual Inspection

```{r}
library(ggplot2)

ggplot(cucumber, aes(x = branches, y = leaves)) +
  geom_point(pch = 21, size = 2, fill = "grey70") +
  theme_linedraw()
```

```{r}
ggplot(cucumber, aes(x = branches, y = earlyfruit)) +
  geom_point(pch = 21, size = 2, fill = "grey70") +
  theme_linedraw()
```

## Correlation Coefficient

We use Pearson's r for linear association between
continuous variables.

```{r}
cor.test(cucumber$branches, cucumber$leaves)
cor.test(cucumber$branches, cucumber$earlyfruit)
```

You can also inspect a full correlation matrix:

```{r}
cor(cucumber)
```

# Simple Linear Regression

Now suppose our question is predictive:

> How much does fruit count change with branches?

Then we use simple linear regression.

## Model Form

$$Y_i = \alpha + \beta X_i + \varepsilon_i$$

Where:

- $\alpha$ is intercept
- $\beta$ is slope
- $\varepsilon_i$ is error

## Fit the Model

```{r}
model1 <- lm(totalfruit ~ branches, data = cucumber)
summary(model1)
```

Use ANOVA and broom outputs to interpret the model:

```{r}
anova(model1)
```

```{r}
library(broom)
tidy(model1)
glance(model1)
```

# Interpreting Coefficients

- Intercept (`alpha`): expected Y when X = 0
- Slope (`beta`): expected change in Y for a one-
  unit increase in X

The slope p-value tests whether the relationship is
statistically distinguishable from zero.

# Residual Diagnostics

Before trusting a model, inspect residuals.

```{r}
par(mfrow = c(1, 2))
plot(model1, which = 1)
plot(model1, which = 2)
par(mfrow = c(1, 1))
```

Look for:

- no strong curvature in residuals vs fitted
- no severe funnel shape
- approximate normality in Q-Q plot

# Visualization with Regression Line

```{r}
ggplot(cucumber, aes(x = branches, y = totalfruit)) +
  geom_point(pch = 21, size = 2, fill = "grey70") +
  geom_smooth(method = "lm", se = TRUE, color = "tomato") +
  theme_linedraw() +
  labs(x = "Branches", y = "Total Fruit")
```

# Prediction

```{r}
new_data <- data.frame(branches = c(3, 5, 7, 9, 11))
new_data$pred_totalfruit <- predict(model1, new_data)
new_data
```

Predictions should stay within the observed range
of X values whenever possible.

# Extrapolation

Extrapolation means predicting outside the data
range used to fit the model.

This is risky because model behavior beyond the
observed range may no longer be linear or valid.

# Conclusion

In this unit, we learned:

- correlation to quantify association
- simple linear regression to model Y from X
- diagnostics to assess model assumptions
- prediction and the limits of extrapolation

These tools set up the next unit, where we extend
regression to multiple predictors and nonlinear
relationships.
