---
title: "Unit 4 Assignment"
output: html_notebook
---


# Question 1a (1 pt) 
Load the dataset "data/cattle_wormer.csv" and examine its top six rows using the head() command.  Calves were divided into two populations, each treated with a different insecticide.  The cattle were paired according to weight, making this a paired experimental design.  Cattle were weighed 133 days after birth.
```{r}

```

# Question 1b (2 pt)
Use a t-test to test whether the difference in cattle weight between the two populations is equal to zero.
```{r}

```

# Question 1c (2 pt)
What does the p-value tell us about the probability of observing a difference of this amount by chance?


# Question 2a (3 pt)
Create a plot plan for a paired experimental design, with 5 replications, that compares two flea controls, "Itchy" and "Scratchy".  use the *design.rcbd()* function from the agricolae package.  Show the "book" view of the plot plan.
```{r}
library(agricolae)

```

# Question 2b (1 pt)
Show the plot map using the "sketch" view.
```{r}

```

# Question 3a (1pt)
Load and examine the "data/tobacco_radiation.csv" dataset.  Tobacco seeds were separated into two populations.  One population was untreated, while the other received a high dose of radiation.  Tobacco seeds were then planted in paired field plots.  Groups of 20 plants were measured, and their combinded heights (in cm) recorded.
```{r}

```

#Question 3b (2pt)
Test whether the height difference between the irradiated and non-irradiated populations is equal to zero
```{r}

```

# Question 3c (1 pt)
Is the difference significantly different from zero?

# Question 3d (2 pt)
Convert the data to "wide format" by using the *spread()* function of the "tidyverse" package to create individual columns for each treatment. 
```{r}

```

