--- 
title: Unit 7 Homework
output: html_document
---



# Question 1a (1 pt)
Fifteen canola entries were compared in a field trial in Mannetoba.  The experimental design was a randomized complete block design with three blocks.

Read in the data from "data/canola_data.csv" and print the first six rows using the *head()* function

```{r}
```

# Question 1b (2 pt)
Run the Analysis of Variance for entry and print the results using the *summary()* function.
```{r}

```

# Question 1c (1 pt)
What can you conclude about the effect of entry on canola yield?  (Don't read too much into this!  What does the p-value tell you?)


# Question 2 (1 pt)
Five barley genotypes were evaluated for their response to three levels of drought stress in an arid environment.  The experimenal design was a randomzed complete block with the two factors (genotype and drought stress).

Load the data from "data/barley_drought.csv" and print the top six rows using the head() command.
```{r}


```

# Question 2b (2 pt)
Conduct the Analysis of Variance and print the results using the *summary()* function.
```{r}


```

# Question 2c (1pt)
What does you initial look tell you about the effects of drought and genotype?  (Are they significant as main effects?  Is their interaction significant?)

# Question 2d (2pt)
Was every genotype affected by drought stress?  Test the drought response at each level of genotype and answer the question.
```{r}
library(tidyverse)
library(broom)

```


# Question 3a (1pt)
The effects of irrigation and tillage on soil nitrite (NO2) emission was studied in a split-plot design, with irrigation as the main factor and tillage as the sub-factor.  The data are loaded for you and saved to the data.frame named nitrite.  Print the top six rows of the dataset with head(). 
```{r}
nitrite = read.csv("data/irrigation_tillage_no2.csv")

```

# Question 3b (2 pt) 
Conduct an analysis of variance, keeping in mind this is a split-plot trial, and print the results using the summary() function.
```{r}


```

# Question 3c (2 pt)
Complete the code below to produce an interaction plot for tillage and irrgation. What does this tell you about their interaction?
```{r}
nitrite %>%
  group_by() %>%
  summarise(NO2 = ) %>%
  ungroup() %>%
  ggplot(aes(x=, y = , group = )) +
  geom_line(aes(color= ))

```

# Extra Credit -- this was a big lesson (2pts)
Create a randomized complete block design with 3 replicates and 3 potato varieties: "Big", "Head", and "Todd".  Don't eat them raw -- they're bittersweet.  Use seed_no=1993.

Report the treatment assignments (not the plot map).

```{r}
library(agricolae)
varieties = c("Big", "Head", "Todd")
reps = 3
seed_no =1993

```


