---
title: "Multiple Treatment Designs"
subtitle: "AGRON 5130"
author: "YOUR NAME HERE"
output: pdf_document
---

# Introduction

The goal of this assignment is to give you hands-on practice working
with **multiple treatment experimental designs** and their analyses. The
exercises build directly on the concepts introduced in this week's
lecture and lab.

You will analyze three agronomic experiments and practice:

-   identifying experimental designs\
-   fitting the correct statistical models\
-   interpreting main effects and interactions\
-   visualizing treatment responses

Submit a **knitted PDF generated from your R Markdown file** to Canvas.

# Question 1

**1 point**

Fifteen canola entries were compared in a field trial in Manitoba. The
experiment was conducted using a **randomized complete block design with
three blocks**. In this dataset block is coded as rep.

Load the dataset `"data/canola_data.csv"` and print the first six rows.

```{r}
canola <- read.csv('./data/canola_data.csv')
head(canola)
```

# Question 2

**2 points**

Fit an appropriate model to evaluate the effect of **canola entry on
yield** while accounting for block effects.

Print the ANOVA table for your model.

```{r}
model1 <- lm(yield ~ entry + rep,
             data = canola)
anova(model1)
```

# Question 3

**2 points**

Based on the ANOVA results from Question 2:

-   Is there evidence that **canola entry affects yield**?\
-   Why is it important to include **block** in the model?

Write 2–3 sentences.

# Question 4

**1 point**

Five barley genotypes were evaluated for their response to **three
levels of drought stress** in an arid environment. The experiment was
conducted as a **factorial treatment structure within a randomized
complete block design**.

Load the dataset `"data/barley_drought.csv"` and print the first six
rows.

```{r}
barley <- read.csv('./data/barley_drought.csv')
head(barley)
```

# Question 5

**2 points**

Fit a factorial model that includes:

-   block (you may treat it as fixed or random)
-   genotype\
-   drought\
-   genotype × drought interaction

Print the ANOVA table.

```{r}
model2 <- lm(yield ~ genotype * drought + block, 
             data = barley)
anova(model2)
```

# Question 6

**2 points**

Based on the ANOVA table from Question 5, interpret the results.

Address the following questions:

-   Is drought stress a significant factor affecting yield?\
-   Do genotypes differ in yield?\
-   Is the **genotype:drought interaction** significant?

# Question 7

**1 point**

In your own words, explain what a significant **genotype:drought**
interaction means in the context of this experiment.

# Question 8

**1 point**

The effects of **irrigation and tillage on soil nitrite (NO2 emission**
were studied using a **split-plot design**, where irrigation was applied
to main plots and tillage was applied to subplots.

Load the dataset `"data/irrigation_tillage_no2.csv"` and assign it to an
object named `nitrite`. Print the first six rows.

```{r}
nitrate <- read.csv('./data/irrigation_tillage_no2.csv')
head(nitrate)
```

# Question 9

**2 points**

Fit an appropriate **split-plot model** for this experiment using a
mixed-effects model and print the ANOVA table.

*hint: there is a package that starts with `n` and ends with `lme` that
fits mixed-effects models.*

```{r}
library(nlme)
model3 <- lme(NO2 ~ tillage * irrigation,
              random = ~ 1 | block/irrigation, 
              data = nitrate)
anova(model3)

```

# Question 10

**1 point**

Based on the ANOVA table and the interaction plot, briefly explain
whether irrigation, tillage, or their interaction affects NO2 emissions.

# Question 11 (Bonus)

**2 points** 

*This one is optional*

In your own words, explain **why split-plot experiments require a
different statistical model than completely randomized or randomized
complete block designs.**
