---
title: "Population statistics"
subtitle: "AGRON 5130"
output: pdf_document
author: "Caio dos Santos"
---


# Question 1

Load the shapefile
"data/elkhart_ia_yield_map/elkhart_ia_yield_map.shp"
using the *st_read()* function, and assign it to
the data frame "soybean"

```{r}


```


# Question 2

Review the structure of the "soybean" data frame using the *head()* function.
```{r}

```

# Question 3

Plot the yield map using the *plot()* function.  Remember you need to designate the column to plot.
```{r}


```

# Question 4

Create a histogram of the yield using the *hist()* function.

```{r}

```

# Question 5

Calculate the mean yield.

```{r}

```

# Question 6
Calculate the median yield
```{r}

```

# Question 7

Calculate the standard deviation for yield.
```{r}

```



# Question 8

Load the planting map "data/moline_il_planting_map/moline_il_planting_map.shp".
```{r}

```

# Question 9
Inspect the data using the *head()* function.
```{r}


```

# Question 10
Map the column "AppldRt", which measures the actual planting rate.
```{r}

```

# Question 11
Create an histogram for the actual planting rate using the *hist()* function.
```{r}

```

#Question 12
What was the mean actual planting rate?
```{r}

```

# Question 13
What was the minimum actual rate planted?
```{r}

```

# Question 14
The standard deviation can be used to gauge the consistency of the planting rate.  What was the standard deviation of the actual planting rate?
```{r}

```


