---
title: "Non-normal yield data"
output: html_document
---

We will begin by importing the libraries we'll use in this exercise

```{r}
library(ggplot2)
library(emmeans)
```


## Skewed yield data

We'll look at an example coming from a strip-trial
that evaluated the effect of multiple foliar treatments on
soybean yield.

The experiment included 30 replications and five treatments:
Control, T1, T2, T3, and T4.

However, the data were stored in separate files, as can happen
in real-world settings where different teams manage different parts
of the experiment.

Each dataset is also in a **wide format**, meaning that each treatment
is stored in its own column.


Let's bring the data into R

First, we will bring the first part of the experiment and inspect it. 
It seems like this first file has the first three treatments. 


```{r}
df1 <- read.csv('./data/soybean-foliar-1.csv')
head(df1)
```
Great! Now, we can do the same for the second part of this trial's data.
This dataset seems to have two treatments but also seems to be out of order. 

```{r}
df2 <- read.csv('./data/soybean-foliar-2.csv')
head(df2)
```

Now, we will tell R to join these data frames into
a single one. One nice thing about using the
`merge()` function and telling it by which columns
to merge the data, is that it does not matter if
they are out of order. R will join them based on
the `rep` values.

```{r}
df <- merge(df1, 
            df2,
            by = 'rep')
head(df)
```


Ok! Now we have the dataset merged but we still
have to restructure it. As we discussed before,
the wide format works great on Excel but it's not
great to build models or graphs in R.

```{r}
df_long <- reshape(df,
                   direction = 'long',
                   varying = c('Yield.Control', 'Yield.T1', 'Yield.T2', 'Yield.T3', 'Yield.T4'),
                   v.names = 'yield',
                   timevar = 'treatment',
                   times = c('Control', 'T1', 'T2', 'T3', 'T4'),
                   idvar = 'rep'
                   )
head(df_long)
```


Great! Now that our dataset is in the right format
for our analysis, we can begin!

Like before, let's take a look at the yield data.
Empty circles represent the raw data, and filled
triangles with error bars represent the mean and
standard deviation.

```{r}
ggplot()+
  geom_point(data = df_long,
             aes(x = treatment, y = yield),
             pch = 1)+
  stat_summary(data = df_long,
               fun.data = 'mean_se',
               aes(x = treatment, y = yield),
               col = 'blue',
               pch = 17)+
  labs(y = 'Soybean yield (bu/ac)', x = 'Treatment')+
  theme_bw()
```

Let's take a look at the distribution of the raw
data. Yield data often deviate from normality,
especially showing right skew. Part of it is
because yield data is strictly positive. Can you
imagine harvesting -20 bu/ac?

Even then, we can note that the normal
distribution model below still attributes a small
probability of negative values happening. The data
below is skewed, with a few large values.

```{r}

xbar <- mean(df_long$yield)
s <- sd(df_long$yield)

ggplot()+
  geom_histogram(data = df_long,
                 aes(x = yield, y = ..density..), 
                 bins = 20,
                 col = 'black',
                 fill = 'lightgrey')+
  geom_line(aes(x = -10:120, y = dnorm(-10:120, xbar, s)), col = 'red')+
  labs(x = 'Soybean yield (bu/ac)', y = 'Density')+
  theme_bw()

```

### Fitting a linear model to this data

Just like with the count data, we can go ahead and
fit a linear model to diagnose whether any of our
assumptions are violated.


```{r, echo=TRUE}
mod_yield <- aov(yield ~ treatment, data = df_long)
summary(mod_yield)
```


Let's take a look at the same diagnostic plots
that we look at before. It seems like there's a
relationship between mean and variance, once
again. Also, the Q-Q plot shows us that the model
errors don't really fit a normal distribution,
indicating skewness.

```{r}
plot(mod_yield, which = 1)
plot(mod_yield, which = 2)
```

Once again, we are violating some model
assumptions. 

Before fixing things, I want to show some of the
implications of mis-specified models like this one.
Let's compute the treatment means and plot them
along with their confidence intervals.

First, we will extract the treatments means using
`emmeans`, then we will convert it to a data frame
so we can use it within `ggplot`

```{r}
means_yield <- emmeans(mod_yield, ~treatment)
means_yield <- data.frame(means_yield)
means_yield
```


Now, let's add this to a plot. A couple of things I want you to notice:

1. Since our model assumes constant variance among
the treatments, look at how the width of the
confidence interval is the same for all
treatments.

2. Does it seem like treatment `T3` has the same
variance as the others? 

3. Think of the residual plot we saw before.

```{r}
ggplot()+
  geom_point(data = df_long,
             aes(x = treatment, y = yield),
             pch = 1)+
  geom_errorbar(data = means_yield,
                aes(x = treatment, ymin = lower.CL, ymax = upper.CL),
                width = 0.1,
                col = 'red')+
  geom_point(data = means_yield,
             aes(x = treatment, y = emmean), col = 'red')+
  labs(y = 'Soybean yield (bu/ac)', x = 'Treatment')+
  theme_bw()
```



## Log transformation 

Let's fit the model using `log(yield)` as our
response variable and see if that helps with the
assumption of normality.

```{r}
mod_log_yield <- lm(log(yield) ~ treatment, data = df_long)
anova(mod_log_yield)
```

Looking at the diagnostic plots, it seems like
this helped quite a bit. Not only is the variance
more stable, the Q-Q plot also shows that the
model residuals are more in line with theoretical
quantiles.

```{r}
par(mfrow = c(1,2))
plot(mod_log_yield, which = 1)
plot(mod_log_yield, which = 2)
par(mfrow = c(1,1))
```


## Extracting and interpreting effects

Once we fit the log-transformed model, we can
recover the treatment means on the **original
scale** using `emmeans`.

```{r}
log_means <- emmeans(mod_log_yield, ~ treatment)
log_means <- data.frame(log_means)
log_means
```


Now that we've recovered the treatment means at the log scale, let's interpret them. 
Let's look at the difference between the `Control` and `T3` for example. 

```{r}
4.23 - 3.47
exp(0.76)
```

This difference of 0.76 exponentiated return a
value of 2.13. This means that, when
back-transformed to the data scale, the median of
treatment 3 is 2.13 times greater than the median of
the control.

Please note that I am talking about the median
once we back-transform the data. Back-transforming
gives something closer to a median than a mean,
due to the nonlinearity of the log transformation
(Jensen’s inequality).

We can verify that using the chunk below.

```{r}
ds_means_yield <- exp(log_means$emmean)
ds_means_yield
68.92 / 32.28
```

Now, we can see that log-transforming is great!
But remember: we changed the data to make the
model work. It solves several of our problems but
it also makes interpretations slightly more
complex.



## Fitting a GLM

Instead of transforming the data, we can fit a
**GLM** appropriate for skewed, strictly positive
data, like the **Gamma GLM** (common for yield
data):

```{r}
mod_gamma <- glm(yield ~ treatment,
                 family = Gamma(link = "log"),
                 data = df_long)
mod_gamma
plot(mod_gamma)
```


Let's extract the treatment means at the data scale. 
There are a couple of things I want to mention here:

1. The model coefficients are still in the log
scale. If we look at the coefficient above,
there's a difference of `0.73` between treatment
`T3` and the `Control`. When exponentiated, this
results in 2.07. This means that the mean of
treatment `T3` is about 2.07 times greater than
the `Control` mean.

2. We can easily convert between predictions and
means at the `link` (log) or `response` (data)
scales.

3. Note that the SE is different for every mean.
Now, we do not make the assumption that the errors
are normally distributed and with constant
variance anymore. The error variance is modeled as
part of the distribution.

```{r}
means_glm <- emmeans(mod_gamma, ~treatment, type = 'response')
means_glm <- data.frame(means_glm)
means_glm
```


Let's put these in a graph. Can you see that the
confidence intervals for treatment `T3` are wider
than for the `Control`, for example? This is
because when we modeled the data, we ackowledged
the data's distribution and that treatment `T3`
varied more than the `Control`.

```{r}
ggplot()+
  geom_point(data = df_long,
             aes(x = treatment, y = yield),
             pch = 1)+
  geom_errorbar(data = means_glm,
                aes(x = treatment, ymin = lower.CL, ymax = upper.CL),
                width = 0.1,
                col = 'red')+
  geom_point(data = means_glm,
             aes(x = treatment, y = response), col = 'red')+
  labs(y = 'Soybean yield (bu/ac)', x = 'Treatment')+
  theme_bw()
```



Notes for interpretation:

* Both approaches give similar treatment differences.
* Log-transformed LM: assumes normal errors after transformation.
* Gamma GLM: naturally handles skewed, positive data and models the mean-variance relationship.
* Ratios of yields can be interpreted directly from the log/Gamma models.

> Key takeaway: Transformations can help, but
using a model that matches the data distribution
often produces more reliable estimates and is
easier to interpret.

