---
title: "Non-normal proportion data"
output: html_document
---

We will begin by importing the libraries we'll use in this exercise

```{r}
library(ggplot2)
library(emmeans)
```

## Skewed proportion data

We'll look at an example coming from a field trial
that evaluated the effect of multiple foliar treatments
on the percent of a plot affected by a disease.

The experiment included 30 replications and five treatments:
Control, T1, T2, T3, and T4.

Like in real-world trials, the data may come from
different sources or teams. In this case, since
you had too many replications, you evaluated the
first 15 replications, and a co-worker evaluated
the last 15 replications. We now need to merge
these datasets.

Let's bring the data into R. First, the first part:

```{r}
df1 <- read.csv('./data/disease-proportion-1.csv')
head(df1)
```

Now the second part of the trial. Note that you
and your co-worker named the columns differently.
This is common in a research setting.

```{r}
df2 <- read.csv('./data/disease-proportion-2.csv')
head(df2)
```

We can merge these data frames vertically using
`rbind`. Note that R will not let us bind these
because they have different names.

```{r, eval=FALSE}
rbind(df1, df2)
```

We will rename the columns in `df2`, to match the columns in `df1`
```{r}
names(df2) <- names(df1)
head(df2)
```

Now, let's try again:
```{r}
df_long <- rbind(df1, df2)
head(df_long)
```


Great! Now that the dataset is ready, let's visualize it.

```{r}
ggplot()+
  geom_point(data = df_long,
             aes(x = treatment, y = disease),
             pch = 1)+
  stat_summary(data = df_long,
               fun.data = 'mean_se',
               aes(x = treatment, y = disease),
               col = 'blue',
               pch = 17)+
  labs(y = 'Percent disease', x = 'Treatment')+
  theme_bw()
```

### Distribution of the raw data

Proportion data often deviate from normality. They
are bounded between 0 and 100%, which creates
skewness, especially when values are near 0 or
100.

```{r}
xbar <- mean(df_long$disease)
s <- sd(df_long$disease)

ggplot()+
  geom_histogram(data = df_long,
                 aes(x = disease, y = ..density..),
                 bins = 20,
                 col = 'black',
                 fill = 'lightgrey')+
  geom_line(aes(x = seq(-10, 100,1), 
                y = dnorm(seq(-10,100,1), xbar, s)), col = 'red')+
  labs(x = 'Percent disease', y = 'Density')+
  theme_bw()
```

### Fitting a linear model

```{r}
mod_prop <- aov(disease ~ treatment, data = df_long)
summary(mod_prop)
```

Let's check the model diagnostics:

```{r}
plot(mod_prop, which = 1) # residuals vs fitted
plot(mod_prop, which = 2) # Q-Q plot
```

Notice how residuals are not normally distributed. Variance increases with the mean—a classic issue with proportion data.





Before fixing things, I want to show some of the
implications of mis-specified models like this one.
Let's compute the treatment means and plot them
along with their confidence intervals.

First, we will extract the treatments means using
`emmeans`, then we will convert it to a data frame
so we can use it within `ggplot`

```{r}
means_prop <- emmeans(mod_prop, ~treatment)
means_prop <- data.frame(means_prop)
means_prop
```


Now, let's add this to a plot. A couple of things I want you to notice:

1. Since our model assumes constant variance among
the treatments, look at how the width of the
confidence interval is the same for all
treatments.

2. Does it seem like treatment `T3` has the same
variance as the others? 

3. Think of the residual plot we saw before.

```{r}
ggplot()+
  geom_point(data = df_long,
             aes(x = treatment, y = disease),
             pch = 1)+
  geom_errorbar(data = means_prop,
                aes(x = treatment, ymin = lower.CL, ymax = upper.CL),
                width = 0.1,
                col = 'red')+
  geom_point(data = means_prop,
             aes(x = treatment, y = emmean), col = 'red')+
  labs(y = 'Soybean yield (bu/ac)', x = 'Treatment')+
  theme_bw()
```

## Logit transformation

To stabilize variance and make residuals closer to normal, we can use a **logit transformation**. We'll add a small constant to avoid 0% or 100% values.

```{r}
df_long$logit_disease <- log((df_long$disease)/(100 - df_long$disease))

mod_logit <- lm(logit_disease ~ treatment, data = df_long)
anova(mod_logit)
```

Check diagnostics:

```{r}
par(mfrow = c(1,2))
plot(mod_logit, which = 1)
plot(mod_logit, which = 2)
par(mfrow = c(1,1))
```

---

## Extracting and interpreting effects

We can recover estimated treatment means on the **original scale** by back-transforming the logit:

```{r}
logit_means <- emmeans(mod_logit, ~ treatment)
logit_means <- data.frame(logit_means)
logit_means

# back-transform to percent
logit_means$percent <- 100 * exp(logit_means$emmean) / (1 + exp(logit_means$emmean))
logit_means
```

---

## Fitting a GLM

Instead of transforming, we can fit a **GLM** suitable for proportion data (Binomial family):

```{r}
# convert percent to proportion
df_long$prop_disease <- df_long$disease / 100

mod_glm <- glm(prop_disease ~ treatment,
               family = quasibinomial(link = "logit"),
               data = df_long)

summary(mod_glm)
plot(mod_glm)
```

Estimated means at the original scale:

```{r}
means_glm <- emmeans(mod_glm, ~ treatment, type = 'response')
means_glm <- data.frame(means_glm)
means_glm
```

Visualizing the results:

```{r}
ggplot()+
  geom_point(data = df_long,
             aes(x = treatment, y = prop_disease*100),
             pch = 1)+
  geom_errorbar(data = means_glm,
                aes(x = treatment, ymin = asymp.LCL*100, ymax = asymp.UCL*100),
                width = 0.1,
                col = 'red')+
  geom_point(data = means_glm,
             aes(x = treatment, y = prob*100), col = 'red')+
  labs(y = 'Percent disease', x = 'Treatment')+
  theme_bw()
```

---

### Notes for interpretation

* Logit-transformed LM: stabilizes variance, residuals closer to normal.
* Binomial/Quasibinomial GLM: naturally models proportions, allows variance to vary with mean.
* Back-transformed estimates can be interpreted as expected percent disease.

> Key takeaway: Transformations can help, but using a model that matches the data distribution is often more reliable and easier to interpret.



