---
title: "Exercise: RCBD from Start to Finish"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

In this exercise we will go through a complete
analysis of a **randomized complete block design
(RCBD)** with one treatment factor.

The goal is to walk through the entire workflow:

* loading the data
* visualizing the experiment
* fitting the ANOVA model
* separating treatment means
* testing agronomic contrasts
* presenting the results in tables and figures

By the end, you should have a clear picture of
what a full RCBD analysis looks like in practice.

# Before You Start: Install Required Packages

If you do not already have these packages
installed, run the command below once on your
machine. If a package is already installed, R
will simply skip it.

```{r eval=FALSE}
install.packages(c("emmeans", "multcomp"))
```

Now let's load the packages we will use in this
exercise.

```{r}
library(ggplot2)
library(emmeans)
library(multcomp)
```

# The Experiment

In this example, a field experiment evaluated four
fungicide programs for wheat under **powdery
mildew pressure**. The response variable is grain
yield (t/ha).

The treatments are:

* `T0`: untreated control
* `T1`: one fungicide application (early)
* `T2`: one fungicide application (late)
* `R`: recommended two-application program

The field had a **soil fertility gradient**, so
the
experiment was conducted using a **randomized
complete block design**. There are nine blocks,
and each block contains all four treatments.

This gives us **nine replicates per treatment**.

## Load the Data

Let's start by loading the dataset.

```{r}
mildew <- read.csv("data/mildew.csv")

mildew$block <- factor(mildew$block)
mildew$trt <- factor(mildew$trt, levels = c("T0", "T1", "R", "T2"))

head(mildew)
```

Notice that we convert `block` and `trt` to
factors. This tells R that these variables should
be treated as **categorical variables** in the
analysis.

## Visualize the Data

Before fitting any statistical model, it is always
a good idea to **look at the data**.

Here we build a simple figure showing the yield
distribution for each treatment.

```{r}
ggplot(mildew, aes(x = trt, y = yield)) +
        geom_boxplot(fill = "gray85", outlier.shape = NA) +
        geom_jitter(width = 0.12, size = 2, alpha = 0.8, color = "steelblue") +
        labs(
                x = "Fungicide Treatment",
                y = "Wheat Yield (t/ha)",
                title = "RCBD Data: Treatment Distributions"
        ) +
        theme_linedraw()
```

The boxplots summarize the treatment
distributions, while the points show the
individual observations. This gives us a quick
visual sense of how the treatments compare.

## Fit the RCBD Model

For a one-factor RCBD, the statistical model is

$$Y_{ij} = \mu + B_i + T_j + \varepsilon_{ij}$$

where

* (B_i) represents the block effect
* (T_j) represents the treatment effect

Now let's fit this model in R.

```{r}
model_rcbd <- aov(yield ~ block + trt, data = mildew)
```

## Interpret the ANOVA

Next, we examine the ANOVA table.

```{r}
anova(model_rcbd)
```

If the treatment *p-value* is small, this
provides evidence that **at least one treatment
mean differs from the others**.

When that happens, we proceed to **means
separation** to identify which treatments differ.

## Means Separation

We will use the **emmeans** package to compute
estimated marginal means and compare treatments.

```{r}
emm_trt <- emmeans(model_rcbd, ~ trt)

# Pairwise treatment comparisons with Tukey adjustment
pairs(emm_trt, adjust = "tukey")
```

This output shows the estimated difference
between each pair of treatments along with the
associated statistical test.

### LSD versus HSD

In `emmeans`, different multiple-comparison
procedures are controlled using the `adjust`
argument.

* `adjust = "none"` corresponds to an
  **LSD-style comparison**
* `adjust = "tukey"` corresponds to **Tukey's
  Honest Significant Difference**

Let's compute both.

```{r}
# LSD-style pairwise comparisons
pairs(emm_trt, adjust = "none")

# HSD (Tukey) pairwise comparisons
pairs(emm_trt, adjust = "tukey")
```

Another way to summarize these comparisons is
with **compact letter displays**.

Treatments sharing the same letter are **not
significantly different**.

Below we compare the groupings produced by LSD
and Tukey.

```{r}
letters_lsd <- as.data.frame(cld(emm_trt, adjust = "none", Letters = letters))
letters_hsd <- as.data.frame(cld(emm_trt, adjust = "tukey", Letters = letters))

lsd_hsd_compare <- data.frame(
        trt = as.character(letters_lsd$trt),
        LSD_group = trimws(letters_lsd$.group),
        HSD_group = trimws(letters_hsd$.group),
        stringsAsFactors = FALSE
)

lsd_hsd_compare
```

Here are the Tukey groupings alone:

```{r}
letters_trt <- as.data.frame(cld(emm_trt, adjust = "tukey", Letters = letters))
letters_trt
```

## Contrasts Between Groups of Treatments

Pairwise comparisons are useful, but many
agronomic questions involve **groups of
treatments** rather than individual ones.

Here we evaluate two planned contrasts:

* untreated control (`T0`) versus the **average
  of all fungicide programs**
* single-application programs (`T1`, `T2`)
  versus the **recommended two-application
  program (`R`)**

```{r}
contrast(emm_trt,
                                 method = list(
                                         "T0_vs_all_treated" = c(-1, 1/3, 1/3, 1/3),
                                         "single_vs_recommended" = c(0, 1/2, -1, 1/2)
                                 ),
                                 adjust = "none")
```

These are **planned contrasts based on
biological questions**, so we do not apply a
multiplicity adjustment here.

## Build a Means Table

Now let's build a simple reporting table with

* treatment means
* standard errors
* confidence intervals
* Tukey grouping letters

```{r}
means_tbl <- data.frame(
        as.data.frame(emm_trt)[, c("trt", "emmean", "SE", "lower.CL", "upper.CL")],
        stringsAsFactors = FALSE
)
colnames(means_tbl) <- c("trt", "mean", "SE", "lower", "upper")

means_tbl <- merge(means_tbl,
                   letters_trt[, c("trt", ".group")],
                   by = "trt")

means_tbl <- means_tbl[order(means_tbl$mean, decreasing = TRUE), ]
means_tbl$group <- trimws(means_tbl$.group)
means_tbl$.group <- NULL

means_tbl
```

This is the type of table you would commonly see
in an agronomy paper reporting treatment results.

## Build a Figure

Finally, let's create a figure showing

* treatment means
* confidence intervals
* Tukey grouping letters

```{r}
plot_tbl <- means_tbl[order(means_tbl$mean, decreasing = TRUE), ]

ggplot(plot_tbl, aes(x = reorder(trt, mean), y = mean)) +
        geom_col(fill = "steelblue", width = 0.65) +
        geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
        geom_text(aes(y = upper + 0.05, label = group), size = 5) +
        labs(
                x = "Fungicide Treatment",
                y = "Estimated Mean Yield (t/ha)",
                title = "RCBD Means with Tukey Grouping"
        ) +
        theme_linedraw()
```

This type of visualization combines the
statistical results with a clear graphical
representation of the treatment effects.

