---
title: "Exercise: RCBD"
output: html_document
---


In this exercise we will go through a complete
analysis of a **randomized complete block design
(RCBD)** with one treatment factor.

The goal is to walk through the entire workflow:

* loading the data
* visualizing the experiment
* fitting the ANOVA model
* separating treatment means
* testing agronomic contrasts
* presenting the results in tables and figures

By the end, you should have a clear picture of
what a full RCBD analysis looks like in practice.

# Before You Start: Install Required Packages

If you do not already have these packages
installed, run the command below once on your
machine. If a package is already installed, R
will simply skip it.

```{r eval=FALSE}
install.packages(c("emmeans", "multcomp", 'multcompView'))
```

Now let's load the packages we will use in this
exercise.

```{r}
library(ggplot2)
library(emmeans)
library(multcomp)
library(nlme)
```

# The Experiment

In this example, a field experiment evaluated four
fungicide programs for wheat under **powdery
mildew pressure**. The response variable is grain
yield (t/ha).

The treatments are:

* `T0`: untreated control
* `T1`: one fungicide application (early)
* `T2`: one fungicide application (late)
* `R`: recommended two-application program

The field had a **soil fertility gradient**, so
the
experiment was conducted using a **randomized
complete block design**. There are nine blocks,
and each block contains all four treatments.

This gives us **nine replicates per treatment**.

## Load the Data

Let's start by loading the dataset.

```{r}
mildew <- read.csv("data/mildew.csv")

mildew$block <- as.factor(mildew$block)
mildew$trt <- as.factor(mildew$trt)

head(mildew)
```

Notice that we convert `block` and `trt` to
factors. This tells R that these variables should
be treated as **categorical variables** in the
analysis.

## Visualize the Data

Before fitting any statistical model, it is always
a good idea to **look at the data**.

Here we build a simple figure showing the yield
distribution for each treatment.

```{r}
ggplot(mildew, aes(x = trt, y = yield)) +
  geom_jitter(width = 0.1, 
              size = 2, 
              alpha = 0.8, 
              color = "steelblue", 
              shape = 1) +
  labs(x = "Fungicide Treatment",
       y = "Wheat Yield (t/ha)") +
  stat_summary(fun.data = 'mean_se')+
  theme_bw()
```

## Fit the RCBD Model

For a one-factor RCBD, the statistical model is

$$Y_{ij} = \mu + b_i + T_j + \varepsilon_{ij}\newline
b_i \sim N(0, \sigma^2_{block})$$

where

* $b_i$ represents the block effect
* $T_j$ represents the treatment effect

Now let's fit this model in R.

```{r}
model1 <- lme(yield ~  trt,
                  random = ~ 1 | block,
                  data = mildew)
```

## Interpret the ANOVA

Next, we examine the ANOVA table.

```{r}
anova(model1)
```

If the treatment *p-value* is small, this
provides evidence that **at least one treatment
mean differs from the others**.

When that happens, we proceed to **means
separation** to identify which treatments differ.

## Means Separation

We will use the **emmeans** package to compute
estimated marginal means and compare treatments.

```{r}
emm <- emmeans(model1, ~ trt)

pairs(emm, adjust = "tukey")
```

This output shows the estimated difference
between each pair of treatments along with the
associated statistical test.

### LSD versus HSD

Here, we have 4 treatments. This means we are
running 6 comparisons.

$$4\times(4-1)/2 = 6$$

With 6 comparisons, our experiment-wise error rate
can be computed as:

```{r}
alpha <- 0.05
experiment_error_rate <- 1 - (1 - alpha)^6
experiment_error_rate
```

This means that we have a 26% chance of making at
least one error of type I (identifying a
difference when in reality there is none).


In `emmeans`, different multiple-comparison
procedures are controlled using the `adjust`
argument.

* `adjust = "none"` corresponds to an
  **LSD-style comparison**
* `adjust = "tukey"` corresponds to **Tukey's
  Honest Significant Difference**

Let's compute both.

LSD
```{r}
pairs(emm, adjust = "none")
```

Tukey
```{r}
pairs(emm, adjust = "tukey")
```

Another way to summarize these comparisons is
with **compact letter displays**.

Treatments sharing the same letter are **not
significantly different**.

Below we compare the groupings produced by LSD
and Tukey.

```{r}
cld(emm, adjust = "none", Letters = letters)
```
```{r}
cld(emm, adjust = "tukey", Letters = letters)
```


## Contrasts Between Groups of Treatments

Pairwise comparisons are useful, but many
agronomic questions involve **groups of
treatments** rather than individual ones.

Here we evaluate two planned contrasts:

* untreated control (`T0`) versus the **average
  of all fungicide programs**
* single-application programs (`T1`, `T2`)
  versus the **recommended two-application
  program (`R`)**

```{r}
contrast(emm,
                                 method = list(
                                         "T0_vs_all_treated" = c(-1, 1/3, 1/3, 1/3),
                                         "single_vs_recommended" = c(0, 1/2, -1, 1/2)
                                 ),
                                 adjust = "none")
```



## Build a Means Table

Now let's build a simple reporting table with

* treatment means
* standard errors
* confidence intervals
* Tukey grouping letters

First, let's store the groups of means in an object
```{r}
cld(emm, adjust = "tukey", Letters = letters)
cld(emm, adjust = "tukey", Letters = letters, decreasing = TRUE)
```

```{r}
groups <- cld(emm, adjust = "tukey", Letters = letters, decreasing = TRUE)
groups
```

Now that we have the `groups` data frame, we can
build a table. We will use a library called
`flextable`. Let's download the library and load
it.

```{r, eval=FALSE}
install.packages('flextable')
```

Let's build a first table and improve it
step-by-step. This first version looks good
already, as the `cld()` function output is
generally well formated. Let's see if we can
Improve it.

```{r}
library(flextable)
groups <- data.frame(groups)
tab <- flextable(groups)
tab
```


Let's start by formatting the number of decimal
places in the numeric columns

```{r}
tab <- colformat_double(tab,
                        j = c("emmean","SE","lower.CL","upper.CL"),
                        digits = 2)

tab
```

Now, I think we can improve this by changing the
header labels

```{r}
tab <- set_header_labels(tab,
                         values = c('Treatment', 'Mean', 'SE', 'DF', 'Lower CL', 'Upper CL', 'Group'))
tab
```



We can also add a caption for the table:
```{r}
tab <- set_caption(tab, 
                   caption = as_paragraph(as_b('Table 1:'),'Mean wheat yield for four fungicide treatments randomized complete block design.'))
tab
```

Adding a footnote about the procedure used to separate means is a good idea too:

```{r}
tab <- footnote(tab, 
                j = 7, 
                part = 'header',
         value = as_paragraph('Letters indicate mean separation groups using a Tukey procedure at α = 0.05; treatments sharing a common letter are not significantly different.'))
tab
```



Finally, we can use the `autofit()` function to adjust columns widths

```{r}
tab <- autofit(tab)
tab
```

Then, we can export this as an image or as a Word
document.

```{r}
save_as_image(tab, path = 'wheat_tab.png')
save_as_docx(tab, path = 'wheat_tab.docx')
```


## Build a Figure

Finally, let's create a figure showing

* treatment means
* confidence intervals
* Tukey grouping letters

```{r}

ggplot(groups) +
  geom_col(aes(x = trt, y = emmean), fill = "grey50", col = 'black') +
  geom_errorbar(aes(x = trt, ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_text(aes(x = trt, y = upper.CL + 0.5, label = .group), size = 5) +
  labs(
    x = "Fungicide Treatment",
    y = "Wheat Yield (t/ha)")+
  theme_bw()
```

This type of visualization combines the
statistical results with a clear graphical
representation of the treatment effects.



Like I mentioned in the lecture, I personally like
the type of graph below. It clearly shows the data
behind the results and effectively communicates
the treatment differences.

```{r}
ggplot(groups) +
  geom_jitter(data = mildew, aes(x = trt, y = yield),  width = 0.1, col = 'grey40')+
  geom_errorbar(aes(x = trt, ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_point(aes(x = trt, y = emmean),  col = 'red', size = 2) +
  geom_text(aes(x = trt, y = upper.CL + 0.5, label = .group), size = 5) +
  labs(
    x = "Fungicide Treatment",
    y = "Wheat Yield (t/ha)")+
  theme_bw()
```


For some reason, I also like to flip the x and y
axes sometimes. This is more visually appealing to
me. I am show you this to give you options.

```{r}
ggplot(groups) +
  geom_jitter(data = mildew, aes(x = trt, y = yield),  width = 0.1, col = 'grey40')+
  geom_errorbar(aes(x = trt, ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_point(aes(x = trt, y = emmean),  col = 'red', size = 2) +
  geom_text(aes(x = trt, y = upper.CL + 0.5, label = .group), size = 5) +
  labs(
    x = "Fungicide Treatment",
    y = "Wheat Yield (t/ha)")+
  theme_bw()+
  coord_flip()
```