---
title: "Exercise: Split-Plot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

In this exercise we will walk through a **complete
split-plot analysis**, from loading the data to
producing final tables and figures.

The goal is to walk through the entire workflow:

* loading the data
* visualizing the experiment
* fitting the mixed model
* separating treatment means
* comparing LSD and HSD results
* testing agronomic contrasts
* presenting the results in tables and figures

By the end, you should have a clear picture of
what a full split-plot analysis looks like in
practice.


```{r}
library(ggplot2)
library(emmeans)
library(multcomp)
library(nlme)
```

# The Experiment

A research team evaluated corn yield responses to
two management factors:

* Cover crop species (`species`): Peas, Clover, Rye
* Termination timing (`termination`): Early, Late

Unlike the fungicide experiment from the previous
exercise, here we have **two treatment factors
applied at different experimental scales**.

Because **termination timing is difficult to apply
at small scales**, it was applied to larger strips
within each block.

Within those strips, the **cover crop species were
randomized to smaller plots**.

This structure creates a **split-plot design**.

Structure of the experiment:

* **Blocks**: `block`
* **Main-plot factor**: `termination`
* **Subplot factor**: `species`
* **Main plots within blocks**: `main_plot`

This structure is important because it determines
**how we specify the statistical model**.

## Load the Data

Let's start by loading the dataset.

```{r}
splitplot <- read.csv("data/cover-crop-and-termination-split.csv")
head(splitplot)
```

Notice that we convert these variables to
**factors** so that R treats them as categorical
variables in the statistical model.

## Visualize the Data

Before fitting any statistical model, it is
always a good idea to **look at the data**.

Here we plot yield distributions for each
treatment combination.

```{r}
ggplot(splitplot,
       aes(x = termination, 
           y = yield)) +
  geom_point(shape = 1)+
  stat_summary(fun.data = 'mean_se')+
  labs(
    x = "Termination x Species",
    y = "Corn Yield (bu/ac)",
    fill = "Termination"
  ) +
  facet_wrap(~ species)+
  theme_linedraw()
```

Another useful visualization is an **interaction
plot**, which shows how treatment means change
across factor levels.

```{r}
ggplot(splitplot,
       aes(x = termination,
           y = yield,
           color = species,
           group = species)) +
  stat_summary(fun = mean,
               geom = "line",
               linewidth = 1) +
  stat_summary(fun = mean,
               geom = "point",
               size = 3) +
  labs(
    x = "Termination Timing",
    y = "Mean Yield (bu/ac)",
    color = "Species"
  ) +
  theme_linedraw()
```

This type of plot is particularly helpful for
**spotting potential interactions**.

# Fit the Split-Plot Model

For a split-plot design we must account for
**two levels of random variation**:

* variation among **blocks**
* variation among **main plots within blocks**

A suitable statistical model is

$$
Y_{ijk} = \mu + S_i + T_j + ST_{ij} + b_k + m_{jk} + \varepsilon_{ijk}
$$

where

* $S_i$ = species effect
* $T_j$ = termination timing effect
* $ST_{ij}$ = interaction
* $b_k$ = block random effect
* $m_{jk}$ = main plot random effect (termination within block)

Now let's fit this model in R.

```{r}
model_split <- lme(
  yield ~ species * termination,
  random = ~1 | block/main_plot,
  data = splitplot
)
```

# Interpret the ANOVA

Now we examine the ANOVA table.

```{r}
anova(model_split)
```


# Means Separation

We will use the **emmeans** package to compute
estimated marginal means.

```{r}
emm <- emmeans(model_split, ~ species * termination)
```

Now we can compare treatment levels.

```{r}
pairs(emm, adjust = "tukey")
```

## LSD versus HSD

For teaching purposes, it is useful to compare
**LSD-style inference** and **Tukey's HSD**.

In `emmeans`:

* `adjust = "none"` corresponds to **LSD-style comparisons**
* `adjust = "tukey"` corresponds to **Tukey's HSD**

```{r}
pairs(emm, adjust = "none")
pairs(emm, adjust = "tukey")
```

We can also compare the grouping letters produced
by each approach.

```{r}
cld(emm, adjust = "none", Letters = letters)
cld(emm, adjust = "tukey", Letters = letters)
```


# Contrasts Between Groups of Treatments

Pairwise comparisons are useful, but many
agronomic questions involve **groups of
treatments**.

For example:

* Are **legumes different from rye**?
* Is **peas different from clover**?

```{r}
emm_species <-  emmeans(model_split, ~ species)

contrast(emm_species,
         method = list(
           legumes_vs_rye = c(1/2, 1/2, -1),
           peas_vs_clover = c(1, -1, 0)
         ),
         adjust = "none")
```


# Build a Means Table

Now let's build a reporting table with

* treatment means
* standard errors
* confidence intervals
* Tukey grouping letters

First, let's store the groups of means in an object.

```{r}
cld(emm, adjust = "tukey", Letters = letters)
cld(emm, adjust = "tukey", Letters = letters, decreasing = TRUE)
```

```{r}
groups <- cld(emm, 
              adjust = "tukey",
              Letters = letters,
              decreasing = TRUE)
groups
```

Now that we have the `groups` data frame, we can
build a table.

```{r}
library(flextable)

groups <- data.frame(groups)

tab <- flextable(groups)
tab
```

Let's format the numeric columns.

```{r}
tab <- colformat_double(tab,
                        j = c("emmean","SE","lower.CL","upper.CL"),
                        digits = 2)

tab
```

Now we improve the column labels.

```{r}
tab <- set_header_labels(tab,
                         values = c("Species",
                                    'Termination',
                                    "Mean",
                                    "SE",
                                    "DF",
                                    "Lower CL",
                                    "Upper CL",
                                    "Group"))

tab
```

We can also add a caption for the table:
```{r}
tab <- set_caption(tab, 
                   caption = as_paragraph(as_b('Table 1:'),'Mean corn yield for species and termination timing combinations in a split-plot experiment.'))
tab
```

Adding a footnote about the procedure used to separate means is a good idea too:

```{r}
tab <- footnote(tab, 
                j = 8, 
                part = 'header',
         value = as_paragraph('Letters indicate mean separation groups using a Tukey procedure at α = 0.05; treatments sharing a common letter are not significantly different.'))
tab
```



Finally, we can use the `autofit()` function to adjust columns widths

```{r}
tab <- autofit(tab)
tab
```

Then we can export the table if needed.

```{r}
save_as_image(tab, path = "species_table.png")
save_as_docx(tab, path = "species_table.docx")
```

# Build a Figure

Finally, let's create a figure showing

* treatment means
* confidence intervals
* Tukey grouping letters

```{r}

ggplot(groups) +
  geom_col(aes(x = termination, y = emmean), fill = "grey50", col = 'black') +
  geom_errorbar(aes(x = termination, ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_text(aes(x = termination, y = upper.CL + 15, label = .group), size = 5) +
  labs(
    x = "Termination x Species",
    y = "Corn Yield (bu/ac)",
    fill = "Termination")+
  theme_linedraw()+
  facet_wrap(~species)
```

This type of visualization combines the
statistical results with a clear graphical
representation of the treatment effects.



Like I mentioned in the lecture, I personally like
the type of graph below. It clearly shows the data
behind the results and effectively communicates
the treatment differences.

```{r}
ggplot(groups) +
  geom_jitter(data = splitplot, aes(x = termination, y = yield),  width = 0.1, col = 'grey40')+
  geom_errorbar(aes(x = termination, ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_point(aes(x = termination, y = emmean),  col = 'red', size = 2) +
  geom_text(aes(x = termination, y = upper.CL + 15, label = .group), size = 5) +
  labs(
    x = "Termination x Species",
    y = "Corn Yield (bu/ac)",
    fill = "Termination")+
  theme_linedraw()+
  facet_wrap(~species)
```


For some reason, I also like to flip the x and y
axes sometimes. This is more visually appealing to
me. I am show you this to give you options.

```{r}
ggplot(groups) +
  geom_jitter(data = splitplot, aes(x = termination, y = yield),  width = 0.1, col = 'grey40')+
  geom_errorbar(aes(x = termination, ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_point(aes(x = termination, y = emmean),  col = 'red', size = 2) +
  geom_text(aes(x = termination, y = upper.CL + 15, label = .group), size = 5) +
  labs(
    x = "Termination x Species",
    y = "Corn Yield (bu/ac)",
    fill = "Termination")+
  theme_linedraw()+
  facet_wrap(~species)+
  coord_flip()
```