---
title: "Means Separation and Data Presentation"
subtitle: "AGRON 5130"
author: "YOUR NAME HERE"
output: pdf_document
---

# Introduction

The goal of this assignment is to give you hands-on
practice with **means separation**, **linear contrasts**,
and **presenting treatment means**.


You will work with an apple yield dataset to practice:

- fitting an ANOVA model
- comparing treatment means using **LSD** and **Tukey** comparisons
- constructing and testing **linear contrasts**
- presenting treatment means in a **figure with grouping letters**

Submit a **knitted PDF/HTML generated from your R Markdown
file** to Canvas.

We can start by loading the necessary libraries

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(emmeans)
library(multcomp)
library(nlme)
```


# Question 1

**1 point**

Load the dataset `"data/apple.csv"` and display
the first few rows of the dataset. This dataset
contains apple yield data from an experiment that
evaluated foliar treatments. The experiment was
conducted as a randomized complete block design
with 4 replications.

```{r}
apple <- read.csv('./data/apple.csv')
head(apple)
```



# Question 2

**2 points**

Build a plot to visualize the relationship between
`yield` and `foliar_trt`

```{r}
ggplot(apple, aes(x = foliar_trt, y = yield)) +
  geom_jitter(width = 0.1, 
              size = 2, 
              alpha = 0.8, 
              color = "steelblue", 
              shape = 1) +
  labs(x = "Foliar Treatment",
       y = "Apple Yield (t/ha)") +
  stat_summary(fun.data = 'mean_se')+
  theme_bw()
```

# Question 3

**2 points**

Fit an ANOVA model testing the effect of `foliar_trt`
on apple yield.

Show the ANOVA table.


```{r}
model <- lme(yield ~ foliar_trt,
             random = ~ 1 | block,
             data = apple)
anova(model)
```


# Question 4

**2 points**

Use the **emmeans** package to compute the estimated
marginal means for `foliar_trt`.

Then conduct **pairwise comparisons using LSD
inference**.

Show the output.


```{r}
emm <- emmeans(model, ~ foliar_trt)
pairs(emm, adjust = 'none')
```


# Question 5

**3 points**

Now conduct the same pairwise comparisons using
**Tukey's HSD adjustment**.

Show the results and briefly explain how they differ
from the LSD-style comparisons.


```{r}
pairs(emm, adjust = 'tukey')
```


# Question 6

**3 points**

Now test a **linear contrast** comparing

**calcium plus additional ingredients**

versus

**calcium alone**.

First examine the order of treatment levels in
`foliar_trt`. Then compute the estimated marginal
means using `emmeans`.

Construct the appropriate **contrast coefficients**
and test the hypothesis using the `contrast()`
function.

Briefly interpret the results.


```{r}
contrast(emm, 
         method = list(
           calcium_vs_alone = c(-1, -1/2, -1/2, 0)
         ))
```


# Question 7

**1 point**

Use `cld()` to obtain **Tukey grouping letters**
for the treatment means.

Show the output.


```{r}
cld(emm, adjust = 'tukey', Letters = letters)
```


# Question 8

**1 point**

Create a **plot using ggplot2** that shows:

- treatment means
- confidence intervals
- grouping letters above the bars

Your final figure should resemble the type of plot
shown in the lecture.


```{r}

```