---
title: "Nonlinear Relationships and Multiple Linear Regression"
subtitle: "AGRON 5130 - Module 11"
author: "Caio dos Santos"
date: "2026-03-11"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    smooth: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Introduction

Simple linear regression is powerful, but many
agronomic systems are more complex than one X and
one straight line.

In this unit, we cover:

- multiple linear regression
- model bloat, collinearity, and cross-validation
- nonlinear relationships and common nonlinear
  curves

# Multiple Linear Regression

## Model Form

$$Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \cdots + \beta_pX_p + \varepsilon$$

This model allows several predictors to explain
variation in a single response.

## Example Dataset

```{r}
set.seed(20260315)
n <- 200
county <- data.frame(
  ppt = rnorm(n, 920, 90),
  whc = rnorm(n, 0.21, 0.03),
  om = rnorm(n, 3.2, 0.8),
  clay = rnorm(n, 26, 8)
)

county$corn <- 70 +
  0.07 * county$ppt +
  80 * county$whc +
  3.6 * county$om -
  0.35 * county$clay +
  rnorm(n, 0, 8)

head(county)
```

## Fit and Interpret

```{r}
mlr1 <- lm(corn ~ ppt + whc + om + clay, data = county)

library(broom)
tidy(mlr1)
glance(mlr1)
```

# Beware of Bloated Models

More predictors are not always better.

Problems with bloated models:

- unstable coefficients
- overfitting
- poor out-of-sample performance
- hard interpretation

## Check Correlation Among Predictors

```{r}
plot(county[, c("ppt", "whc", "om", "clay")])
cor(county[, c("ppt", "whc", "om", "clay")])
```

## Partial Correlations

```{r}
library(psych)
partial.r(county[, c("ppt", "whc", "om", "clay")])
```

## Cross-validation

```{r}
library(caret)
set.seed(123)
ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 3)

cv_fit <- train(corn ~ ppt + whc + om + clay,
                data = county,
                method = "lm",
                trControl = ctrl)
cv_fit
```

Use cross-validated metrics (e.g., RMSE, Rsquared)
to compare alternative model forms.

# Nonlinear Relationships

Some biological relationships are curved by
nature, not linear.

Common examples:

- nutrient response
- growth over time
- asymptotic biomass accumulation

## Exponential Relationship

```{r}
x <- seq(1, 100, by = 1)
y_exp <- 2 * exp(0.03 * x) + rnorm(length(x), 0, 1.2)

plot(x, y_exp, pch = 21, bg = "grey70",
     xlab = "X", ylab = "Y", main = "Exponential Pattern")
```

## Parabolic Relationship

```{r}
y_par <- 40 + 1.8 * x - 0.015 * x^2 + rnorm(length(x), 0, 2)

plot(x, y_par, pch = 21, bg = "grey70",
     xlab = "X", ylab = "Y", main = "Parabolic Pattern")
```

## Monomolecular (Asymptotic)

```{r}
set.seed(20260316)
pop <- seq(1, 10, by = 0.2)
y_mono <- 18 - 13 * exp(-0.5 * pop) + rnorm(length(pop), 0, 0.5)

mono_dat <- data.frame(pop = pop, y = y_mono)
mono_fit <- nls(y ~ SSasymp(pop, Asym, R0, lrc), data = mono_dat)
summary(mono_fit)
```

## Logistic Growth

```{r}
gdd <- seq(100, 600, by = 10)
la <- 8 / (1 + exp(-(gdd - 360) / 55)) + rnorm(length(gdd), 0, 0.25)

log_dat <- data.frame(gdd = gdd, la = la)
log_fit <- nls(la ~ SSlogis(gdd, Asym, xmid, scal), data = log_dat)
summary(log_fit)
```

# Choosing Starting Values

For custom nonlinear equations, starting values
can determine whether optimization converges.

Good starts often come from:

- plotting first
- biological constraints
- rough parameter estimates from data range

# Conclusion

This unit focused on model realism:

- multiple predictors improve explanatory power
- but model complexity must be controlled
- cross-validation helps avoid overfitting
- many agronomic responses are nonlinear and should
  be modeled accordingly

These tools improve both predictive performance and
biological interpretability.
