---
title: "Exercise: Avoiding Bloated Models"
output: html_document
---

# Introduction

In this exercise, you will evaluate model complexity,
identify collinearity, and assess generalization with
cross-validation.

# Case Study

```{r}
library(tidyverse)
county_corn_data = read.csv("data/county_environments.csv")

ia_and_il_corn = county_corn_data %>%
  filter(state %in% c("IL", "IA"))

ia_and_illinois_model = lm(corn ~ ppt + whc + sand + silt + clay + om, data = ia_and_il_corn)
```

## Covariance Matrix

```{r}
variables_of_interest = ia_and_il_corn %>%
  dplyr::select(ppt, whc, sand, silt, clay, om)
```

```{r}
plot(variables_of_interest)
```

## Partial Correlation Matrix

```{r}
library(psych)

partial.r(variables_of_interest)
```

## Cross-Validation

```{r}
library(caret)

set.seed(123)
train.control <- trainControl(method = "repeatedcv",
                              number = 10, repeats = 3)

model <- train(corn ~ ppt + whc + sand + silt + clay + om,
               data=ia_and_il_corn,
               method = "lm",
               trControl = train.control)

print(model)
```

# Practice: Thompson Corn Data

Use the Thompson dataset to inspect potential
collinearity in a four-predictor model.

```{r}
thompson_corn = read.csv("data/thompson_corn.csv")

thompsom_model = lm(yield ~ rain6 + rain7 + temp6 + temp7, thompson_corn)
```

1. Plot the covariance matrix. Which predictors look
correlated?

```{r}

```

2. Compute partial correlations. Which relationships
appear significant?

```{r}

```

3. Cross-validate the model. Target cross-validated
`Rsquared` is approximately `0.537`.

```{r}
library(caret)

set.seed(123)
train.control <- trainControl(method = "repeatedcv",
                              number = 10, repeats = 3)


```
