---
title: "Exercise: Multiple Linear Regression"
output: html_document
---

# Introduction

This exercise introduces multiple linear regression,
where one response is modeled as a function of
multiple predictors.

# Case Study: County Corn Yields

```{r}
library(tidyverse)
county_corn_data = read.csv("data/county_environments.csv")
head(county_corn_data)

ia_and_il_corn = county_corn_data %>%
  filter(state %in% c("IL", "IA"))

head(ia_and_il_corn)
```

## Fit the Model

```{r}
ia_and_illinois_model = lm(corn ~ ppt + whc + sand + silt + clay + om, data = ia_and_il_corn)
```

## Inspect Coefficients

```{r}
library(broom)
tidy(ia_and_illinois_model)
```

## Evaluate Model Fit

```{r}
glance(ia_and_illinois_model)
```

# Practice: Thompson Corn Data

Build an initial model of `yield` as a function of
`rain6`, `rain7`, `temp6`, and `temp7`.

```{r}
thompson_corn = read.csv("data/thompson_corn.csv")
head(thompson_corn)



```

Use `tidy()` to inspect coefficient estimates.
Your estimates for `rain7` and `temp7` should be
near `4.00` and `2.31`.

```{r}
library(broom)


```

Use `glance()` to evaluate model fit. You should see
`r.squared` near `0.55`.

```{r}

```
