---
title: "Introduction to Shapefiles"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Introduction

One of my more recent areas of growth as a data
scientist has been learning to work with spatial
data. If you have worked with yield monitor data,
*ArcGIS*, or used an application map to vary the
application rate of your fertilizer, you have
worked with spatial data. We will learn how to
plot spatial data in detail in Module 12, but here
is a brief introduction to importing and
inspecting spatial data.

# An R package for spatial data

To look at spatial data, we will use the R package
`sf` (short for *simple features*). Remember,
when we use a new package in R for the first time,
we must install and launch it with the following
commands. The `sf` package takes a while to
install. 

```{r}
install.packages('sf')
library(sf)
```

# Reading Shapefiles

To read in a shape file, we use the `st_read()`
function. We read in a shape file the same way we
read in a .csv file in the previous exercise. We
follow `st_read()` with the path to the shape
file, which has a .shp suffix.

We will read in a shape file from Indiana that
includes county summaries for a couple of dozen
agricultural statistics.

```{r}
indiana <- st_read("data/Agriculture_Census-shp/Agriculture_Census.shp")
```

# Examining Simple Feature Data Frames

Let's inspect the first six rows of the spatial
feature data frame using the `head()` command,
where the name of the data frame is given between
the parentheses.

```{r}
head(indiana)
```

Spatial feature data frames are very similar to
the data frames that were introduced to in the
last exercise. There are a few differences,
however. First, every spatial feature data frame
consists of rows of observations that are linked,
in the geometry column, with a georeference. That
georeference may be a simple point on the map, or
it may be a polygon that defines an area, such as
a field, a county, or a state.

Spatial feature data frames also include some
metadata about the geometry of the file, including
geometry type (POLYGON in this case, since we are
working with county outlines) and -- this is
critical -- *CRS*. CRS stands for **coordinate
reference system**, the "projection" we use to
draw a map, and the georeferences used to define a
place or area.

Maps are flat representations of a round object,
and there are different ways to draw these maps.
We need not go into a deep review of those
projects at this point; suffice it to know that
each projection represents tradeoffs in how
accurately it portrays areas in the center and
periphery of the map and how much it represents
the curvature. For most of my work, I use the CRS
of 4326. You know this system more commonly as the
latitude/longitude system.

The bounding box defines where the map is
centered. Latitude/longitude maps are more
accurate in portraying areas near the center of
the map than the periphery, so it is important the
map be centered on the area of interest.

Don't worry about committing the above information
to memory for now, we will discuss projection is
much greater detail in Unit 12.

# Visualizing Data

To visualize the data, we can use the `plot()`
function from the sf package. You might assume we
could map a column of data the same way we
visualized its contents earlier, by indicating the
data frame and column name, separated by the
dollar sign.

```{r}
plot(indiana)
```


We will try that with TOTCROP_AC, the
total number of acres per county in crop
production.

```{r}
tot_crop <- indiana$TOTCROP_AC

plot(tot_crop)
```

Instead of a map, however, this shows us the range
of values for TOTCROP_AC in the dataset.

To view a map, we need to first subset the data a
little differently than we did before: we can't
use the dollar-sign approach. That only assigns
the values of our single column to our new R
object.

See what happens if we look at the top of our
tot_crop object using the `head()` command

```{r}
head(tot_crop)
```

The object only has a single column of numbers
with the county values. It lacks information on
the geometry of those counties, so that they can
be drawn.

To draw our map, we need to assign two columns to
our new object: TOTCROP_AC and geometry columns.
We do this by referencing the desired measure as
`indiana["TOTCROP_AC"]`. This tells R to pass both
the TOTCROP_AC values and their geometries to the
new object, `tot_crop`.

```{r}
tot_crop <- indiana["TOTCROP_AC"]
head(tot_crop)
```

When we examine the top six rows of our object,
tot_crop, we see it contains both the county
values and the county geometries.

Note: we only use this approach, with the desired data column in quotes and brackets, with shape file data. The rest of the time, the data frame and desired column should be separated by the dollar sign, as we learned above.

We can then map these data:

```{r}
plot(tot_crop)
```


Let's look at the mean farm income per county.
This is indicated by the "DOL_FARM" column. Again,
we first subset the data and then plot it.

```{r}

farm_income <- indiana["DOL_FARM"]
plot(farm_income)
```

It looks like the larger farms, by income, are in
northwest Indiana, just south of Lake County and
the Chicago suburbs. Where are the chickens in
Indiana? Let's find out:

```{r}
chickens <- indiana["N_CHICKENS"]
plot(chickens)
```


# Create Your Own Maps

Where are the hogs in Indiana? Create a map for
"N_PIGS".

```{r}

```

Where are most of the wheat acres? Create a map
for "WHEAT_AC".

```{r}

```

Which counties have greater farm ownership by
minorities? Create a map for "MINORTY_OP" (note
the missing "I").
```{r}

```

