---
title: "Module 10 Assignment"
subtitle: "AGRON 5130"
author: "Caio dos Santos"
output: pdf_document
---

# Introduction

This assignment extends regression analysis to
multiple and nonlinear settings. You will build,
inspect, and evaluate models using diagnostics,
collinearity checks, and cross-validation.

```{r}
library(tidyverse)
library(broom)
library(caret)
```

# Question 1: Multiple Linear Regression (7 points)

```{r}
thompson_soy <- read.csv("data/thompson_soy.csv")
head(thompson_soy)
```

## Question 1a (1 point)
Model `yield` as a function of all four weather
measurements.

```{r}

```

## Question 1b (1 point)
Report coefficient values using `tidy()`.

```{r}

```

## Question 1c (1 point)
Report model fit statistics using `glance()`.

```{r}

```

## Question 1d (1 point)
Examine the covariance structure among predictor
variables. Name one pair that appears related.

```{r}

```

## Question 1e (1 point)
Examine partial correlations among predictor
variables and name one correlated pair.

```{r}
library(ppcor)

```

## Question 1f (2 points)
Cross-validate the model with repeated 10-fold CV
and report performance.

```{r}
set.seed(123)
train.control <- trainControl(method = "repeatedcv",
                              number = 10, repeats = 3)

# Complete model training code below

```

# Question 2: Nonlinear Regression (13 points)

```{r}
tomato <- read.csv("data/tomato_days_volume.csv")
head(tomato)
```

## Question 2a (2 points)
Plot the original data (`days` vs `fruit_volume`).

```{r}

```

## Question 2b (3 points)
Fit and summarize a logistic nonlinear model of
`fruit_volume` as a function of `days`.

```{r}

```

## Question 2c (2 points)
Create a new dataset, `tomato_predicted`, with
`days` from 7 to 49 in increments of 1.

```{r}

```

## Question 2d (2 points)
Predict fruit volume for each day in
`tomato_predicted`.

```{r}

```

## Question 2e (2 points)
Add the modeled logistic curve to the initial
scatterplot.

```{r}

```

## Question 2f (2 points)
Interpret the fitted curve.

- Where does growth accelerate most rapidly?
- Where does growth begin to plateau?
- Why is a logistic model more appropriate than a
  straight line for this process?
