---
title: "Correlation and Simple Regression"
subtitle: "AGRON 5130 - Unit 10 Assignment"
author: "Caio dos Santos"
output: pdf_document
---

# Introduction

The goal of this assignment is to help you practice
the full workflow from Module 10:

- visualizing quantitative relationships
- interpreting correlation carefully
- fitting and interpreting a simple linear regression
- checking model assumptions
- making predictions and distinguishing interpolation
  from extrapolation

You will use a wheat–weed dataset collected in India.
The key biological question is whether increasing weed
pressure is associated with lower wheat yield.

Submit a knitted PDF/HTML generated from your R
Markdown file to Canvas.

```{r}
library(ggplot2)

weed <- read.csv("data/wheat_weed_india.csv")
head(weed)
```



# Question 1

**1 point**

Create a scatterplot matrix for all quantitative
variables in `weed`.

```{r}
plot(weed)
```

---

# Question 2

**1 point**

Create a correlation matrix for all quantitative
variables in `weed`. Briefly identify which pair
of variables shows the strongest association.


```{r}
round(cor(weed), 2)
```


# Question 3

**1 point**

Create a scatterplot showing the relationship between
`total_weed_pop` and `yield`.

Optional but recommended: overlay a linear trend line.

```{r}
plot(weed$total_weed_pop, weed$yield)
```


# Question 4

**1.5 points**

Run a correlation test between `total_weed_pop` and
`yield` using `cor.test()`.

```{r}
cor.test(weed$total_weed_pop, weed$yield)
```

In 2–3 sentences, interpret:

1. direction of association
2. strength of association
3. whether this result implies causation


# Question 5

**1.5 points**

Fit a simple linear regression model for `yield` as a
function of `total_weed_pop`.

Display coefficient estimates using `coef()`.

```{r}
model1 <- lm(yield ~ total_weed_pop, data = weed)
coef(model1)
```


# Question 6

**1 point**

Display $R^2$ for the model using `summary()`.

```{r}
summary(model1)
```


# Question 7

**2 points**

Interpret the model from Questions 5–6.

Address all of the following:

1. Is the slope positive or negative?
2. Is the slope significantly different from zero?
3. What percentage of yield variation is explained by
   `total_weed_pop`?
4. In plain language, what does this model imply
   biologically?


# Question 8

**1.5 points**

Check model assumptions for the regression from Question 5.

Produce:

1. residuals vs fitted plot (`which = 1`)
2. normal Q-Q plot (`which = 2`)

```{r}
plot(model1, which = 1)
plot(model1, which = 2)
```

Then write 2–3 sentences on whether a straight-line model
appears adequate.


# Question 9

**1 points**

Using `ggplot()`, create a plot of `yield` vs
`total_weed_pop` with:

1. observed points
2. fitted regression line

```{r}

ggplot(weed, aes(x = total_weed_pop, y = yield))+
  geom_point()+
  geom_smooth(method = 'lm', se = F)
```


# Question 10

**2 points**

Using the `new_weed_pops` data frame below, predict yield
for each value of `total_weed_pop`.

*Note:* negative values are included intentionally to
explore extrapolation.

```{r}
new_weed_pops <- data.frame(
  total_weed_pop = c(-1, 1, 3, 5, 7)
)

```

Generate:

1. predicted values
2. 95% confidence intervals for the mean response

*Hint:* use `predict(..., interval = "confidence")`

```{r}
predict(model1, newdata = new_weed_pops, interval = 'conf')
```


# Question 11

**1.5 points**

From Question 10, classify each prediction as:

1. interpolation
2. extrapolation

Justify your classification using the observed range of
`total_weed_pop` in the original data.

Which predictions would you trust the most? Why?


# Optional Bonus Question

**+1 point**

Based on your residual plots, do you see any signs that
the assumptions of the linear model may be violated?

If so, what alternative approach might be more appropriate?
