---
title: "Exercise - Multiple Linear Regression"
output: html_document
---

# Introduction

In this exercise, you will practice the core move
of Module 11: interpreting coefficients as
**partial effects**.

The workflow is simple:

1. Start with a simple two-predictor model to lock in intuition.
2. Expand to a fuller model.
3. Interpret coefficients using biology, not p-values alone.

Before you start, make sure the `car` package is
installed. We will use it later in this module for
VIF.

```{r}
install.packages('car')
```


```{r}
library(car)
```

# Case Study: Corn Yield in Minnesota and Wisconsin

```{r}
library(ggplot2)

counties <- read.csv("data/county_environments.csv")

model_data <- subset(counties, state %in% c("MN", "WI"))
model_data <- na.omit(model_data[, c("corn", "ppt", "sand", "clay", "om", "sph")])

head(model_data)
```

## Plot first

```{r}
plot(model_data, pch = 20, cex = 0.3)
```

Quick check:

- Which predictors appear correlated?
- Which relationships with `corn` look roughly linear?

## Start simple: two predictors

Fit yield as a function of rainfall and organic matter.

```{r}
model_simple <- lm(corn ~ ppt + om, data = model_data)
summary(model_simple)
```

Now inspect coefficients directly from the model summary:

```{r}
coef(model_simple)
```

Interpretation prompt:

- Write one sentence interpreting the `om` coefficient as a partial effect.
- Start with: "Holding rainfall constant, ..."

## Visualize partial effects

```{r}
om_levels <- quantile(model_data$om, probs = c(0.10, 0.90), na.rm = TRUE)

pred_grid <- expand.grid(
  ppt = seq(min(model_data$ppt), max(model_data$ppt), length.out = 200),
  om = as.numeric(om_levels)
)

pred_grid$om_group <- factor(
  pred_grid$om,
  levels = as.numeric(om_levels),
  labels = c("Lower OM (10th percentile)", "Higher OM (90th percentile)")
)

pred_grid$corn_pred <- predict(model_simple, newdata = pred_grid)

ggplot() +
  geom_point(
    data = model_data,
    aes(x = ppt, y = corn),
    color = "gray65",
    alpha = 0.35,
    size = 1.7
  ) +
  geom_line(
    data = pred_grid,
    aes(x = ppt, y = corn_pred, color = om_group),
    linewidth = 1.2
  ) +
  scale_color_manual(values = c("#1b9e77", "#d95f02")) +
  labs(
    title = "Partial Effects in a Two-Predictor Model",
    subtitle = "Predicted lines from corn ~ ppt + om",
    x = "Rainfall (ppt)",
    y = "Predicted corn yield",
    color = "Organic matter"
  ) +
  theme_minimal(base_size = 12)
```

Pause and reflect:

- At the same rainfall value, what does the vertical gap between lines represent?
- Why are the lines parallel here?

## Expand to a fuller model

```{r}
model_full <- lm(corn ~ ppt + sand + clay + om + sph, data = model_data)
summary(model_full)
AIC(model_full)
```


Quick check:

- Did any coefficient signs or p-values change vs the simple model?
- Why can that happen when predictors are correlated?

# Practice 1: Thompson Corn Data

Fit a model of `yield` using `rain6`, `rain7`, `temp6`, and `temp7`.

```{r}
thompson_corn <- read.csv("data/thompson_corn.csv")
head(thompson_corn)
```

1. Fit the model.
2. Use `summary(model)` to inspect coefficients.
3. Report fit metrics (`adj.r.squared`, `AIC`).

```{r}
# your code here
```

Write 2-3 sentences:

- Which coefficients are easiest to interpret biologically?
- Which ones feel unstable or uncertain?

# Practice 2: Partial-Effect Language

Using your Thompson model, write one partial-effect interpretation for:

- `rain7`
- `temp7`

Template:

"Holding the other predictors constant, a one-unit increase in ___ is associated
with an expected ___-unit change in yield."

Does this prove intervention-level causation? Why or why not?
