---
title: "Exercise - Avoiding Bloated Models"
output: html_document
---

# Introduction

In this exercise, you will diagnose when a model has too many overlapping
predictors.

Key idea: a model can predict reasonably well and still give unstable
coefficient interpretations when predictors are strongly correlated.

# Case Study: Corn Yield and Collinearity

```{r}
library(ggplot2)
library(car)

counties <- read.csv("data/county_environments.csv")

model_data <- subset(counties, state %in% c("MN", "WI"))
model_data <- na.omit(model_data[, c("corn", "ppt", "sand", "silt", "clay", "om", "sph")])

head(model_data)
```

## Start with a fuller model

```{r}
model_full <- lm(corn ~ ppt + sand + silt + clay + om + sph, data = model_data)
summary(model_full)
```

## Correlation matrix as first warning sign

```{r}
round(cor(model_data[, c("ppt", "sand", "silt", "clay", "om", "sph")]), 2)
```

Quick check:

- Which pairs are most strongly correlated?
- Which predictors might compete to explain the same variation?

## VIF as direct inflation check

```{r}
vif(model_full)
```

Interpretation guide:

- Around 1: little inflation
- Around 5: meaningful inflation
- Around 10+: severe inflation

Pause and reflect:

- Could a predictor with a weak p-value still matter biologically?
- Could a predictor with a significant p-value still be unstable under collinearity?

## The compositional texture trap

Sand, silt, and clay are compositional percentages. Together, they sum to 100.
That can cause near-perfect linear dependence.

Try dropping one texture class (silt) and refit:

```{r}
model_drop_silt <- lm(corn ~ ppt + sand + clay + om + sph, data = model_data)
summary(model_drop_silt)
vif(model_drop_silt)
```

Then try dropping a different texture class (clay):

```{r}
model_drop_clay <- lm(corn ~ ppt + sand + silt + om + sph, data = model_data)
summary(model_drop_clay)
vif(model_drop_clay)
```

Quick check:

- Which reduced model gives more stable VIF values?
- How do coefficient signs and uncertainty change across models?

# Practice: Thompson Corn Data

```{r}
thompson_corn <- read.csv("data/thompson_corn.csv")
head(thompson_corn)
```

1. Fit `lm(yield ~ rain6 + rain7 + temp6 + temp7, data = thompson_corn)`.
2. Compute a correlation matrix for the predictors.
3. Compute VIF values.
4. Drop one predictor you think is redundant and refit.

```{r}
# your code here
```

Write 3-4 sentences:

- What evidence suggested redundancy?
- Did dropping one variable improve interpretability?
