---
title: "Exercise - Tuning the Model"
output: html_document
---

# Introduction

In this exercise, you will make a practical model-selection decision.

The question is not "Which model has the most predictors?"
The question is "Which model is simple, biologically sensible, and still useful?"

# Case Study: Full vs Reduced Model

```{r}
counties <- read.csv("data/county_environments.csv")

model_data <- subset(counties, state %in% c("MN", "WI"))
model_data <- na.omit(model_data[, c("corn", "ppt", "sand", "clay", "om", "sph")])
```

Fit full and reduced models:

```{r}
model_full <- lm(corn ~ ppt + sand + clay + om + sph, data = model_data)
model_reduced <- lm(corn ~ ppt + sand + om + sph, data = model_data)
```

## Compare coefficients and fit

```{r}
coef(model_full)
coef(model_reduced)
```

```{r}
summary(model_full)
summary(model_reduced)
```


## Compare models directly

```{r}
anova(model_reduced, model_full)
AIC(model_reduced, model_full)
```

Interpretation prompts:

- If ANOVA p-value is large, what does that suggest about the dropped predictor?
- If AIC is similar or lower for reduced model, what does that suggest?

Important note:

This ANOVA comparison assumes models are nested and use the same observations.


# Practice: Thompson Corn Data

```{r}
thompson_corn <- read.csv("data/thompson_corn.csv")
```

Build and compare:

- Full: `yield ~ rain6 + rain7 + temp6 + temp7`
- Reduced candidate: drop one predictor you think is least useful

Tasks:

1. Fit both models.
2. Compare model summaries and fit metrics (`r.squared`, `adj.r.squared`, `sigma`, `AIC`).
3. Run nested `anova()`.
4. Compare AIC.
5. (Optional) Run the same 80/20 held-out RMSE check.

```{r}
# your code here
```

Final prompt (write 4-6 sentences):

Defend your final model for a real agronomy report. Your explanation should
include statistical evidence and biological reasoning.
