---
title: "Exercise - Nonlinear Regression"
output: html_document
---

# Introduction

In this exercise, you will fit two nonlinear models used constantly in agronomy:

- linear-plateau (`SSlinp`) for nutrient response
- logistic (`SSlogis`) for growth curves

Key idea: nonlinear models are worth the extra effort when they match biology
and give parameters you can actually use.

# Case Study 1: Nutrient Response (Linear-Plateau)

```{r}
install.packages('nlraa')
```


```{r}
library(ggplot2)
library(nlraa)

corn_n <- read.csv("data/corn_nrate_mono.csv")
head(corn_n)
```

Plot first:

```{r}
ggplot(corn_n, aes(x = n_rate, y = yield)) +
  geom_point(pch = 21, size = 2) +
  labs(x = "N rate", y = "Yield") +
  theme_bw()
```

Fit linear-plateau model:

```{r}
model_linp <- nls(yield ~ SSlinp(n_rate, a, b, xs), data = corn_n)
summary(model_linp)
```

Parameter intuition:

- `a`: intercept (yield at zero N)
- `b`: slope before the join point
- `xs`: join point (the N rate where yield stops responding)

Create predicted curve:

```{r}
grid_n <- data.frame(n_rate = seq(min(corn_n$n_rate), max(corn_n$n_rate), length.out = 200))
grid_n$pred <- predict(model_linp, newdata = grid_n)

ggplot(corn_n, aes(x = n_rate, y = yield)) +
  geom_point(pch = 21, size = 2) +
  geom_line(data = grid_n, aes(y = pred), color = "steelblue", linewidth = 1) +
  labs(x = "N rate", y = "Yield") +
  theme_bw()
```

Pause and reflect:

- Why is `xs` often more actionable than a quadratic coefficient?
- What management recommendation can you make from this curve?

# Case Study 2: Logistic Growth

```{r}
soy_tdm <- read.csv("data/soybean_dae_tdm_logis.csv")
head(soy_tdm)
```

Plot first:

```{r}
ggplot(soy_tdm, aes(x = dae, y = tdm)) +
  geom_point(pch = 21, size = 2) +
  labs(x = "Days after emergence", y = "Total dry matter") +
  theme_bw()
```

Fit logistic model:

```{r}
model_logis <- nls(tdm ~ SSlogis(dae, Asym, xmid, scal), data = soy_tdm)
summary(model_logis)
```

Parameter intuition:

- `Asym`: maximum biomass
- `xmid`: day of fastest growth
- `scal`: curve steepness

Predicted curve:

```{r}
grid_dae <- data.frame(dae = seq(min(soy_tdm$dae), max(soy_tdm$dae), length.out = 200))
grid_dae$pred <- predict(model_logis, newdata = grid_dae)

ggplot(soy_tdm, aes(x = dae, y = tdm)) +
  geom_point(pch = 21, size = 2) +
  geom_line(data = grid_dae, aes(y = pred), color = "darkgreen", linewidth = 1) +
  labs(x = "Days after emergence", y = "Total dry matter") +
  theme_bw()
```

# Practice 1: Ryegrass and Soil P (Linear-Plateau)

```{r}
ryegrass <- read.csv("data/ryegrass_soil_p_mono.csv")
head(ryegrass)
```

Tasks:

1. Plot `dm` vs `soil_p`.
2. Fit `nls(dm ~ SSlinp(soil_p, a, b, xs), data = ryegrass)`.
3. Build a prediction grid for `soil_p`.
4. Plot observed points + fitted curve.
5. Interpret `xs` agronomically.

```{r}
# your code here
```

# Practice 2: Aspergillus Growth (Logistic)

```{r}
aspergillus <- read.csv("data/aspergillus_hours_density_logist.csv")
head(aspergillus)
```

Tasks:

1. Plot `log_density` vs `hours`.
2. Fit logistic model with `SSlogis`.
3. Predict on a dense `hours` grid.
4. Plot observed points + fitted curve.
5. Interpret `xmid` in biological terms.

```{r}
# your code here
```

