---
title: "Spatial Data Analysis"
subtitle: "AGRON 5130 - Module 12 Assignment"
author: "Caio dos Santos"
output: pdf_document
---

# Introduction

The goal of this assignment is to practice the full
workflow from Module 12:

- loading and inspecting spatial vector data
- checking and using coordinate reference systems
- joining spatial datasets
- interpreting scale and resolution effects
- building interpolated surfaces (rasters) with kriging
- interpreting interpolated spatial surfaces

You will complete two parts:

- **Part 1:** vector data exploration and spatial joins
- **Part 2:** raster creation and spatial pattern analysis

Submit a knitted HTML/PDF generated from your R
Markdown file to Canvas.




# Setup

```{r}
library(sf)
library(gstat)
library(stars)
```


# Part 1: Vector Data and Spatial Joins

## Question 1

**2 points**

Read in the SSURGO polygon layer
(`data/sample-ssurgo.shp`) and the soil test data
(`data/soil_tests.shp`).

```{r}
ssurgo <- read_sf('./data/sample-ssurgo.shp')
soil <- read_sf('./data/soil_tests.shp')
```

For each dataset, report:

1. the geometry type
2. the CRS
3. whether the CRS is longlat or projected

**Interpretation:** If the CRS is longlat, which one
should you transform to a projected system, and why?


## Question 2

**2 points**

Transform both datasets to the same projected CRS
(use 32615 for UTM zone 15N).

```{r}
ssurgo <- st_transform(ssurgo, 32615)
soil <- st_transform(soil, 32615)
```

Make a map showing both layers together. Use
`plot()` to overlay SSURGO polygons (border only)
with soil test points on top.


## Question 3

**2 points**

Use `st_join()` with `st_within` to assign SSURGO
attributes (for example `mukey`, `muname`, `om_r`)
to each soil test point.

```{r}
soil_b <- st_join(soil, ssurgo, join = st_within)
soil_b
```


## Question 4

**2 points**

Compute the mean `measure` value by SSURGO map unit
and attribute (use `aggregate()`).

Include `mukey` in the grouping variables.

```{r}
aggregate(measure ~ mukey + attribute,
          soil_b,
          FUN = mean)
```



# Part 2: Raster Creation and Spatial Pattern

## Question 5

**1 point**

Load the field boundary (`data/boundary.shp`) and
reproject (using `st_transform`) it to EPSG 32615.
Create a raster grid that covers the boundary and
has a 10 m resolution using `st_as_stars()` and
`st_crop()`.

Plot your empty grid.

```{r}
boundary <- st_read('./data/boundary.shp')
boundary <- st_transform(boundary, 32615)

grd <- st_as_stars(st_bbox(boundary), 
                   dx = 10,
                   dy = 10)

grd <- st_crop(grd, boundary)
plot(grd)
```



## Question 6

**2 points**

Choose the soil attribute `P_bray` from the
original soil test points (not the boundary).
Compute the empirical variogram with `variogram()`
and plot it.

```{r}
soil$attribute |> unique()

# Example: select pH
selected <- soil[soil$attribute == "P_bray", ]
v <- variogram(measure ~ 1, selected)
v
plot(v)
```


## Question 7

**2 points**

Using your selected attribute from Question 6, fit a
spherical model (`"Sph"`) with `fit.variogram()` and
plot empirical + fitted variogram together.

```{r}
m <- fit.variogram(v, vgm("Sph"))
m
plot(v, model = m)
```


## Question 8

**2 points**

Using the fitted model from Question 7, run
`krige()` across the grid and map the predicted
surface (`var1.pred`).

```{r}
kr <- krige(measure ~ 1, selected, grd, model = m)

plot(kr["var1.pred"])
plot(st_geometry(boundary), add = TRUE, border = "black", col = NA, lwd = 1)
plot(st_geometry(selected), add = TRUE, pch = 16, cex = 0.5, col = "grey20")
```



