---
title: "Module 12 Assignment"
subtitle: "AGRON 5130"
author: "Caio dos Santos"
output: pdf_document
---

# Introduction

This assignment applies Module 12 spatial analysis
skills to real field data. You will work with
shapefiles, build raster interpolation workflows,
and create interactive maps for interpretation.

```{r}
library(tidyverse)
library(sf)
library(leaflet)
library(RColorBrewer)
library(stars)
library(gstat)
library(leafem)
```

# Question 1: Shapefile Mapping (8 points)

For this question, use the grid-sampled field at
`data/Paulines_grid_sample.shp`.

## Question 1a (1 point)
Read the dataset and name it `soil_data`.

```{r}

```

## Question 1b (1 point)
Filter rows where `attribute == "Ph"`.

```{r}

```

## Question 1c (1 point)
Create a color palette for `measure` using
`"RdYlGn"`.

```{r}

```

## Question 1d (1 point)
Create a basic leaflet map with circle markers and
`Esri.WorldImagery` background.

```{r}

```

## Question 1e (1 point)
Color-code markers using `fillColor`.

```{r}

```

## Question 1f (1 point)
Adjust marker appearance:

- `fillOpacity = 0.8`
- `weight = 0`
- `radius = 6`

```{r}

```

## Question 1g (1 point)
Add a legend to the map.

```{r}

```

## Question 1h (1 point)
Interpret the spatial pattern of soil pH in the
field. Which third (upper, middle, lower) appears
lowest?

# Question 2: Raster and Kriging (12 points)

## Question 2a (1 point)
Read the boundary shapefile needed to build the
raster framework.

```{r}

```

## Question 2b (2 points)
Create the grid object (`grd`) that will be used for
kriging predictions.

```{r}

```

## Question 2c (1 point)
Filter the point data for `attribute == "P_bray"`.

```{r}

```

## Question 2d (2 points)
Create and plot a variogram model for `P_bray`.

```{r}

```

## Question 2e (2 points)
Generate kriged predictions using `gstat::krige()`.

```{r}

```

## Question 2f (2 points)
Map the kriged surface.

```{r}

```

## Question 2g (1 point)
Add interactivity so clicking the map reveals
predicted values.

```{r}

```

## Question 2h (1 point)
In 2 to 3 sentences, explain one management insight
a grower could obtain from your map.
