---
title: "Multiple Linear and Nonlinear Regression"
subtitle: "AGRON 5130 - Unit 11 Assignment"
author: "Caio dos Santos"
output: pdf_document
---

# Introduction

The goal of this assignment is to help you practice
the full workflow from Module 11:

- interpreting partial effects in multiple regression
- diagnosing multicollinearity and unstable coefficients
- comparing full vs reduced models
- fitting and interpreting a nonlinear growth model
- connecting model choice to biological reasoning

You will complete two parts:

- **Part 1:** multiple linear regression using soybean yield data
- **Part 2:** nonlinear modeling using tomato growth data

Submit a knitted PDF/HTML generated from your R
Markdown file to Canvas.

**Total: 15 points**

# Part 1: Multiple Linear Regression (Thompson Soy)

```{r}
library(ggplot2)
library(car)

thompson_soy <- read.csv("data/thompson_soy.csv")
head(thompson_soy)
```

For this dataset:

- `yield` = soybean yield
- `rain6` = total June precipitation
- `rain7` = total July precipitation
- `temp6` = mean June temperature
- `temp7` = mean July temperature

The goal is to ask which of these weather variables
are associated with yield **after accounting for the
others**.

# Question 1

**1 point**

Fit a multiple linear regression model of `yield` as a
function of all four weather variables:

- `rain6`
- `rain7`
- `temp6`
- `temp7`

```{r}
```

# Question 2

**1 point**

Report model coefficients.

```{r}
```

Write 2-3 sentences interpreting:

1. one precipitation coefficient as a partial effect
2. one temperature coefficient as a partial effect

Use language like: "Holding the other predictors
constant..."

# Question 3

**1 point**

Report model fit statistics from `summary()`:

```{r}
```

# Question 4

**1 point**

Create a scatterplot matrix of model variables.

```{r}
```

Then name one pair of predictors that appears
strongly related.

# Question 5

**1 point**

Create a correlation matrix for the predictor
variables only.

```{r}
```


# Question 6

**1 point**

Describe and interpret the correlations between
variables. Are there any varaibles you would be
concerned about including in the model? Why?


# Question 7

**1 point**

Compute VIF values for the full model.

```{r}
```

Then answer:

1. Which predictor has the largest VIF?
2. Is multicollinearity likely to affect interpretation?

# Question 8

**1 point**

Create a reduced model by dropping one predictor you
believe is least useful or most redundant.

```{r}
```

Compare full vs reduced model using:

1. ANOVA (`anova(reduced, full)`)
2. AIC (`AIC(reduced, full)`)

```{r}
```

In 3-4 sentences, justify which model you would keep
for interpretation.

# Question 9

**1 point**

Check assumptions for your **selected** model using:

1. residuals vs fitted plot (`which = 1`)
2. normal Q-Q plot (`which = 2`)

```{r}
```

Write 2-3 sentences on whether the linear model is
adequate.

# Part 2: Nonlinear Regression (Tomato Growth)

```{r}
tomato <- read.csv("data/tomato_days_volume.csv")
head(tomato)
```

# Question 10

**1 point**

Plot `fruit_volume` as a function of `days`.

```{r}
```

In one sentence, describe the shape you see.

# Question 11

**2 points**

Fit and summarize a logistic nonlinear model:

`fruit_volume ~ SSlogis(days, Asym, xmid, scal)`

```{r}
```

Interpret each parameter in biological terms:

1. `Asym`
2. `xmid`


# Question 12

**1 point**

Create a prediction grid for `days` from 7 to 49 by 1,
then predict `fruit_volume` on that grid.

```{r}
```

# Question 13

**1 point**

Add the fitted logistic curve to your original tomato
scatterplot.

```{r}
```

# Question 14

**1 point**

Fit a polynomial model to the same tomato data:

`lm(fruit_volume ~ days + I(days^2), data = tomato)`

```{r}
```

Then, in 3-4 sentences, compare polynomial vs logistic
for this dataset using:

1. biological interpretability
2. plausibility of curve shape
3. model communication to an agronomy audience


