---
title: "Population Statistics"
subtitle: "AGRON 5130"
author: "YOUR NAME HERE"
output: pdf_document
---


# Introduction

The goal of this assignment is to give you
hands‑on practice working with population‑level
summary statistics and spatial data in R, building
directly on the concepts introduced in this week’s
videos. You will load shapefiles, explore their
structure, visualize agronomic datasets, and
compute basic descriptive statistics that help
characterize variability across a field. Submit a
knitted PDF/HTML generated from your R Markdown
file to Canvas.

# Loading libraries

Remember that, before we are able to use the
functions within the *sf* package, we will need to
load the package. Since this is the first
assignment in which you will need to load an R
package, I will include this chunk of code here
for you. For later assignments, I will trust that
you will remember this.

```{r}
library(sf)
```


# Question 1  
**1 point**  

Load the shapefile
`"data/elkhart_ia_yield_map/elkhart_ia_yield_map.shp"`
using `st_read()` and assign it to an object
called `soybean`.

```{r}

```


# Question 2
**2 point**  

Inspect the structure of the `soybean` data frame
using `head()` and `summary()`.

```{r}

```


# Question 3
**2 points**
Plot the yield map and create a histogram of the
yield values.

```{r}

```


# Question 4
**2 points**

Calculate the **mean**, **median**, and **standard
deviation** of the soybean yield.

```{r}

```


# Question 5
**1 point**

Load the planting map
`"data/moline_il_planting_map/moline_il_planting_map.shp"`
and assign it to an object called moline.

```{r}

```


# Question 6
**1 point**  

Inspect the `moline` data using `head()`.

```{r}

```


# Question 7
**2 points**  

Plot the `"AppldRt"` (actual planting rate) column
and produce a histogram of the planting rate.

```{r}

```


# Question 8
**1 point**  

Calculate the mean actual planting rate.

```{r}

```


# Question 9

**1 point**  
Calculate the minimum actual planting rate.

```{r}

```


# Question 10
**2 points**

Calculate the standard deviation of the actual
planting rate and write 1–2 sentences about what
this indicates for variability in planting rate.

```{r}

```

Write your brief interpretation here:

    (Your answer)



