---
title: "Confidence Interval of Difference"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Introduction

In this unit, we learned that, in the case of
two-treatment experiments, we can test whether the
populations that receive each treatment are
different from each other by testing whether their
difference is different from zero. While we will
often use the **t-test** to calculate the
probability the two populations are equal, given
their measured values, there are times when we may
want to report the confidence interval around
their difference.

These confidence intervals also be can be gained
from the `t.test()` function, but calculating them
manually in these exercises is a good opportunity
to sharpen your R skills and hopefully will give
you are greater conceptual sense of how the
confidence interval works.

To create the confidence interval for the
difference between two populations, there are the
following steps:

1.  Calculate the differences between the
populations in each pair or block.

2.  Calculate the overall mean difference between
the two populations.

3.  Calculate the standard deviation and standard
error of the difference.

4.  Look up the appropriate t-value based on the
desired level of confidence and degrees of freedom

5.  Add and subtract the product of standard error
and t-value from the mean difference.

# Example 1

We will work with the soybean_fungicide data from
the previous exercise:

```{r}
soybean <- read.csv("data/soybean_fungicide.csv")
head(soybean)
```

## Compute Differences

We will use the `reshape()` function from the
previous exercises to calculate a the population
difference within each block.

```{r}
soybean_wide <- reshape(soybean,
                        idvar = "Block",
                        timevar = "Treatment",
                        direction = "wide")

soybean_wide$diff <- soybean_wide$Yield.A - soybean_wide$Yield.B
soybean_wide
```

## Calculate the Mean Difference

This is the mean for the new column, *diff*. As a
reminder, `soybean_wide$diff` tells R to use the
diff column of the soybean_wide data frame.

```{r}
xbar <- mean(soybean_wide$diff)
xbar
```

## Calculate the Standard Error


```{r}
n <- length(soybean_wide$diff)
s <- sd(soybean_wide$diff)/sqrt(n)
s
```


## Compute the t-value

We calculate the t-value to use in our confidence
interval by using the `qt()` function of R.
Remember, the qt function uses two arguments. The
first is the desired level of confidence. We want
a 95% confidence interval, so we will use 0.975.
That will leave 0.025, or 2.5%, beyond the upper
and lower limits of our confidence interval.

The second argument is the degrees of freedom.
Since we have observed ten differences, we have
nine degrees of freedom.

```{r}
t_value <- qt(c(0.025, 0.975), 9)
t_value
```

Our t_value is approximately 2.26.

# Calculate the Confidence Interval

The confidence interval is bound by **upper** and
**lower confidence limits**. The lower limit is
equal to the mean difference minus the product of
the standard error of the difference and t-value.
The upper limit is the mean difference plus the
product of the standard error of the difference.

```{r}
ci <- xbar + (s * t_value)
ci
```

The 95% confidence interval, with 9 df, for the
difference is (-5.21, -1.85). Note the confidence
interval does not include zero, therefore the two
populations (and, thus, the treatments they
received) are significantly different at the
*P\<0.05* level.

Which treatment was greater? Since we subtracted B
from A, the negative confidence interval indicates
B was greater.



Like we saw in the lecture, this entire process
can be summarized using the `t.test()` function.

```{r}
t.test(soybean_wide$Yield.A,
       soybean_wide$Yield.B,
       paired = TRUE)
```


The t.test function allows a certain level of
flexibility too. Let's say, for example, that the
company claims a 5-bushel difference when using
this fungicide.

We could test the following hypothesis:
$$H_0: \mu_1 - \mu_2 = 5 \newline
H_a: \mu_1 - \mu_2 \ne 5$$


```{r}
t.test(soybean_wide$Yield.A,
       soybean_wide$Yield.B,
       mu = -5,
       paired = TRUE)
```