---
title: "Re-Structuring Data"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Restructuring Columns and Rows

As we begin comparing **two treatments**, it
becomes very important to think about **how our
data are organized**.

The way a dataset is structured often determines:
- how easy it is to compute statistics,
- how easy it is to plot results,
- and whether certain tests (like paired t‑tests) are straightforward or frustrating.

In this exercise, we will practice moving data
between **wide** and **long** formats.


# Wide vs Long Data

There are two very common ways to organize data.

### Wide format
Each treatment gets its **own column**.


```{r}
wide <- data.frame(Rep = c(1, 2),
                   Trt_A = c(1, 2),
                   Trt_B = c(5, 6))
wide
```


### Long format

Each **row is one observation** and we include a
column telling us which treatment it came from.

```{r}
long <- data.frame(Rep = c(1, 2, 1, 2),
                   Treatment = c('A', 'A', 'B', 'B'),
                   Yield = c(1, 2, 5, 6))
long
```


In statistics and plotting, we often prefer **long
format**. But for **paired differences**, wide
format can be more convenient.

So we need to know how to move between both.


# Example 1: Create a Wide Dataset

First, let's manually create a simple wide dataset.

```{r}
Rep   <- 1:4
Trt_A <- 1:4
Trt_B <- 5:8

wide_df <- data.frame(Rep, Trt_A, Trt_B)

wide_df
```

Each treatment is currently in its own column.  


# Convert Wide to Long

To convert formats in base R, we use the
**reshape()** function.

Let's take a look at the help page for reshape. 
```{r}
?reshape
```



Ok, now let's convert a wide dataset to a long
dataset

```{r}
long_df <- reshape(wide_df,
                   varying = c("Trt_A", "Trt_B"),
                   v.names = "Yield",
                   timevar = "Treatment",
                   times = c("A", "B"),
                   idvar = "Rep",
                   direction = "long")

long_df
```

Now:
- Treatment names are in one column
- Observations are stacked vertically

This format is often preferred for plotting and
modeling.


# Convert Long to Wide

Sometimes we need to go back to wide format. This
is especially useful when computing **paired
differences**.

```{r}
wide_again_df <- reshape(long_df,
                         idvar = "Rep",
                         timevar = "Treatment",
                         direction = "wide")

wide_again_df
```

---

# Creating New Columns

When the data are wide, we can compute differences
very easily.

```{r}
wide_again_df$Trt_Diff <- wide_again_df$Yield.A - wide_again_df$Yield.B

wide_again_df
```


# Case Study: Soybean Fungicide

Now let's try with a real dataset.

```{r}
soybean <- read.csv("data/soybean_fungicide.csv")
head(soybean)
```

This dataset is currently **long**.

To compute differences between treatments, we
first reshape to wide.

```{r}
soybean_wide <- reshape(soybean,
                        idvar = "Block",
                        timevar = "Treatment",
                        direction = "wide")

soybean_wide
```

Now compute differences.

```{r}
soybean_wide$diff <- soybean_wide$Yield.A - soybean_wide$Yield.B
soybean_wide
```

Notice how easy subtraction becomes once treatments are columns.



# Practice 1

The dataset "darwin.csv" is from a trial by
Charles Darwin in which he compared the heights of
progeny of corn plants that were self-fertilized
and cross-fertilized. Three pairs of plants were
grown per pot.  Measurements are in inches.

```{r}
darwin <- read.csv("data/darwin.csv")
head(darwin)
```

The data are currently in the long format.  We
want to convert it to the wide format, where each
treatment has its heights in a separate column.
Re-structure the type and height columns by adding
the spread function to the code chunk below

```{r}
```

The first few rows of the resulting dataset should
look like:

I	a	23.500	17.375	
I	b	12.000	20.375	
I	c	21.000	20.000	
II	d	22.000	20.000	


# Practice 2

These data are from a trial that compared apple
genotypes.

```{r}
apple <- read.csv('./data/apple_genotype.csv')
head(apple)
```

Calculate the difference between the two apple genotypes (Golden and Redspur).
You will have to generate a wide dataset, and then compute the differences.

```{r}
```
The first two rows should look like:

R1	121.7750	127.5900	-5.81500	
R2	146.7222	126.7625	19.95972	


# Practice 3

Manganese was applied to soybean at the V5 stage near New Baltimore, Ohio.  

```{r}
manganese <- read.csv("data/soybean_manganese.csv")
head(manganese)
```

Convert the data to wide format so the two
treatments are in separate columns
```{r}
```
The first two rows of the resulting dataset should
look like:

1	78.9	81.7		
2	80.9	84.3	

Then calculate the difference between treatments A
and B by subtracting treatment B from treatment A.

```{r}

```

The first couple of rows of the final dataset
should look like:

1	78.9	81.7	-2.8	
2	80.9	84.3	-3.4	


# Practice 4

Data are from a wheat trial where fungicide was
applied during flag leaf stage.

```{r}
wheat <- read.csv("data/wheat_fungicide.csv")
head(wheat)
```

Convert the data to wide format.  The first few rows should look like:

1	95.9	93.8	
2	97.5	92.3	
3	98.8	98.7	

```{r}
```


Calculate the difference between treatments by subtracting Treatment A from Treatment B.  The first few rows should look like:

1	93.8	95.9	-2.1	
2	92.3	97.5	-5.2	
3	98.7	98.8	-0.1	



```{r}


```

