---
title: "Exercise: Machine Learning for Canopy Classification"
output: html_document
---

# Introduction

In this exercise, you'll get hands-on with the
same canopy/soil classification workflow we
discussed in lecture. We'll use real drone imagery
and labeled data. The goal: see how simple rules
and machine learning models work in practice, and
what happens when you try to generalize.

Here's what you'll do:

- Explore the structure of multispectral raster data
- Try a simple ExG threshold rule
- See how well it generalizes
- Train and test a random forest model
- Compare results and reflect on what you find


# Installing relevant libraries

In this module, we will introduce a new library.
To fit random forest models, we will use the
`randomForest` library. Let's begin by intalling
it.

```{r, eval=FALSE}
install.packages('randomForest')
```


# 1. Load and Explore the Data

Let's start by loading the libraries you'll need.
These handle rasters, vectors, and plotting. If
you get an error, check that the package is
installed.

```{r}
library(stars)
library(sf)
library(ggplot2)
```


Now, load the first flight image. This is a
multispectral raster, each pixel has several bands
(not just RGB).

```{r}
img_first <- read_stars("data/2023-07-06_resampled.tif")
```


Let's see what the image looks like in color.
We'll make an RGB composite and plot it. Does it
look like a field?

```{r}
img_rgb <- st_rgb(img_first[,,,c(3,2,1)], stretch = TRUE)
plot(img_rgb, main = "First Flight Image (RGB)")
```


Take a look at the structure of the raster. How
many bands? What are the dimensions?

```{r}
st_dimensions(img_first)
```

# 2. Compute and Visualize ExG

Next, let's compute the Excess Green (ExG) index.
This is a quick way to highlight green vegetation.
The formula is simple: $2G - R - B$.

```{r}
exg <- st_apply(img_first, 
                MARGIN = c("x", "y"), 
                FUN = function(bands) bands[2] * 2 - bands[1] - bands[3])
```


Plot the ExG surface. Where do you see the highest
values? Do they match where you expect canopy?

```{r}
plot(exg,
     main = "Excess Green (ExG) across the field", 
     col = function(n) hcl.colors(n, 'Greens', rev = TRUE))
```


Now, plot a histogram of ExG values. This helps
you pick a threshold to separate canopy from soil.
Where would you draw the line?

```{r}
exg_vals <- as.vector(exg[[1]])
hist(exg_vals, breaks = 80, col = "darkgreen", border = "white", main = "Distribution of ExG Values", xlab = "ExG Value")
```

# 3. Apply ExG Threshold and Evaluate

Pick a threshold for ExG. Try 0.035 to start. This
will be your rule: above = canopy, below = soil.

```{r}
binary_exg <- exg[[1]] > 0.035
```


Plot the result. Do the canopy and soil regions
look reasonable?

```{r}
plot(st_as_stars(binary_exg),
     col = c('chocolate4', 'darkgreen'),
     main = "ExG Threshold Classification")
```


Let's see how well this rule works on new data.
Load the labeled test points for the second
flight. These are your "ground truth" for
evaluation.

```{r}
test_pts <- st_read("data/2023-07-24_testing-set.geoJSON", quiet = TRUE)
test_pts$class <- factor(test_pts$class, levels = c("0", "1"))
```

Let's visualize the second flight

```{r}
img_test <- read_stars("data/2023-07-24_resampled.tif")
img_rgb2 <- st_rgb(img_test[,,,c(3,2,1)], stretch = TRUE)
plot(img_rgb2, main = "Second Flight Image (RGB)")
```


Now, compute ExG for the second flight image. This
is a new date and conditions may have changed. 

```{r}
exg_test <- st_apply(img_test, 
                     MARGIN = c("x", "y"),
                     FUN = function(bands) bands[2] * 2 - bands[1] - bands[3])
plot(exg_test,
     main = "Excess Green (ExG) across the field", 
     col = function(n) hcl.colors(n, 'Greens', rev = TRUE))
```


Apply your threshold to the new image. This gives
you a prediction for every pixel.

```{r}
test_pred <- as.numeric(exg_test[[1]] > 0.035)
pred_raster <- st_as_stars(matrix(test_pred, nrow = dim(exg_test)[1], ncol = dim(exg_test)[2]))
st_dimensions(pred_raster) <- st_dimensions(exg_test)[1:2]
st_crs(pred_raster) <- st_crs(exg_test)
plot(pred_raster,
     col = c('chocolate4', 'darkgreen'),
     main = "ExG Threshold Classification")
```


Extract the predicted class at each test point.
This lets you compare your rule to the ground
truth.

```{r}
pred_extract <- st_extract(pred_raster, at = test_pts)
pred_factor <- factor(ifelse(as.vector(pred_extract[[1]]) == 1, "1", "0"), levels = c("0", "1"))
```


Make a confusion matrix. How many pixels did your
rule get right? Where did it make mistakes?

```{r}
conf_mat <- table(Predicted = pred_factor, Observed = test_pts$class)
print(conf_mat)
```

# 4. Train a Random Forest Model

Now let's try a machine learning model. Load the
randomForest library. If you get an error, install
the package first.

```{r}
library(randomForest)
```


Load the hand-labeled training points for the
first flight. These are the pixels you'll use to
train the model.

```{r}
train_pts <- st_read("data/2023-07-06_training-set.geoJSON", quiet = TRUE)
train_pts$class <- factor(train_pts$class, levels = c("0", "1"))
```


Pull out the RGB band values at each training
point. This is your training data.

```{r}
train_x <- st_extract(img_first, at = train_pts)
train_vals <- as.data.frame(train_x[[1]])
names(train_vals) <- paste0("2023-07-06_resampled.tif.", 1:6)
train_df <- cbind(st_drop_geometry(train_pts), train_vals)
train_df <- train_df[complete.cases(train_df), ]
```


Train a random forest model using just the RGB
bands. This is a classic "supervised learning"
setup: you give the model examples, it learns a
rule.

```{r}
x_cols_rgb <- c("2023-07-06_resampled.tif.1",
                "2023-07-06_resampled.tif.2",
                "2023-07-06_resampled.tif.3")

rf_fit_rgb <- randomForest(x = train_df[, x_cols_rgb],
                           y = train_df$class, 
                           ntree = 300,
                           importance = TRUE)
print(rf_fit_rgb)
```

# 5. Test the Random Forest Model

Now, let's see how the model does on new data.
Prepare the RGB bands from the second flight image
for prediction.

```{r}
rgb_data_test <- img_test[,,,1:3]
dims_test <- dim(rgb_data_test)
pixel_matrix_test <- matrix(NA, nrow = dims_test[1] * dims_test[2], ncol = 3)
pixel_matrix_test[, 1] <- as.vector(rgb_data_test[[1]][,,1])
pixel_matrix_test[, 2] <- as.vector(rgb_data_test[[1]][,,2])
pixel_matrix_test[, 3] <- as.vector(rgb_data_test[[1]][,,3])
```


Use your trained model to predict canopy/soil for
every pixel in the new image.

```{r}
pred_vec_test <- predict(rf_fit_rgb, pixel_matrix_test, type = "response")
pred_numeric_test <- as.numeric(pred_vec_test == "1")
pred_matrix_test <- matrix(pred_numeric_test,
                           nrow = dims_test[1], 
                           ncol = dims_test[2])
pred_test_rgb <- st_as_stars(pred_matrix_test)
st_dimensions(pred_test_rgb) <- st_dimensions(rgb_data_test)[1:2]
st_crs(pred_test_rgb) <- st_crs(rgb_data_test)
plot(pred_test_rgb,
     col = c('chocolate4', 'darkgreen'))
```


Extract the predicted class at each test point,
just like before.

```{r}
pred_extract <- st_extract(pred_test_rgb, at = test_pts)
pred_factor_rf <- factor(ifelse(as.vector(pred_extract[[1]]) == 1, "1", "0"), levels = c("0", "1"))
```


Make a confusion matrix. Did the random forest do
better than your simple rule? Where did it help?
Where did it still struggle?

```{r}
conf_mat_rf <- table(Predicted = pred_factor_rf, Observed = test_pts$class)
print(conf_mat_rf)
```

# 6. Reflection Questions

- How did the ExG threshold and random forest models compare in accuracy?
- What are the strengths and weaknesses of each approach?
- If you wanted to improve the models, what would you try next?

