---
title: "Means Separation and Data Presentation"
subtitle: "AGRON 5130 - Module 8"
author: "Caio dos Santos"
date: "2026-03-11"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    smooth: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


# Introduction

The previous two units focused on the design and
analysis of effects in multiple-treatment trials.
Our focus there was to determine whether treatment
effects explained more of the variation among
individual observations than residual error.

In this unit, we move to the next practical
question:

> Once ANOVA says treatments differ, how do we
decide which means are different from each other?

We will cover two classical means-separation
methods:

- Fisher's Least Significant Difference (LSD)
- Tukey's Honest Significant Difference (HSD)

Then, we will introduce linear contrasts, which
allow us to test targeted questions between groups
of treatments.

Finally, we will discuss how to present means in
tables and figures so results are both accurate
and useful to the reader.


# Case Study

Our case study is a saffron nutrient-rate trial.
Saffron was grown hydroponically under four
nutrient concentrations:

- 1X
- 2X
- 3X
- 4X

The trial was arranged as a randomized complete
block design with 4 blocks. The response variable
is corm number.

```{r}
library(dplyr)
library(ggplot2)

saffron <- read.csv("../exercises/data/saffron.csv")
saffron$block <- as.factor(saffron$block)
saffron$rate <- as.factor(saffron$rate)

head(saffron)
```


# ANOVA First

Before separating means, we fit the ANOVA model:

$$Y_{ij}=\mu + B_i + T_j + BT_{ij}$$

Where:

- $Y_{ij}$ is observed corm number
- $\mu$ is the overall mean
- $B_i$ is block effect
- $T_j$ is nutrient-rate effect
- $BT_{ij}$ is residual error

```{r}
saffron_model <- aov(corm_number ~ block + rate, data = saffron)
anova(saffron_model)
```

The treatment effect (`rate`) is significant. This
is our green light to perform means separation.


# Least Significant Difference

The Least Significant Difference test is direct:

1. Calculate a minimum difference threshold (LSD)
1. Compare every pair of treatment means
1. Any pair with a difference >= LSD is considered
   significantly different

The formula is:

$$LSD_{df,\alpha} = t_{df,\alpha} \times SED$$

This is conceptually very similar to the logic we
used for confidence intervals and two-treatment
comparisons.

```{r}
library(agricolae)

lsd_saffron <- LSD.test(saffron_model, "rate")
lsd_saffron
```


## Reading LSD Output

The output is organized in sections. The most
important are:

- `$statistics`: includes residual mean square,
  degrees of freedom, CV, t-value, and LSD value
- `$means`: treatment means and within-treatment
  descriptive statistics
- `$groups`: letter grouping of means

```{r}
lsd_saffron$statistics
```

```{r}
lsd_saffron$means
```

```{r}
lsd_saffron$groups
```

In the grouping table, treatments sharing a letter
are not significantly different at the chosen
alpha level.


# Comparisonwise vs Experimentwise Error

Why not always use LSD?

With each pairwise comparison run at significance
level $\alpha_C$ (the **comparisonwise error rate**),
the probability of at least one false positive
across the entire experiment grows. For $M$
independent tests, the **experimentwise error rate**
is approximately:

$$\alpha_E \approx 1 - (1 - \alpha_C)^M$$

For $K$ treatment means there are
$M = K(K-1)/2$ possible pairs. With $K = 10$
treatments ($M = 45$ pairs) running each at
$\alpha_C = 0.05$:

$$\alpha_E \approx 1 - (0.95)^{45} \approx 0.90$$

That is nearly a 90% chance of at least one false
positive across the experiment. Conversely, if we
fix $\alpha_E = 0.05$ for the same 45 comparisons,
the per-comparison error drops to roughly
$\alpha_C \approx 0.001$, making it very difficult
to detect modest but real differences.

No procedure can keep both error rates small
simultaneously at a fixed sample size. The choice
of procedure depends on which error matters more
in your specific research context.


## Procedure Comparison

Four widely used procedures and what they control:

| Procedure | Error rate controlled |
|---|---|
| Protected LSD | $\alpha_C$ (partial control of $\alpha_E$) |
| Bonferroni adjusted t | $\alpha_E$ |
| Tukey HSD | $\alpha_E$ |
| Scheffé | $\alpha_E$ |

- If a **false positive** is costly (e.g., recommending
  an inferior treatment to farmers), prioritize
  $\alpha_E$ control with Tukey or Bonferroni.
- If **missing a real difference** is costly (e.g., in
  early screening trials), F-protected LSD preserves
  more power.


# Tukey's Honest Significant Difference

Tukey HSD is similar in structure to LSD, but it
uses the studentized range distribution (`Q`) and
adjusts the threshold based on:

- alpha
- residual degrees of freedom
- number of treatments

So as the number of treatments increases, the
minimum significant difference becomes more
stringent.

```{r}
hsd_saffron <- HSD.test(saffron_model, "rate", group = TRUE)
hsd_saffron
```

```{r}
hsd_saffron$groups
```

Compared with LSD, Tukey tends to be more
conservative, especially when many means are being
compared.

The formula for HSD differs slightly depending on
whether sample sizes are equal or unequal. When
all groups have the same $n$:

$$\text{HSD} = q_{\alpha_E} \sqrt{\frac{MSError}{n}}$$

When sample sizes differ (the **Tukey-Kramer**
procedure), each pair gets its own threshold:

$$\text{HSD}_{ij} = q_{\alpha_E}
\sqrt{\frac{1}{2}\left(\frac{1}{n_i} +
\frac{1}{n_j}\right) MSError}$$

In both cases, $q_{\alpha_E}$ is the critical
value of the studentized range distribution,
based on $\alpha_E$, the residual degrees of
freedom, and $K$ (the number of treatments).


# Bonferroni Adjusted t

The Bonferroni procedure also controls $\alpha_E$
but uses the standard t-distribution with an
adjusted critical level. For $M = K(K-1)/2$
comparisons, each test is conducted at
$\alpha_E / M$, giving the critical difference:

$$\text{BonAdj} = t_{\alpha_E^*}
\sqrt{\left(\frac{1}{n_i} + \frac{1}{n_j}\right) MSError}$$

where $t_{\alpha_E^*}$ is tighter than the LSD
critical value and grows as $M$ increases.

In R, we can apply Bonferroni-adjusted pairwise
comparisons using the `multcomp` package:

```{r}
library(multcomp)

bon_saffron <- glht(
  saffron_model,
  linfct = mcp(rate = "Tukey")
)

summary(bon_saffron, test = adjusted("bonferroni"))
```

Compared to LSD, the Bonferroni critical
difference is larger, so fewer pairs will be
declared significantly different. In small
experiments (few treatments) Bonferroni and Tukey
often give the same groupings. As the number of
treatments grows, Tukey HSD is generally
preferred because the Bonferroni penalty grows
proportionally with $M$ while the studentized
range distribution adjusts more efficiently.


# Scheffé's Procedure

Scheffé's method is the most conservative of the
four procedures. It controls $\alpha_E$ and, unlike
the others, has a formal connection to the ANOVA
F-test: after a significant F-test, it guarantees
that at least one contrast will be declared
significant.

The critical difference for Scheffé is based on
$\sqrt{(K-1)F_{\alpha_E}}$ rather than a t or
studentized-range value, which makes it
substantially larger than the Tukey or Bonferroni
thresholds. When the goal is pairwise mean
comparison, this conservatism is usually
unnecessary.

**When Scheffé is appropriate:**

- Post-hoc exploration where hypotheses were not
  specified before seeing the data
- Situations involving many possible contrasts,
  not just pairwise differences

For routine pairwise comparisons in designed
agricultural experiments, Tukey HSD is preferred.


# Linear Contrasts

LSD and Tukey are pairwise methods. But many
agronomic questions are not pairwise.

Sometimes we want to compare **groups** of
treatments, for example:

- low rates vs high rates
- middle rates vs extremes

This is what linear contrasts are for.


## Contrast Example 1

Question: Is the average response of low rates
(1X and 2X) different from high rates (3X and 4X)?

Coefficients:

- `1X`: +1/2
- `2X`: +1/2
- `3X`: -1/2
- `4X`: -1/2

The coefficients sum to zero, as required.

```{r}
library(multcomp)

levels(saffron$rate)

K_low_vs_high <- matrix(c(1/2, 1/2, -1/2, -1/2), 1)

saffron_model_contrast <- aov(corm_number ~ 0 + rate, data = saffron)
low_vs_high <- glht(saffron_model_contrast, linfct = K_low_vs_high)
summary(low_vs_high)
```


## Contrast Example 2

Question: Are middle rates (2X and 3X) different
from the extremes (1X and 4X)?

Coefficients:

- `1X`: -1/2
- `2X`: +1/2
- `3X`: +1/2
- `4X`: -1/2

```{r}
K_middle_vs_extremes <- matrix(c(-1/2, 1/2, 1/2, -1/2), 1)

middle_vs_extremes <- glht(
  saffron_model_contrast,
  linfct = K_middle_vs_extremes
)

summary(middle_vs_extremes)
```

The estimate gives the size and direction of the
difference between the two treatment groups.


# Means Presentation

Once means are separated, we need to communicate
results clearly.

There are two standard formats:

- Means tables
- Means plots


## How Letters Are Assigned

Before presenting results, it helps to understand
how statistical software assigns letters to means.
The algorithm works as follows:

1. Order sample means from largest to smallest.
2. Compare the largest mean to the smallest. If
   significantly different, that pair gets
   different letters.
3. Work inward, comparing the largest to the
   second-smallest, and so on, until a
   non-significant pair is found.
4. Place a common letter on that group, then
   restart the process using the second-largest
   mean.

**Consistency rule**: if means A and B are not
significantly different, all means *between* them
in the ordered list are also declared not
significantly different without further testing.
This prevents the logical contradiction where X
and Z are not different but X and Y (between them)
are.

In most applications the consistency rule applies
cleanly. When sample sizes are unequal and
multiple critical differences exist, rare apparent
contradictions can arise — these should be
reported in a table footnote.

The practical message: **means sharing a letter
are not significantly different** at the chosen
$\alpha$.


## Means Table

A practical means table usually includes:

- treatment mean
- replication count (`N`)
- SED or SE
- LSD or HSD value (with alpha)
- ANOVA treatment p-value

```{r}
means_tbl <- saffron %>%
  group_by(rate) %>%
  summarise(
    corm_number = mean(corm_number),
    N = n(),
    .groups = "drop"
  ) %>%
  left_join(
    tibble::rownames_to_column(lsd_saffron$groups, var = "rate"),
    by = "rate"
  ) %>%
  arrange(desc(corm_number))

means_tbl
```


## Plotting Means

For categorical treatments, use bar plots or point
plots with error bars.

If you are presenting LSD/HSD groupings, letters
can be added above bars to indicate significance
groups.

```{r}
plot_df <- means_tbl %>%
  mutate(letter_y = corm_number + 8)

ggplot(plot_df, aes(x = rate, y = corm_number)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_text(aes(y = letter_y, label = groups), size = 5) +
  labs(
    x = "Nutrient Rate",
    y = "Corm Number"
  ) +
  theme_linedraw()
```

This gives the reader both the treatment means and
a direct visual guide to which means differ.


# Conclusion

In this unit, we moved from ANOVA significance to
interpretable treatment comparisons.

Key takeaways:

- Use ANOVA to test whether treatment effects exist
- Use LSD (preferably F-protected) for pairwise
  means separation
- Use Tukey HSD when you need stronger control of
  experimentwise error
- Use linear contrasts to answer focused,
  biologically meaningful group comparisons
- Present means with enough supporting statistics
  that others can interpret and reuse your results

These tools are not just statistical procedures;
they are how we convert trial results into useful
agronomic decisions.



```{r}

true_means <- c(195, 202, 215, 205, 198, 193)
mu <- mean(true_means)


block_effects <- c(`1` = -50,
                   `2` = -50,
                   `3` = 0,
                   `4` = 0,
                   `5` = 0)

rcbd_layout$yield <- 
  mu +
  true_means[rcbd_layout$trt] - mean(true_means) +
  block_effects[rcbd_layout$block] +
  rnorm(30, mean = 0, sd = 8)
```

Let's take a look at the data from the randomized
complete block trial. One of the things that stand
out is that for every treatment, there are two
points below the mean. These two points correspond
to to the two water logging-prone blocks.

```{r}
ggplot(rcbd_layout, aes(x = trt, y = yield)) +
  geom_point(pch = 21, size = 2) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "point", shape = 17, size = 3) +
  coord_flip() +
  theme_linedraw() +
  labs(x = "Fungicide",
       y = "Corn Yield (bu/ac)")
```


## The linear additive model

One important thing that we can think about is
how, by accounting for soil variability using a
blocked structure, we explicitly remove the soil
variability from the residuals ($\epsilon_ij$).
Our linear additive model now includes a block
effect, as we can see below:

$$Y_{ij}=\mu + \tau_i + B_j + \epsilon_{ij}$$

Where, 

$Y_{ij}$ = observed yield 

$\mu$ = overall mean yield

$B_j$ = block effect

$\tau_i$ = fungicide treatment effect 

$\varepsilon_{ij}$ = random error


What does this mean for the ANOVA? 

Below, we can see the result of the analysis of
variance table. Check how much of the MSE is
allocated in the block effect. All of that error,
if left unaccounted for, would inflate our
residual MSE, which would decrease our $F$-value.

```{r, echo=TRUE}
lm3 <- lm(yield ~ trt + block, data = rcbd_layout)
anova(lm3)
```

We can see that blocking the design is meant to
account for pre-existing variability. The main
assumption here is that experimental units within
a block are exposed to uniform conditions.

By modeling the block effect explicitly, we
prevent environmental variability from inflating
the residual error. This increases the sensitivity
of the experiment and improves our ability to
detect treatment differences.


## A brief digression about Fixed and Random Effects

Before we move forward, it is important to briefly
discuss the difference between **fixed effects** and
**random effects**. This distinction becomes important
when we start working with blocked experiments
and, later on, with more complex experimental
designs.

In agricultural experiments, **fixed effects** are
factors where we are interested in comparing the
specific levels included in the experiment. In
other words, the treatments we include are the
ones we want to draw conclusions about.

For example, in the fungicide trial we have been
simulating, the treatments are the five
fungicides:

> F1, F2, F3, F4, F5, and F6

In this case, these fungicides are not a random
sample of all possible fungicides in the world.
Instead, they are specific products that we
intentionally selected because we want to compare
their performance. Because of that, fungicide
treatment is considered a **fixed effect**.


**Random effects** are different. A **random effect**
represents a source of variability where the
individual levels themselves are not of primary
interest. Instead, the levels are viewed as a
sample from a larger population of possible
levels.


When a statistical model includes both **fixed
effects** and **random effects**, it is called a mixed
model.

This means that, if I do not intend to make any
statements about the effect of block itself, but
just want to account for that variability, we can
model the block effect as a **random effects** Here,
we do not intend to make a statement about how
waterlogging will affect yield in this trial, all
we want to do is to account for the block
variability.

If we treat blocks as a **random effects** we assume
that the blocks in our experiment represent a
random sample of possible field conditions.


#### Modeling block as a **random effects**

In this case, the linear additive model looks like
this:

$$Y_{ij}=\mu + \tau_i + b_j + \epsilon_{ij}\newline b_j \sim N(0, \sigma^2_{block})$$

Where, 

$Y_{ij}$ = observed yield 

$\mu$ = overall mean yield

$b_j$ = random effect of block

$\tau_i$ = fungicide treatment effect 

$\varepsilon_{ij}$ = random error



To fit mixed effect models in R, we will have to
use a package. There's no way around it. There are
two main packages for this `lme4` and `nlme`.
Here, we will use `nlme` for simplicity.

```{r, echo=TRUE}
library(nlme)
model_mixed <- lme(yield ~ trt, 
                   random = ~ 1 | block, 
                   data = rcbd_layout)

anova(model_mixed) 
```

If we want to see how much variability is
attributed to the **random effects**, we can examine
the variance components estimated by the model.

```{r}
VarCorr(model_mixed)
```


Plots within the same block tend to experience
similar environmental conditions. Because of that,
their yields are often more similar to each other
than to plots located in other parts of the field.
In other words, observations within a block are
not completely independent.

If we ignore this structure and treat every
observation as independent, some of that shared
block variability ends up inside the residual
error term. This inflates the residual variance
and can make it harder to detect treatment
differences.

By including block as a random effect, the model
explicitly accounts for this shared variability.
Instead of being treated as unexplained noise, the
block-to-block variation is separated from the
residual error.


# Factorial Design

Now, in agronomy we are often interested in more
than one factor at a time. We have just looked at
a simple example that investigated 6 fungicide
treatments.

Let's now imagine that we have another factor in
which we are interested as well. We want to know
wether the fungicide treatments behave differently
with irrigation. In this case, we are interested
in **two** factors. 

Since we want to evaluate the effect of fungicide,
the effect of irrigation, and whether the effect
of fungicide is different when there is
irrigation, we will set up a **factorial design**
with two factors.

From a design perspective, factorial experiments
allow us to estimate both the main effects and the
interaction between factors, because every
combination of the factor levels appears in the
experiment. This keeps our assumptions aligned
with reality: we model the interactions we
observe, maintain independent errors, and can
properly test all effects.



## Design example with two factors

We will keep using the same trial location, with
30 experimental units. To do so, and still include
another level, I will choose fewer fungicide
treatments. Let's pick treatments **F1, F3, and
F4**.

Now, we can include another factor: with or
without irrigation.

Since we have 3 fungicides and 2 irrigation
levels, the factorial design produces 3 × 2 = 6
treatment combinations.


```{r, echo=TRUE}
fungicide_treatments <- c('F1', 'F3', 'F4')
irrigation_treatments <- c('Yes', 'No')
factorial_treatments <- expand.grid(fungicide = fungicide_treatments,
                                    irrigation = irrigation_treatments)
factorial_treatments
```

```{r}
set.seed(20260307)
# Define blocks
block <- factor(rep(1:5, each = 6))

# Define fungicide and irrigation treatments
fungicide <- factor(rep(c("F1", "F3", "F4"), times = 2 * 5 ))
irrigation <- factor(rep(c("Yes", "No"), each = 3, times = 5))

# Combine into a data frame
trial_layout <- data.frame(block = block,
                           fungicide = fungicide,
                           irrigation = irrigation)

# Randomize within each block
for(b in levels(trial_layout$block)) {
  rows <- trial_layout$block == b
  trial_layout[rows, c("fungicide", "irrigation")] <- trial_layout[rows, c("fungicide", "irrigation")][sample(1:6), ]
}

trial_layout$label <- paste(trial_layout$fungicide, trial_layout$irrigation, sep = ' & ')
eus$trt <- trial_layout$label
```

Let's arrange our treatments in 5 blocks, like
before, and let's randomize the treatments within
block.

Below, the graph shows that every block contains a
treatment that corresponds to a combination of our
factors.


```{r}
ggplot()+
  geom_sf(data = eus, aes(fill = trt), col = 'black')+
  geom_sf(data = field, fill = 'transparent', col = 'black')+
  geom_sf_text(data = eus, aes(label = trt))+
  theme_void()+
  scale_fill_manual(values = hcl.colors(6, 'Dynamic'))+
  labs(fill = 'Treatment')
```


```{r}

fungicide_means <- c(F1 = 195, F3 = 215, F4 = 205)

irrigation_effects <- c(Yes = 0, No = 25) 

interaction_effects <- matrix(
  c(0, 5,  
    0, 25, 
    0, 5), 
  nrow = 3, 
  byrow = TRUE,
  dimnames = list(c("F1", "F3", "F4"), c("Yes", "No"))
)

block_effects <- c(`1` = -50, `2` = -50, `3` = 0, `4` = 0, `5` = 0)

trial_layout$yield <- with(trial_layout,
                           205 +
                           fungicide_means[fungicide] - mean(fungicide_means) +
                           irrigation_effects[irrigation] +
                           mapply(function(f, i) interaction_effects[f, i], fungicide, irrigation) +
                           block_effects[block] +
                           rnorm(30, mean = 0, sd = 8)  
)

factorial_layout <- trial_layout
names(factorial_layout)[4] <- 'trt'
```


Let's take a look at the data from this
experiment. On the Y axis, we have the different
fungicide-irrigation combinations.

Comparing within the same fungicide, it seems like
on average, irrigation increased yield as well.

Now, is the effect of irrigation and fungicide
different for the different combinations? That's
what the interaction term of the model will answer.

```{r}
ggplot(factorial_layout, aes(x = trt, y = yield)) +
  geom_point(pch = 21, size = 2) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "point", shape = 17, size = 3) +
  coord_flip() +
  theme_linedraw() +
  labs(x = "Fungicide & Irrigation",
       y = "Corn Yield (bu/ac)")
```

We can visualize the relationship between out
treatments and yield using what is called an
interaction plot.

Interaction plots provide a useful visual way to
detect interactions. If the lines are roughly
parallel, it suggests that the effect of one
factor is consistent across the levels of the
other factor. When the lines are not parallel,
this indicates a possible interaction between the
factors.

```{r}
ggplot(factorial_layout,
       aes(x = irrigation, y = yield,
           color = fungicide,
           group = fungicide)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point", size = 3) +
  theme_linedraw() +
  coord_flip()+
  labs(x = "Irrigation",
       y = "Corn Yield (bu/ac)",
       color = "Fungicide")
```


## The linear additive model

Now, our model is getting a little more
complicated, right? The simple "treatment" effect
($\tau_i$), has been replaced by $F$ and $I$,
which indicate the effects of our two treatments.
In addition, we also have the interaction term
($FI_{ij}$), which indicates whether fungicide and
irrigation treatments affect each other.

In this case, I kept block as a random effect, but
I could just as easily have included block as a
**fixed effect**. I made this choice based on the fact
that we do not want to make any statements about
the two first blocks, which are
waterlogging-prone. We want to include block in
the model to account for this systematic
variability coming from the different soil
conditions.

$$Y_{ijk}=\mu + F_i + I_j + FI_{ij} +  b_k + \epsilon_{ijk}\newline b_k \sim N(0, \sigma^2_{block})$$

Where, 

$Y_{ijk}$ = observed yield 

$\mu$ = overall mean yield

$F_i$ = fungicide treatment effect 

$I_j$ = irrigation treatment effect 

$FI_{jk}$ = interaction between fungicide and irrigation  

$b_k$ = random effect of block

$\varepsilon_{ijk}$ = random error

## Fitting this model in R

Just like before, we will use the `lme()`
function, which is design to fit linear mixed
effects models.

```{r, echo=TRUE}
model_fac <- lme(yield ~ fungicide * irrigation, 
                 random = ~ 1 | block, 
                 data = factorial_layout)
anova(model_fac)

```

The ANOVA table output shows us strong evidence
that both the fungicide and irrigation treatments
have an effect on corn yield. Furthermore, the
interaction row (fungicide:irrigation) shows us
that our fungicide treatments behave differently
whether there's irrigation or not.




# Split-Plot Design

Up to this point, we have been assuming that all
treatments can be randomized at the **plot level**.
In other words, we assumed that each experimental
unit can independently receive any treatment.

However, in real agricultural experiments this is
not always practical.

For example, in the factorial experiment we just
constructed, we included an irrigation factor with
two levels: **Yes** and **No**. In our simulated
design we randomly assigned irrigation treatments
to individual plots.

In practice, however, irrigation is often
difficult to apply on a plot-by-plot basis. It may
be much easier to irrigate **larger strips of
land** rather than individual plots.

To account for this practical limitation, we can
use a **split-plot design**.

In a split-plot experiment, one factor is applied
to **large experimental units** (called *main
plots* or *whole plots*), while another factor is
applied to
**smaller units within them** (called *subplots*).

In our case:

- **Irrigation** will be the **main plot factor**
- **Fungicide** will be the **split-plot factor**

## Constructing the split-plot layout

We will keep using our field with **5 blocks** and
**30 experimental units**.

Instead of assigning irrigation at the plot level,
we will assign irrigation to **large strips within
each block**.

Each block will contain:

- **one irrigated strip**
- **one non-irrigated strip**

Each strip will contain the three fungicide
treatments.

This means that fungicides will still be
randomized, but **only within irrigation strips**.

The graph below shows that the irrigation strips
start at the left side of each block and run to
the middle of the block.

```{r}
eus$irrigation <- 'Yes'
eus$irrigation[c(4:6, 10:12, 16:18, 22:24, 28:30 )] <-  'No'
ggplot()+
  geom_sf(data = eus, aes(fill = irrigation), col = 'black')+
  geom_sf(data = field, fill = 'transparent', col = 'black')+
  geom_sf_text(data = eus, aes(label = id))+
  theme_void()+
  scale_fill_manual(values = hcl.colors(2, 'Terrain', rev = TRUE))+
  labs(fill = 'Irrigation')
```


```{r}
set.seed(20260305)
# Define blocks
block <- factor(rep(1:5, each = 6))

# Define irrigation strips (main plots)
irrigation <- factor(rep(rep(c("Yes", "No"), each = 3), times = 5))

# Define fungicide treatments (subplot factor)
fungicide <- factor(rep(c("F1", "F3", "F4"), times = 10))

# Combine into layout
splitplot_layout <- data.frame(block,
                               irrigation,
                               fungicide)


for(b in levels(splitplot_layout$block)) {
  
  for(i in levels(splitplot_layout$irrigation)) {
    
    rows <- splitplot_layout$block == b &
            splitplot_layout$irrigation == i
    
    splitplot_layout$fungicide[rows] <-
      sample(splitplot_layout$fungicide[rows])
  }
}
```

```{r}
fungicide_means <- c(F1 = 195, F3 = 215, F4 = 205)

irrigation_effects <- c(Yes = 0, No = 25)

interaction_effects <- matrix(
  c(0, 5,
    0, 25,
    0, 5),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(c("F1", "F3", "F4"), c("Yes", "No"))
)

block_effects <- c(`1` = -50,
                   `2` = -50,
                   `3` = 0,
                   `4` = 0,
                   `5` = 0)

splitplot_layout$yield <- with(splitplot_layout,
  205 +
  fungicide_means[fungicide] - mean(fungicide_means) +
  irrigation_effects[irrigation] +
  mapply(function(f, i) interaction_effects[f, i],
         fungicide, irrigation) +
  block_effects[block] +
  rnorm(30, mean = 0, sd = 8)
)
```



```{r}
splitplot_layout$trt <- paste(splitplot_layout$fungicide,
                              splitplot_layout$irrigation,
                              sep = " & ")

eus$trt <- splitplot_layout$trt
```


In the figure below, irrigation treatments form
large strips within each block, while fungicides
are randomized within those strips.

```{r}
ggplot()+
  geom_sf(data = eus, aes(fill = trt), col = 'black')+
  geom_sf(data = field, fill = 'transparent', col = 'black')+
  geom_sf_text(data = eus, aes(label = trt))+
  theme_void()+
  scale_fill_manual(values = hcl.colors(6, 'Dynamic'))+
  labs(fill = 'Treatment')
```

Notice that irrigation treatments are applied to
groups of plots, not individual plots. This is
what creates the split-plot structure and requires
a different statistical model.


```{r}
ggplot(splitplot_layout,
       aes(x = paste(fungicide, irrigation, sep = '&'),
           y = yield)) +
  geom_point(pch = 21, size = 2) +
  stat_summary(fun.data = mean_se,
               geom = "errorbar",
               width = 0.2) +
  stat_summary(fun = mean,
               geom = "point",
               shape = 17,
               size = 3) +
  coord_flip() +
  theme_linedraw() +
  labs(x = "Fungicide & Irrigation",
       y = "Corn Yield (bu/ac)")
```



## The linear additive model

Split-plot experiments introduce multiple
sources of variability.

Because irrigation is applied to large plots, the
experimental units receiving irrigation are not
independent at the same level as the fungicide
subplots.

We therefore include two sources of random variation:

1. Block-to-block variability
2. Main-plot variability (irrigation strips within blocks)

The linear model becomes:

$$Y_{ijk}=\mu + F_i + I_j + FI_{ij} +  b_k + m_{jk}+ \epsilon_{ijk}\newline b_k \sim N(0, \sigma^2_{block})\newline m_{jk} \sim N(0, \sigma^2_{strip})$$


Where:

$Y_{ijk}$ = observed yield

$\mu$ = overall mean

$F_i$ = fungicide effect

$I_j$ = irrigation effect

$FI_{ij}$ = interaction

$b_k$ = random block effect

$m_{jk}$ = random main plot effect (irrigation within block)

$\varepsilon_{ijk}$ = subplot error


Fitting the split-plot model in R

```{r,echo=TRUE}
model_split <- lme(
  yield ~ fungicide * irrigation,
  random = ~ 1 | block/irrigation,
  data = splitplot_layout
)

anova(model_split)
```
The nested random structure

block / irrigation

tells the model that irrigation treatments are
applied within blocks, and that the fungicide
subplots share variability within those main
plots.

This accounts for the fact that irrigation was not
randomized at the plot level.



# Conclusion

Experimental design of multiple treatment
experiments plays a critical role in the quality
of our data and the inferences that can be made.

**Randomized Complete Block Design**: In most cases,
you will probably run RCBD experiments where
blocking will reduce the error among plots within
each block, making it easier to detect true
treatment differences.

**Factorial Design**: The factorial design allows
greater efficiency if you are interested in
studying multiple factors simultaneously. It is
also the only way to study the interactions
between factors – and these interactions are often
the most important findings.

**Split-Plot Design**: Where one factor is known to
be more subtle (and therefore harder to test for
significance) than a second factor, or where
differences in equipment size make it more
practical to nest one factor inside the other, the
split-plot design can be very useful.

## A Common Thread

Behind every experimental design choice lies a
single principle: we must **model all known
sources of variation** in order to preserve the
independence assumption that underlies ANOVA.
Whether we are blocking to account for
environmental gradients, including interaction
terms to model how factors work together, or using
separate error terms for nested factors, we are
explicitly accounting for systematic patterns so
that what remains (error) is truly independent and
random.

When you choose an experimental design, you are
essentially asking: what sources of variation
exist in my experimental environment and in how I
can apply treatments? Model those sources. What
remains becomes your error term. If your model is
well-designed, errors will be independent, your
statistical tests will be valid, and you will have
a much better chance of detecting real treatment
effects.

Each design serves a different purpose, and
choosing the right design is critical to the
success and relevance of your research.
