---
title: "Linear Additive Model"
output: html_document
---

# Introduction

The linear additive model may seem a very esoteric
concept in data science, but it is critical to
understand how we make predictions from
statistical models, and how we test whether those
models significantly explain the variance in our
observations. It also underscores an important
concept in statistical modelling: the assumption
that the different effects that combine to explain
an observed value are *additive*.

# Case Study: Barley Effects

We will look at the barley dataset, as we have
before. Except that, for simplicity, we will only
look at the 10 first plots.

```{r}
barley <- read.csv('./data/barley.csv')
barley <- head(barley, n = 10)
```


Before diving into the data analysis, I always
find it useful to look at the data. It helps me
better understand what kind of differences I want
to see.

Like I mentioned before, ggplot2 is really useful
for this. Let's see the dataset:

```{r}
library(ggplot2)
ggplot(barley)+
  geom_point(aes(x = yield, y = gen), pch = 1)+
  labs(x = 'Yield (bu/ac)', y = 'Genotype')+
  stat_summary(aes(x = yield, y = gen),
               pch = 17,
               fun.data = 'mean_se')+
  theme_bw()
```




## The linear aditive model

Recall that the linear additive model looks like this:

$$Y_{ij} = \mu + \tau_i + \epsilon_{ij}$$ 
$Y_{ij}$ = observed yield 
$\mu$ = overall mean yield
$\tau_i$ = treatment effect (old or new)
$\epsilon_{ij}$ = random error



Below, we will load the same data set, except that
I have already computed the linear additive model
parameters/effects so that we can better visualize
these effects.

```{r}
barley_effects <-  read.csv("data/barley_effects_table.csv")
barley_effects
```


## Plotting the Effects

Below is the code to create a bar plot.  This kind
of bar plot is *stacked*: it shows multiple
measures for each observation (plot), stacked on
top each other. Normally, a stacked bar plot can
be created in a couple of lines of code. The
intricate code below was so I could customize the
order of stacking for each plot.  Run the code and
observe the plot.

```{r, fig.width=12, fig.height=10}
ggplot(barley_effects) +
  geom_segment(aes(x=plot, y=0, xend=plot, yend=mu, col = 'mu'), linewidth=14) +
  geom_segment(aes(x=plot, xend=plot, y=trt_bar_min, yend=trt_bar_max, col = 'tau'), linewidth=14) +
  geom_segment(aes(x=plot, xend=plot, y=err_bar_min, yend=err_bar_max, col = 'error'), linewidth=14) +
  geom_text(aes(x=plot, y=mu/2, label= mu)) +
  geom_text(aes(x=plot, y=trt_bar_min+(trt_bar_max-trt_bar_min)/2, label=trt_effect)) +
  geom_text(aes(x=plot, y=err_bar_min+(err_bar_max-err_bar_min)/2, label=error_effect)) +
  geom_segment(aes(x=plot+0.4, xend=plot+0.2, y=y, yend=y), arrow = arrow(length = unit(0.01, "npc")), linewidth=1) +
  labs(y = 'Yield', col = 'Effects')+
  geom_text(aes(x=plot+0.4, y=y, label=y), hjust=0) +
  scale_x_continuous(breaks=c(1:10))+
  scale_color_manual(values = c('mu' = 'lightblue',
                                'tau' = 'lightgreen',
                                'error' = 'tomato'))+
  theme_bw()+
  theme(legend.position = 'top')
```

If this plot is too big to fit comfortably on this
screen, remember you can knit this code using the
menu immediately above this window, and either
view it in HTML or print it out to use.

What I have tried to do with this plot is to
visualize the additive effects of the population
mean (light blue bar), treatment effect (green
bar), and error effect (red bar).  The sum of
their effects (the observed value) is shown by a
black arrow to the right of each bar.

The blue bars are the same, 312.3, reflecting that
the population mean for the whole trial does not
change: every plot starts out with this same
value, with the effects then adding or subtracting
from its observed value.

Looking at this plot, you can see half of the
green bars have a treatment effect of 26.8, while
the other half have a treatment effect of -26.8.
In a two treatment trial, the treatment effects
will be exactly opposite for the two levels of
treatment.  If we look at the plot above, we can
see the positive treatment effect is associated
with the new genotype, and the negative effect
with the old genotype.

The red bars represent the error.  This is the
unexplained variance in observed value.  Part of
it reflects how, try as hard as we might, not
every plot has exactly the same environment.  Soil
characteristics and microclimates differ.  Plot
lengths may differ slightly among plots.  One plot
may be planted more evenly than others.  A
fertilizer applicator might skip.  An ear might
bounce out of the combine.

Note that the errors are negative in
some cases and positive in others.  In addition,
they are not consistent -- their values are
random.


## Examining Individual Plots

So let's plug a few plots into this equation and
see how it works.

In plot 2: 
- Yij = 223.9
- mu = 312.3, 
- Ti, associated with the old hybrid = -26.8
- e(i)j, associated with plot 2, = -61.6

Our model for plot 2, then is:  223.9 = 312.3 - 26.8 - 61.6


What about for plot 5?

Yij = 399.3
mu = 312.3
- Ti, associated with the new hybrid = +26.8
- e(i)j, associated with plot 5 =  +60.2

Our model for plot 5, then is:  399.3 = 312.3 + 26.8 + 60.2


What about for plot 8?

Yij = 349.8
mu = 312.3
- Ti, associated with the old hybrid = -26.8
- e(i)j, associated with plot 8 =  +64.3

Our model for plot 8, then is:  349.8 = 312.3 - 26.8 + 64.3

What is interesting in this last example is that
the positive error "masks" the negative effect of
the treatment -- the yield with the old hybrid, in
this case, is greater than plots 3 and 9, which
received the old hybrid!  This underscores why it
is important to break down the effects behind an
observation to better understand the true
contribution of each treatment level.


## Practice: Groudnut

Let's load and plot the groundnut data from the
other exercises this unit.  Like the barley data
above, we have added several columns in order to
calculate treatment and error effects and to draw
the plot below.

```{r}
groundnut_effects <-  read.csv("data/groundnut_effects_table.csv")
groundnut_effects
```
## Plotting the Effects

Below is the code to create a bar plot.  This kind
of bar plot is *stacked* -- it shows multiple
measures for each observation (plot), stacked on
top each other.  Normally, a stacked bar plot can
be created in a couple of lines of code -- the
intricate code below was so I could customize the
order of stacking for each plot.  Run the code and
observe the plot.

```{r, fig.width=12, fig.height=10}
ggplot(groundnut_effects) +
  geom_segment(aes(x=plot, y=0, xend=plot, yend=mu, col = 'mu'), linewidth=14) +
  geom_segment(aes(x=plot, xend=plot, y=trt_bar_min, yend=trt_bar_max, col = 'tau'), linewidth=14) +
  geom_segment(aes(x=plot, xend=plot, y=err_bar_min, yend=err_bar_max, col = 'error'), linewidth=14) +
  geom_text(aes(x=plot, y=mu/2, label= mu)) +
  geom_text(aes(x=plot, y=trt_bar_min+(trt_bar_max-trt_bar_min)/2, label=trt_effect)) +
  geom_text(aes(x=plot, y=err_bar_min+(err_bar_max-err_bar_min)/2, label=error_effect)) +
  geom_segment(aes(x=plot+0.4, xend=plot+0.2, y=y, yend=y), arrow = arrow(length = unit(0.01, "npc")), linewidth=1) +
  labs(y = 'Yield', col = 'Effects')+
  geom_text(aes(x=plot+0.4, y=y, label=y), hjust=0) +
  scale_x_continuous(breaks=c(1:10))+
  scale_color_manual(values = c('mu' = 'lightblue',
                                'tau' = 'lightgreen',
                                'error' = 'tomato'))+
  theme_bw()+
  theme(legend.position = 'top')
```

Model plots 1, 5, and 6, using the linear model as
we did above.

