---
title: "One-Sided Hypotheses"
output:
  html_document
---

# Introduction

We learned in the lesson there are times when it
is appropriate to use a one-sided hypothesis.  A
one-sided hypothesis specifies how two treatments
will rank in a trial, for example that variety B
will have greater yield than variety A:

$$H_0: \mu_{A} \geq \mu_{B} \newline  
H_a: \mu_{A} < \mu_{B} $$

A two-sided hypothesis, in contrast, only
specifies that variety A and variety B will be
different:

$$H_0: \mu_{A} = \mu_{B} \newline  
H_a: \mu_{A} \neq \mu_{B} $$


# Case Study: Groundnut

In this study, the wet weight of groundnut, in
kg/plot, was measured for two genotypes, coded A
and C.  The plots were paired.

```{r}
library(ggplot2)
groundnut <- read.csv("data/groundnut.csv")
head(groundnut)

ggplot(groundnut)+
  geom_point(aes(x = wet, y = gen), pch = 1)+
  labs(x = 'Yield (bu/ac)', y = 'Treatment')+
  stat_summary(aes(x = wet, y = gen),
               pch = 17,
               fun.data = 'mean_se')+
  theme_bw()

```

# One-Sided T-Test

In the last unit, we learned to use the *t.test()*
function to conduct a paired two-sided t-test.
Let's first analyze the groundnut data that way.

Our hypotheses are 

$$H_0: \mu_{A} = \mu_{C} \newline  
H_a: \mu_{A} \neq \mu_{C} $$

```{r}
groundnut_wide <- reshape(groundnut,
                          timevar = 'gen',
                          idvar = 'block',
                          direction = 'wide')

t.test(groundnut_wide$wet.A, 
       groundnut_wide$wet.C, 
       paired=TRUE)
```

We see we have a p-value of 0.0649.  The two
genotypes do not produce different wet weights of
groundnuts at the p < 0.05 level of significance.



Let's say our hypotheses are these:

$$H_0: \mu_{A} \leq \mu_{C} \newline  
H_a: \mu_{A} > \mu_{C} $$



Since A comes before C, a will compute:

$$\mu_A - \mu_C > 0$$

Therefore, it's important to specify the
alternative hypothesis accordingly:

```{r}
t.test(groundnut_wide$wet.A, 
       groundnut_wide$wet.C, 
       paired=TRUE,
       alternative = 'greater')
```

We can now see the p-value for the test is 0.03 --
genotype A produces a greater wet weight of
groundnut than genotype C at the p < 0.05 level of
significance.

# Practice: Barley

In this study, yield of a new and old genotype
were compared.  Treatments were paired.

```{r}
barley <-  read.csv("data/barley.csv")
```

1) Restructure the data to the wide format

2) Run a two-sided t.test to compare the yield of
the two genotypes. You should get a p-value of
2.158e-06.

3) Run a one-sided t.test to test the hypothesis
the new hybrid yields greater than the old hybrid.
Going by alphabetical order, R will subtract the
mean of "old" from "new".  Given our hypothesis
that the yield of the "new" genotype will be
greater than that of the "old", our difference
will again be positive.  Again, use the
"alternative="greater"* argument with the
*t.test()* function.  Your answer should have a
p-value = 1.079e-06.


# Practice: Strawberry

The yield of two strawberry genotypes was tested
in a paired treatment design.

```{r}
strawberry <-  read.csv("data/strawberry.csv")
```

1) Restructure the data to the wide format

2) Test the difference between genotypes using a
two-sided test.  You should get p-value = 0.055.

3) Test the hypothesis that genotype F is greater
than genotype R1.  Since R will subtract R1 from
F, our difference will be positive.  You should
get p-value = 0.0275.
