---
title: "Sample Statistics"
subtitle: "AGRON 5130"
author: "YOUR NAME HERE"
output: pdf_document
---


# Introduction

The goal of this assignment is to give you hands-on
practice working with **sample statistics** and
**confidence intervals**, building directly on the
concepts introduced in this week’s videos. You will
work with real agronomic datasets, compute sample-based
summary statistics, quantify uncertainty using the
standard error, and construct confidence intervals
using the *t* distribution.

Submit a knitted PDF or HTML generated from your
R Markdown file to Canvas.


# Question 1

**1 point**

Load the dataset
`"data/millet_uniformity.csv"`
and examine the first six rows using the `head()`
function.

Pearl millet yield was sampled at 160 locations
within a field in Kundewadi, Niphad, in the district
of Nasik, India.

```{r}
```


# Question 2

**1 points**

Calculate the **mean**, **standard deviation**, and
**number of observations** for millet yield.

```{r}
```


# Question 3

**1 points**

Calculate the **standard error of the mean** for
millet yield.

```{r}
```


# Question 4

**2 points**

Calculate the *t*-value (using the `qt()` function)
needed to construct a **95% confidence interval**
for the mean yield.

A 95% confidence interval leaves 2.5% of the
probability in each tail, so use `0.975`. Remember
that the degrees of freedom are one less than the
number of observations.

```{r}
```


# Question 5

**2 points**

Calculate the **95% confidence interval** for millet
yield using the statistics computed above.

```{r}
```


# Question 6

**1 point**

Load the dataset
`"data/safflower_uniformity.csv"`
and examine the first six rows using the `head()`
function.

Safflower yield was sampled 340 times within a
safflower field in Eloy, Arizona.

```{r}
```


# Question 7

**1 points**

Calculate the **mean**, **standard deviation**, and
**number of observations** for safflower yield.

```{r}
```


# Question 8

**2 points**

Calculate the **standard error of the mean** for
safflower yield.

```{r}
```


# Question 9

**2 points**

Calculate the *t*-value (using the `qt()` function)
needed to construct a **90% confidence interval**
for the mean safflower yield.

A 90% confidence interval leaves 5% of the
probability in each tail, so use `0.95`.

```{r}
```


# Question 10

**2 points**

Calculate the **90% confidence interval** for
safflower yield using the statistics computed above.

```{r}
```

`