---
title: "Introduction to R programming"
subtitle: "AGRON 5130"
author: "YOUR NAME HERE"
output: pdf_document
---

## Introduction

My main objectives with this first assignment are to get to know you a
little better, which will help me shape the class to some extent, and
verify that your R and R Studio installations are working properly.
Submit a knitted PDF generated from an R Markdown file to Canvas.

## Question 1

Write a short paragraph introducing yourself. Include:

-   Your background (education or work)
-   Why you are taking this course
-   One way you expect to use data or statistics in your professional
    work

You can replace this text with your answer

## Question 2

Write valid R commands that will conduct the following mathematical
operations:

a.  $18 + 58$

```{r}

```

b.  $3 \times 9$

```{r}

```

c.  $\sqrt{16}$

```{r}

```

## Question 3

Create a vector named "my_vector" and store the values 2, 4, 6, and 8 in
this vector.

```{r}

```

## Question 4

Now, let's find out how many elements this vector has. Use the function
"length" to count the number of values in the object "my_vector".

```{r}

```

## Question 5

Multiply all elements in the object "my_vector" by 2

```{r}

```

## Question 6

Multiply the first element in the object "my_vector" by 1, the second
element by 2, the third element by 3, and the fourth element by 4.
\small Hint: You may want to create another vector.

```{r}

```

## Question 7

In your own words, explain at a high level what the following piece of
code does. You do not need to explain each line.

```{r, eval=FALSE}
for (i in 1:5) {
  
print(i)
  
}
```

You can replace this text with your answer