---
title: "Exercise - Scatterplots and Correlation"
output: html_document
---

# Introduction

In this exercise, we stay with the first half of
Module 10. The workflow is the same one from
lecture: plot first, quantify second, interpret
last.

In this one, you will practice:

- visualizing quantitative relationships with scatterplots
- quantifying linear association with correlation
- separating "strong pattern" from "strong linear pattern"

We will use the corn allometry data and one short example where the
relationship is curved.

# Case Study: Corn Allometry

Allometry studies how plant structures scale with
one another. Think of this as a field team trying
to figure out which plant traits track with grain
yield well enough to justify deeper follow-up
work.

```{r}
library(ggplot2)

allometry <- read.csv("data/corn_allometry.csv")
head(allometry)
```

## Step 1: Start with a scatterplot matrix

Before computing anything, just look.

```{r}
plot(allometry)
```

This matrix is crowded, but it gives you a fast
first pass across all variable pairs. At this
stage, you are not trying to be precise. You are
trying to spot obvious direction, spread, and any
possible curve.

## Step 2: Focus on one biologically meaningful pair

Now isolate one pair with clear biological
meaning: yield and total leaf biomass.

```{r}
ggplot(allometry,
       aes(x = total_leaves, y = yield)) +
  geom_point(pch = 21, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  labs(x = "Total leaves (dry matter)",
       y = "Yield") +
  theme_bw()
```

The points suggest a positive linear association.
Pause and ask yourself whether the relationship
looks mostly linear, or if there is any obvious
curvature hidden in the cloud.

## Step 3: Quantify linear association

We can quantify the linear association between two
variables using the `cor()` function

```{r}
cor(allometry$yield, allometry$total_leaves)
```


We can also test whether we're likely to observe
this correlation value even if the true
correlation is actually 0.

```{r}
cor.test(allometry$yield, allometry$total_leaves)
```



## Step 4: Build a correlation matrix

```{r}
round(cor(allometry), 2)
```

Use the yield row as your first screening pass for
potentially useful predictors. This is not a final
model decision.

## Step 5: Important limitation of correlation

Correlation measures linear association. A strong
curved relationship can still produce low $r$.

Let's simulate a dataset that is strongly
associated (one of them is a function of the
other) but in a nonlinear way.

```{r}
set.seed(7)
x_nl <- seq(-3, 3, length.out = 50)
y_nl <- -(x_nl ^ 2) + rnorm(50, 0, 0.4)

nonlinear <- data.frame(x = x_nl, y = y_nl)
```

Then we can compute the correlation between them

```{r}
cor(nonlinear$x, nonlinear$y)
```



```{r}
ggplot(nonlinear, aes(x, y)) +
  geom_point(pch = 21, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  labs(title = paste0("Curved relationship, r = ",
                      round(cor(nonlinear$x, nonlinear$y), 2)),
       x = "x", y = "y") +
  theme_bw()
```

This is one reason we always plot before
interpreting $r$. If you only look at the
coefficient, you can miss a real relationship that
is simply not linear.

## If you are curious: what is $r$ actually doing?

Since I know you're curious, let's do this quick
"by hand" check

```{r}
x <- allometry$total_leaves
y <- allometry$yield

x_c <- x - mean(x)
y_c <- y - mean(y)

num <- sum(x_c * y_c)
den <- sqrt(sum(x_c ^ 2) * sum(y_c ^ 2))

r_manual <- num / den
r_builtin <- cor(x, y)

r_manual
r_builtin
```


# Practice 1

Using `allometry`, create a scatterplot with `height` on the x-axis and
`yield` on the y-axis.

After plotting, add one sentence: does a linear trend look believable
for this pair, yes or no?

```{r}
# your code here
```

# Practice 2

Using `allometry`, run `cor.test()` for `height` and `yield`.

Then write one sentence interpreting:

- the sign of $r$
- the magnitude of $r$
- whether this proves causation

Keep your sentence in agronomic language, not only statistical terms.

```{r}
# your code here
```

# Practice 3

Load the fertility dataset and inspect it.

```{r}
fertility <- read.csv("data/corn_fertility.csv")
head(fertility)
```

Create a scatterplot matrix for all quantitative variables.

As you inspect it, note one variable pair that looks roughly linear and
one that looks weak or noisy.

```{r}
# your code here
```

# Practice 4

Create a correlation matrix for all variables in `fertility`.

Then identify:

- one variable with relatively strong positive correlation with yield
- one variable with weak correlation with yield

Add one sentence on why correlation alone is not enough to claim a
cause-and-effect relationship for either variable.

```{r}
# your code here
```
