---
title: "Exercise - Simple Linear Regression"
output: html_document
---

# Introduction

In the previous exercise, we focused on scatterplots and correlation.
In this one, we move to the second half of Module 10: fitting and interpreting
simple linear regression models.

Keep one bridge in mind from earlier modules. This is the same playbook,
just with a continuous predictor:

- ANOVA with categorical predictors: $Y_{ij} = \mu + \tau_i + \epsilon_{ij}$
- Regression with a continuous predictor: $Y_i = \beta_0 + \beta_1 x_i + \epsilon_i$

The assumptions and diagnostics follow the same workflow you already
know.

# Case Study: Corn Allometry

```{r}
library(ggplot2)

allometry <- read.csv("data/corn_allometry.csv")
head(allometry)
```

## Step 1: Fit the model

Model yield as a function of total leaf biomass.

Start simple, then build your interpretation around that first model.

```{r}
model_yield_leaves <- lm(yield ~ total_leaves, data = allometry)
summary(model_yield_leaves)
```

## Step 2: Plot the relationship

```{r}
ggplot(allometry, aes(x = total_leaves, y = yield)) +
	geom_point(pch = 21, size = 2) +
	geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
	labs(x = "Total leaves (dry matter)",
			 y = "Yield") +
	theme_bw()
```

## Step 3: Interpret coefficients

```{r}
coef(model_yield_leaves)
```

Write the fitted equation:

$$
\hat{y} = \hat{\beta}_0 + \hat{\beta}_1x
$$

and interpret:

- intercept (expected yield at $x = 0$, if meaningful in range)
- slope (expected change in yield per one-unit increase in leaves)


## Step 4: Test significance two equivalent ways

```{r}
summary(model_yield_leaves)
anova(model_yield_leaves)
```

For simple regression, testing whether the slope is zero is equivalent
to the model F-test. You are seeing the same hypothesis presented in two
output formats.

## Step 5: Check model fit and assumptions

```{r}
summary(model_yield_leaves)$r.squared

plot(model_yield_leaves, which = 1)
plot(model_yield_leaves, which = 2)
```

Pause and think about:

- any strong curvature in residuals vs fitted
- any funnel pattern
- whether Q-Q points stay reasonably close to a line

If one diagnostic looks poor, do not ignore it just because the slope is
significant.

# Practice 1

Now model `yield` as a function of `total_leaf_area`.

1. Fit the model.
1. Plot the fitted line over the scatterplot.
1. Use `coef()` and `summary()` to inspect slope, p-value, and $R^2$.
1. Plot residual diagnostics (`which = 1` and `which = 2`).

```{r}
# your code here
```

Quick checkpoint:

- Is the slope positive or negative?
- Is the relationship statistically significant?
- Do diagnostics support a linear mean trend?

Write one additional sentence on model adequacy: would you trust this
model for interpretation only, for prediction, or for neither?

# Practice 2

Now use the fertility dataset and model `yield` as a function of
`organic_matter`.

```{r}
fertility <- read.csv("data/corn_fertility.csv")
head(fertility)
```

1. Fit `lm(yield ~ organic_matter, data = fertility)`.
1. Summarize coefficients and write the fitted equation.
1. Report $R^2$.
1. Create residual and Q-Q diagnostics.

```{r}
# your code here
```

