---
title: "Exercise - Predictions from Regression Models"
output: html_document
---

# Introduction

In this exercise, we focus on one of the main goals of regression:
prediction.

In this one, focus on three ideas from Module 10:

- building the prediction equation from `lm()`
- making predictions for new x-values
- distinguishing interpolation from extrapolation

As you work, keep one rule in mind: a prediction is only as good as the
model and data range behind it.

# Case Study: Corn Allometry

```{r}
library(ggplot2)

allometry <- read.csv("data/corn_allometry.csv")
head(allometry)
```

## Step 1: Fit the model

We model yield as a function of stalk circumference.

This gives us the equation we will use for the rest of the exercise.

```{r}
model_circ <- lm(yield ~ stalk_circ, data = allometry)
coef(model_circ)
summary(model_circ)$r.squared
```


## Step 2: Visualize observed points and fitted trend

```{r}
ggplot(allometry, aes(x = stalk_circ, y = yield)) +
  geom_point(pch = 21, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  labs(x = "Stalk circumference", y = "Yield") +
  theme_bw()
```

The grey ribbon is the confidence band for the mean trend.
Notice that uncertainty is usually smallest near the center of observed
$x$ values and larger at the edges.

## Step 3: Predict at new values of x

```{r}
new_stalk <- data.frame(
  stalk_circ = c(1.5, 2.0, 2.5, 3.0, 3.5)
)

new_stalk$fit <- predict(model_circ, newdata = new_stalk)
new_stalk
```

## Step 4: Confidence interval vs prediction interval

For each x-value, we can estimate:

- confidence interval for the mean response
- prediction interval for a new individual observation

```{r}
pred_ci <- predict(model_circ, newdata = new_stalk, interval = "confidence")
pred_ci

pred_pi <- predict(model_circ, newdata = new_stalk, interval = "prediction")
pred_pi
```

The table shows the numbers, but the difference
becomes clearer on a plot. Below, we build a dense
prediction grid across the observed range, compute
both intervals, and overlay them on the original
data.

```{r}

x_grid <- data.frame(
  stalk_circ = seq(min(allometry$stalk_circ), max(allometry$stalk_circ), length.out = 100)
)

ci_band <- as.data.frame(predict(model_circ, newdata = x_grid, interval = "confidence"))
pi_band <- as.data.frame(predict(model_circ, newdata = x_grid, interval = "prediction"))

x_grid$fit <- ci_band$fit
x_grid$ci_lwr <- ci_band$lwr
x_grid$ci_upr <- ci_band$upr
x_grid$pi_lwr <- pi_band$lwr
x_grid$pi_upr <- pi_band$upr

ggplot(allometry, aes(x = stalk_circ, y = yield)) +
  geom_ribbon(data = x_grid, aes(y = fit, ymin = pi_lwr, ymax = pi_upr),
              fill = "orange", alpha = 0.25) +
  geom_ribbon(data = x_grid, aes(y = fit, ymin = ci_lwr, ymax = ci_upr),
              fill = "steelblue", alpha = 0.4) +
  geom_line(data = x_grid, aes(y = fit), color = "steelblue", linewidth = 0.8) +
  geom_point(pch = 21, size = 2) +
  labs(x = "Stalk circumference", y = "Yield",
       caption = "Blue: 95% confidence interval for the mean | Orange: 95% prediction interval for a new observation") +
  theme_bw()
```

The blue band captures uncertainty about the mean trend.
The orange band is wider because it also accounts for individual-level
scatter around that mean — it tells you where a single new plant is
likely to fall.

## Step 5: Interpolation vs extrapolation

```{r}
range(allometry$stalk_circ)
```

Now include one value clearly outside the observed range.

```{r}
new_stalk_ext <- data.frame(stalk_circ = c(1.8, 2.4, 3.0, 4.2))
new_stalk_ext$fit <- predict(model_circ, newdata = new_stalk_ext)
new_stalk_ext
```

Identify which of these values are interpolation and which are
extrapolation.


## If you are curious: significant slope does not always mean the line is right

Here is a quick reminder from lecture: a
significant linear slope does not guarantee that a
straight-line model is adequate.

```{r}
set.seed(13)
x_demo <- rep(seq(0, 120, length.out = 9), each = 4)
y_demo <- 8 + 0.06 * x_demo - 0.00025 * (x_demo ^ 2) + rnorm(length(x_demo), 0, 0.5)

demo_curve <- data.frame(x = x_demo, y = y_demo)

ggplot(demo_curve, aes(x = x, y = y)) +
  geom_point(pch = 21, size = 2) +
  theme_bw()
```

Let's fit a linear model to this data and look at
the summary

```{r}
demo_lm <- lm(y ~ x, data = demo_curve)
summary(demo_lm)
```

The summary looks great overall, but what do the
residuals tell us?

```{r}
plot(demo_lm, which = 1)
plot(demo_lm, which = 2)
```

Think about this: if p-value is small but
residuals are curved, which signal do you trust
for model adequacy, the p-value or the residual
pattern?

# Practice 1

Model `yield` as a function of `area_highest_leaf`.

1. Fit the model and inspect coefficients with `coef()` and $R^2$ with `summary()`.
1. Plot points plus fitted line and confidence band.
1. Predict for `area_highest_leaf = 900, 1000, 1100, 1200, 1300`.
1. Add confidence and prediction intervals for those values.

```{r}
# your code here
```


# Practice 2

Now use the fertility dataset and model `yield` as a function of
`organic_matter`.

```{r}
fertility <- read.csv("data/corn_fertility.csv")
head(fertility)
```

1. Fit `lm(yield ~ organic_matter, data = fertility)`.
1. Plot points and fitted line.
1. Predict yield for `organic_matter = 1.5, 1.8, 2.1, 2.4, 2.7`.
1. Add confidence and prediction intervals.
1. Based on the observed range, label each new value as interpolation or
   extrapolation.

```{r}
# your code here
```

Final checkpoint:

If the model has low $R^2$ but a significant slope, should we still use
it for prediction? Answer with one practical rule for agronomic use,
including how you would check whether predictions are still useful.