---
title: "Spatial Statistics"
subtitle: "AGRON 5130 - Module 12"
author: "Caio dos Santos"
date: "2026-03-11"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    smooth: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Introduction

Spatial data are central to agronomy: yield maps,
soil maps, management zones, and prescription
rasters.

In this unit, we focus on three foundations:

- projection and coordinate systems
- shapefiles and geometry operations
- rasters, interpolation, and map creation

# Projection (General)

Projection is how we represent points from a
curved earth on a flat map.

Different projections preserve different
properties (area, distance, direction), and no
projection is perfect for all purposes.

## Why Projection Matters

If layers use different coordinate systems, they
may not align correctly. This can break overlays,
joins, area calculations, and distance analyses.

## Common CRS Workflows in R

```{r}
library(sf)

# Example: read and inspect CRS
# field <- st_read("data/field_boundary.shp")
# st_crs(field)
```

Typical steps:

1. inspect CRS with `st_crs()`
1. transform layers with `st_transform()`
1. verify alignment before analysis

# Shape Files

A shapefile stores geometry plus attributes.
Common geometry types:

- POINT
- LINESTRING
- POLYGON

## Core Operations

```{r}
# Example geometry operations
# st_intersection(a, b)
# st_difference(a, b)
# st_union(a, b)
```

These operations allow us to combine and compare
spatial layers for practical agronomic questions.

## Case Study Concept

A common use case is linking sampled points (yield
or soil tests) with polygon-based information
(e.g., SSURGO map units).

# SSURGO Integration

SSURGO can be pulled into R and joined to field
boundaries to enrich local analyses with:

- texture class
- organic matter estimates
- drainage indicators
- available water capacity

The key principle is that map-unit level
attributes are modeled estimates and should be
validated against local measurements when possible.

# Rasters

Rasters organize space into equal-size cells.
Each cell stores one value, such as:

- elevation
- pH
- phosphorus
- recommendation rate

## Interpolation

From point samples, we can predict values for
unsampled cells.

Common approach used here: kriging.

Workflow:

1. prepare points and boundary
1. build interpolation grid
1. fit semivariogram
1. run kriging
1. map interpolated surface

## Semivariogram Concept

The semivariogram models how similarity decreases
with distance.

If nearby points are similar, kriging can use that
spatial dependence to produce stronger predictions.

# Practical Mapping

For decision support, maps should be:

- interpretable
- transparent about units and scale
- accompanied by legends and context

For recommendations (e.g., lime), map outputs
should clearly show both predicted values and
management classes.

# Minimal Spatial Template

```{r}
# Pseudocode template
#
# points <- st_read("data/samples.shp")
# boundary <- st_read("data/boundary.shp")
#
# points <- st_transform(points, 3857)
# boundary <- st_transform(boundary, 3857)
#
# grid <- st_bbox(boundary) |> st_as_stars() |> st_crop(boundary)
# v <- gstat::variogram(measure ~ 1, points)
# m <- gstat::fit.variogram(v, gstat::vgm("Sph"))
# kr <- gstat::krige(measure ~ 1, points, grid, model = m)
```

# Conclusion

Spatial workflows in R let us move from raw field
points to interpretable decision layers.

Core takeaways:

- always verify CRS and projection alignment
- understand your geometry type before analysis
- use interpolation carefully and validate where
  possible
- communicate results with clear map design and
  units

These skills are directly transferable to precision
ag and on-farm analytics.
