---
title: "Module 9 Assignment: Data Quality and Model Diagnostics"
subtitle: "AGRON 5130"
author: "Caio dos Santos"
output: pdf_document
---

# Introduction

This assignment guides you through a complete statistical workflow:
explore, fit, check assumptions, decide, and predict. You will apply
what you learned in the lecture and exercises about inspecting data
for assumption violations and deciding whether a model is appropriate
before trusting its conclusions.

Our case study is a wheat yield trial from India where the primary
research question is: how does weed infestation reduce yield? You'll
use simple linear regression—but with a critical eye on whether the
model assumptions hold. Before making predictions, you must verify
that the model respects its assumptions. If it does not, you should
discuss what the violations mean for your conclusions.

# Question 1

**6 points**

Load the wheat-weed dataset and examine its structure. Before fitting
any model, explore the relationships visually and numerically. This
is how we spot problems before they hide in residuals.

```{r}
library(ggplot2)

weed <- read.csv("data/wheat_weed_india.csv")
head(weed)
```

1. Create a scatterplot matrix for all numeric variables in the
dataset. Which pairs of variables appear most strongly related?

```{r}

```

2. Compute the correlation coefficient between `total_weed_dm` (weed
dry matter biomass) and `yield`. Use `cor.test()` to test whether
this correlation is statistically significant.

```{r}

```

3. Create a scatterplot with `total_weed_dm` on the x-axis and `yield`
on the y-axis. Add a title and axis labels. Describe what you see
in one or two sentences. Does the relationship look linear?

```{r}

```

4. What is the correlation coefficient? How would you describe the
strength of the relationship: weak, moderate, or strong?

# Question 2

**4 points**

Now fit a simple linear regression model to answer the question:
how much does yield decline as weed biomass increases?

1. Fit a linear model: `yield ~ total_weed_dm`. Display the model
coefficients and their significance using `summary()` or similar.

```{r}

```

2. Interpret the slope coefficient. What is the numeric value? Is it
statistically significant at $\alpha = 0.05$? In plain language, what
does the slope mean for a farmer? (Hint: "For every unit increase in
weed biomass, yield changes by...")

# Question 3

**8 points**

The linear regression model makes three key assumptions about
residuals: normality, homogeneity of variance, and independence. 
In this question, you'll check whether these assumptions hold.

1. Extract the residuals from your model and create a histogram. Does
the distribution appear approximately normal? (Rough symmetry around
zero is what you're looking for.) Briefly explain whether normality
seems reasonable.

```{r}

```

2. Create a residuals vs. fitted values plot. (Hint: `plot(model)` 
will produce diagnostic plots including this one, or you can make it 
with ggplot using `fitted(model)` and `residuals(model)` as 
coordinates.) Examine the plot carefully and answer: Do residuals 
scatter randomly around a horizontal line at zero? Do you see any 
pattern (curve, funnel shape, or clustering)? What would a 
problematic pattern suggest about the model?

```{r}

```

3. Create a Q-Q plot to check normality more formally. (Hint: 
`plot(model)` produces this automatically as plot 2, or use 
`ggplot(aes(sample = residuals)) + stat_qq() + stat_qq_line()`.) 
Do the points fall approximately along the diagonal line? Describe 
any systematic departures. Are they severe enough to undermine the 
model, or minor?

```{r}

```

# Question 4

**5 points**

Based on your assumption checks in Question 3, assess whether the 
linear model is trustworthy.

1. Summarize what you found in your diagnostic plots. Do residuals 
appear normally distributed? Is variance homogeneous across fitted 
values? Are there influential outliers?

```{r}

```

2. Given your findings, do you think a simple linear regression is 
appropriate here? Why or why not? What assumption(s) might be 
questionable?

# Question 5

**4 points**

Now that you've evaluated the model's assumptions, use it to make 
predictions—but with awareness of any limitations you discovered.

1. Using your model, predict yield for the following weed biomass 
levels. Display the predictions along with 95% confidence intervals.

```{r}
new_data <- data.frame(
  total_weed_dm = c(10, 15, 20, 25)
)

```

2. Interpret your predictions. What yield does your model predict 
when weed biomass is 10 kg/ha? What trend does yield follow as weed 
biomass increases? Given your assumption checks from Question 3, how 
much confidence do you have in these predictions? Explain.

# Question 6

**3 points**

Reflect on the complete workflow you just completed. You explored the
data and found a correlation, fit a model and estimated the effect 
size, checked model assumptions, and made predictions.

Write a short paragraph (~150 words) summarizing:

- What question did we answer?
- What did the model tell us about the effect of weeds on yield?
- What limitations did the assumption checks reveal?
- If your model violated key assumptions, what might you do next?
  (Hint: Refer back to the lecture on transformations or alternative
  approaches.)

```{r}

```
