---
title: "Module 9 Assignment"
subtitle: "AGRON 5130"
author: "Caio dos Santos"
output: pdf_document
---

# Introduction

This assignment practices core skills from Module 9:
visualizing relationships, computing correlation,
fitting simple linear regression models, checking
model diagnostics, and generating predictions.

```{r}
library(tidyverse)
library(broom)

weed <- read.csv("data/wheat_weed_india.csv")
head(weed)
```

# Question 1: Correlation and Visualization (6 points)

## Question 1a (1 point)
Create a scatterplot matrix for the `weed` dataset.

```{r}

```

## Question 1b (1 point)
Create a correlation matrix for all numeric variables
in the dataset.

```{r}

```

## Question 1c (1 point)
Create a scatterplot for `total_weed_dm` (x-axis)
and `yield` (y-axis).

```{r}

```

## Question 1d (1 point)
Test the correlation between `total_weed_dm`
and `yield` using `cor.test()`.

```{r}

```

## Question 1e (2 points)
Interpret the relationship between `total_weed_dm`
and `yield`.

- Is it strong or weak?
- Is it positive or negative?
- Explain it in plain language.

# Question 2: Regression and Prediction (14 points)

## Question 2a (2 points)
Create a linear model of `yield` as a function of
`total_weed_pop`. Display coefficient results using
`tidy()`.

```{r}

```

## Question 2b (1 point)
Display model fit statistics using `glance()`.

```{r}

```

## Question 2c (2 points)
Interpret the model.

- Is the slope positive or negative?
- Is the slope significant?
- Based on `r.squared`, what percentage of yield
  variation is explained?

## Question 2d (1 point)
Plot model residuals.

```{r}

```

## Question 2e (1 point)
Based on the residual plot, does a linear model
seem appropriate? Explain briefly.

## Question 2f (2 points)
Use `ggplot()` and `geom_smooth(method = "lm")`
to plot the regression line for `yield` vs
`total_weed_pop`.

```{r}

```

## Question 2g (2 points)
Using the data frame below, predict `yield`
for each weed population and display results.

```{r}
new_weed_pops <- data.frame(
  total_weed_pop = c(-1, 1, 3, 5, 7)
)

```

## Question 2h (3 points)
Interpret your predictions.

- What trend do you see as weed population increases?
- Which prediction is biologically unrealistic,
  and why?
- What does this suggest about prediction limits
  and extrapolation?
