---
title: "Analyzing Non-Normal Data"
subtitle: "AGRON 5130"
author: "YOUR NAME HERE"
output: pdf_document
---

# Introduction

The goal of this assignment is to give you hands-on
practice with **diagnosing model assumptions** and
working with **non-normal data**.

You will work with a grape yield dataset from a
vineyard experiment evaluating canopy management
strategies.

In this assignment, you will practice:

- visualizing skewed data
- diagnosing violations of model assumptions
- applying a **log transformation**
- fitting a **GLM**
- interpreting treatment effects across models

Submit a **knitted PDF/HTML generated from your R Markdown
file** to Canvas.



## Background

Grapevine productivity is strongly influenced by
canopy structure, which affects light
interception, air movement, and ultimately fruit
development.

In commercial vineyards, growers often use
different **canopy management practices** to
balance vegetative growth and fruit production.
However, these practices can have variable effects
depending on environmental conditions and
vine-to-vine variability.

A field trial was conducted to evaluate the effect
of different canopy management strategies on
grapevine productivity in a mature vineyard.

The treatments included:

- **Control** (no canopy management)  
- **Shoot thinning** (removal of excess shoots early in the season)  
- **Leaf removal** (removal of basal leaves around the fruiting zone)  
- **Shoot thinning + leaf removal**  

The experiment was conducted using a **completely
randomized design** with 30 replications.

At the end of the growing season, the number of
**grape clusters per vine** was recorded for each
experimental unit.

Due to natural variability among vines and the
biological processes governing fruit set, the
response variable is expected to show:

- substantial variability across treatments  
- increasing variance with increasing mean  
- right-skewed distributions  

These characteristics make this dataset a good
candidate for exploring **violations of model
assumptions** and methods to address them.

We can start by loading the necessary libraries

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(emmeans)
```


# Question 1

**1 point**

Load the dataset `"data/grape.csv"` and display
the first few rows.

**note:**  make sure to call the dataset `grape`.
This is important for the models I have pre built
for you after question 10.

```{r}
```


# Question 2

**2 points**

Create a plot showing the relationship between
`treatment` and `clusters`.

```{r}
```


# Question 3

**1 point**

Create a histogram of `clusters`.

Briefly describe the distribution (normal vs skewed).

```{r}
```


# Question 4

**2 points**

Fit a linear model and extract the treatment
means. Then, comment on the confidence intervals.
Do any of the confidence intervals include negative values?
Why might this be a problem for this type of data?

```{r}
```


# Question 5

**1 points**

Create diagnostic plots and comment on:

* normality
* constant variance

```{r  fig.show="hold", out.width="50%"}
```


# Question 6

**1 points**

Fit a model using a **log transformation** of the response.

```{r}
```


# Question 7

**1 point**

Create diagnostic plots for the log-transformed model.

Did things improve?

```{r  fig.show="hold", out.width="50%"}
```


# Question 8

**1 point**

Compute estimated marginal means for the log model.

```{r}
```


# Question 9

**1 point**

Back-transform the means using `exp()`.

```{r}
```


# Question 10

**1 point**

Compute the difference between the treatments
`Both` and `Control` presented on the log scale
then interpret this difference in the data scale

```{r}
```



Now we will fit a **GLM appropriate for this type
of data** using a **Gamma distribution with a log
link**. We will use the model `mod_glm` for the
next questions.

```{r}
mod_glm <- glm(clusters ~ treatment,
               family = Gamma(link = "log"),
               data = grape)
```



# Question 11

**1 point**


Create diagnostic plots for the GLM model. Is this
model a good fit? Are we concerned about violating
assumptions?

```{r  fig.show="hold", out.width="50%"}
```


# Question 12

**1 point**

Use `emmeans()` to extract treatment means on the
**response scale**. Then, comment on the confidence intervals.
Do any of the confidence intervals include negative values?

```{r}
```


# Question 13

**1 point**

Briefly compare the GLM results to the
log-transformed model (are the treatment effects
similar?). Discuss how these two approaches are an
improvement uppon the first approach in which we
ignore the non-normality of the residuals.
