---
title: ANOVA Exercise
output: html_document
editor_options: 
  markdown: 
    wrap: 100
---

# Introduction

In this unit we were introduced to the Analysis of
Variance (ANOVA) and the most basic experimental
design, the Completely Randomized Design. Running
an analysis of variance for this trial is very
easy, as we will now see.

# Case Study: Barley

```{r}
barley <- read.csv("data/barley_crd.csv")
head(barley)
summary(barley)
```


This dataset contains:

- 5 genotypes
- 10 replications per genotype
- 50 total experimental units


# ANOVA

Before running any statistical test, we should
always look at the data.

This gives us intuition about:

- The magnitude of treatment differences
- The variability within treatments
- Whether an ANOVA result will make sense

```{r}
library(ggplot2)
ggplot(barley, aes(x = gen, y = yield)) +
  geom_point(pch = 21, size = 2) +
  stat_summary(fun = mean, geom = "point", shape = 17, size = 3) +
  stat_summary(fun.data = 'mean_se', geom = "errorbar", width = 0.2) +
  theme_linedraw() +
  coord_flip()+
  labs(x = "Genotype",
       y = "Yield (bu/ac)")
```

# Running the ANOVA in R

Conceptually, we are modeling:

> Yield as a function of genotype.

In R, that is written as:

yield ~ gen

Let’s fit the model.

```{r}
barley_aov <-  aov(yield ~ gen, data = barley)
barley_aov
```

Now, to obtain the ANOVA table:

```{r}
summary(barley_aov)
```

And that’s it. But, again, the goal here is not
just to press buttons.

We want to understand what this table represents.

So let’s build it ourselves.



## Linear additive model

Remember our model:


$$Y_{ij} = \mu + \tau_i + \epsilon_{ij}$$

$Y_{ij}$ = observed yield 

$\mu$ = overall mean yield

$\tau_i$ = treatment effect 

$\varepsilon_{ij}$ = random error


## Step 1: Overall Mean ($\mu$)

```{r, echo=TRUE}
mu <- mean(barley$yield)
mu
```

This represents the grand mean across all 50
observations.


## Step 2: Treatment Effects ($\tau_i$)

Now we compute genotype means and subtract the
overall mean.

```{r, echo=TRUE}
taus <- aggregate(barley$yield,
                  barley['gen'],
                  mean)
names(taus) <- c('gen', 'gen_means')
taus$mu <- mu
taus$tau <- taus$gen_means - taus$mu
taus
```

Each $\tau_i$ represents how far a genotype mean
deviates from the overall mean.

Some are positive.
Some are negative.

Importantly:

The sum of all τᵢ values (weighted properly)
equals zero

## Step 3: Error Terms ($\epsilon_{ij}$)

Now we compute the residuals, the deviation of
each observation from its predicted value.

```{r, echo=TRUE}
barley_eff <- merge(barley, taus)
barley_eff$epsilon <- with(barley_eff, yield - mu - tau)
head(barley_eff)
```

These $\epsilon$ terms represent the background noise in the experiment.



To make it easier for us to visualize these effects, 
let's plot them:

```{r}
ggplot(data = barley_eff)+
  geom_vline(aes(xintercept = mu, col = 'mu'), linetype = 1)+
  geom_segment(aes(x = mu, xend = gen_means, y = rep, yend = rep, col = 'tau'), linetype = 1)+
  geom_segment(aes(x = gen_means, xend = yield, y = rep, yend = rep, col = 'epsilon'), linetype = 1)+
  geom_point(aes(x = yield, y = rep))+
  labs(x = 'Yield (bu/ac)', y = 'Replication', col = 'Additive Effect')+
  facet_wrap(~gen, ncol = 1)+
  theme_gray()+
  scale_y_continuous(breaks = 1:8)+
  theme(panel.border = element_rect(colour = 'black', fill = 'transparent'),
        strip.background = element_rect(colour = 'black', fill = 'grey'),
        panel.spacing = unit(0, 'cm'))
```

## Setting up the ANOVA Table

Great! Now we have all the effects we need to
compute the $F$-statistic!

I will set up a table for us to keep track of
these numbers and we will populate this table as
we go:

| Source of Variation | SS  | DF  | MSE | F   |
|:--------------------|:----|:----|:----|:----|
| Treatment           |     |     |     |     |
| Residuals (Error)   |     |     |     |     |
| Total               |     |     |     |     |

## Sums of Squares

Let's compute the Sums of Squares for all three rows

```{r, echo=TRUE}
SS_treatment <- sum(barley_eff$tau ^ 2) ## Treatment
SS_error <- sum(barley_eff$epsilon ^ 2) ## Error
SS_total <- sum((barley_eff$yield - barley_eff$mu)^2) ## Total

SS_treatment
SS_error
SS_total
```

Now, let's add these values to our table:

| Source of Variation | SS   | DF  | MSE | F   |
|:--------------------|:-----|:----|:----|:----|
| Treatment           | 3632 |     |     |     |
| Residuals (Error)   | 2313 |     |     |     |
| Total               | 5945 |     |     |     |

Great! Let's continue populating our table. Next,
let's compute the degrees of freedom!

## Degrees of Freedom

We can compute the total number of degrees of freedom as:

$DF_{total}$ = 50 - 1 = 49

The degrees of freedom for the treatment can be
computed as $K - 1$, where K is the number of
means we are estimating.

$DF_{treatment}$ = 5 - 1 = 4

The degrees of freedom for the error, is simply
the difference between the total degrees of
freedom, and the degrees of freedom taken by the
treatment:

$DF_{error}$ = 49 - 4 = 45

Our table now becomes:

| Source of Variation | SS   | DF  | MSE | F   |
|:--------------------|:-----|:----|:----|:----|
| Treatment           | 3632 | 4   |     |     |
| Residuals (Error)   | 2313 | 45  |     |     |
| Total               | 5945 | 49  |     |     |

## Mean Square Error

Next, we need to take the ratio between the SS and
DF columns. Remember, this will give us an
estimate for the variance of each effect. This
estimate is called Mean Square Error (MSE).

```{r, echo=TRUE}
MS_treatment <- SS_treatment/4 # MSE Treatment
MS_error <- SS_error/45 # MSE Residuals

MS_treatment
MS_error
```

| Source of Variation | SS   | DF  | MSE | F   |
|:--------------------|:-----|:----|:----|:----|
| Treatment           | 3632 | 4   | 908 |     |
| Residuals (Error)   | 2313 | 45  | 51  |     |
| Total               | 5945 | 49  |     |     |

## Computing the $F$-statistic

Now, since the $F$-statistic is computed as the
ratio between the treatment and error variances,
and we are estimating them using the MSE, $F$ can
be computed as follows:

$$F = \frac{\sigma^2_{treatment}}{\sigma^2_{error}} = \frac{MSE_{treatment}}{MSE_{error}}$$


If genotypes truly do not differ, both MS terms
estimate the same variance.

F should be close to 1.

If F is large, treatment variability is much
larger than background noise.

That is evidence against the null hypothesis

```{r, echo=TRUE}
MS_treatment / MS_error
```


Alright! This concludes our ANOVA table!!

| Source of Variation | SS   | DF  | MSE  | F     |
|:--------------------|:-----|:----|:-----|:------|
| Treatment           | 3632 | 4   | 1210 | 17.66 |
| Residuals (Error)   | 2313 | 45  | 82   |       |
| Total               | 5945 | 49  |      |       |

## P-value

Now, like we have done before, we have to compare
the $F$-statistic value we found with the
$F$-distribution and compute the probability of
observing a more extreme value given that $H_0$ is
true.

We will use the `fastGraph` library for that once
again:

```{r, echo=TRUE}
library(fastGraph)
shadeDist(xshade = 17.66,
          ddist = 'df',
          parm1 = 4,
          parm2 = 45,
          lower.tail = FALSE)

```


If this probability is small, we reject the null
hypothesis:

$$H0: \mu_0 = \mu_1 = \ldots = \mu_5$$

And conclude that at least one genotype differs.

# Practice: Beet Data

```{r}
beet_data <- read.csv("data/beets_crd.csv")
head(beet_data)
```

Compute the treatment means for the beet dataset

```{r}
beet_trt_means <- aggregate(beet_data['yield'],
                            beet_data['fert'],
                            mean)
beet_trt_means
```


Model yield as a function of fertilizer.

You should obtain:

Pr(>F) = 1.299e-10

```{r}
beet_model <-  aov(, data = )
summary(beet_model)
```


# Practice: Potato Data

Model infection as a function of treatment.

```{r}
potato_data <-  read.csv("data/potato_crd.csv")
head(potato_data)
```

Compute the treatment means for the beet dataset

```{r}
potato_trt_means <- aggregate(potato_data['inf'],
                            potato_data['trt'],
                            mean)
potato_trt_means
```


You should obtain:

Pr(>F) = 0.0103

```{r}

```

