---
title: "Probability and Distributions"
subtitle: "AGRON 5130"
author: "YOUR NAME HERE"
output: pdf_document
---

# Introduction

The goal of this assignment is to give you hands-on
practice working with probability, probability
distributions, and the normal distribution in R.
You will work with both simulated data and spatial
yield data to estimate probabilities, visualize
distributions, and connect summary statistics to
probabilistic statements. Submit a knitted
PDF/HTML generated from your R Markdown file to
Canvas.


# Question 1

**1 point**

Load the `peanut_uniformity` dataset and assign it
to an object called `peanut`.

*note: * this might be a good place to include
a code chunk loading the `fastGraph` library


```{r}
```

# Question 2

**1 point**

Inspect the dataset using the `head()` function.

```{r}
```

# Question 3

**1 point**

Create a histogram of yield using the `hist()`
function.

```{r}
```

# Question 4

**2 points**

Calculate the mean yield and assign it to the
variable `peanut_mu`. Calculate the standard
deviation of yield and assign it to the variable
`peanut_sigma`.

```{r}
```

# Question 5

**2 points**

Plot the normal distribution for the peanut yield
data and calculate the probability of observing a
yield less than 1.5.

```{r}
```

# Question 6

**1 point**

Calculate the probability of observing a yield
greater than 3.5.

```{r}
```

# Question 7

**1 point**

Calculate the probability of observing a yield
greater than or equal to 1.6 and less than 2.9.

```{r}
```

# Question 8

**1 point**

Load the shapefile
`"data/merriweather_yield_map/merriweather_yield_map.shp"`
using `st_read()` and assign it to an object called
`soybean`.

*note:* Hear that? It sounds like there's someone saying "Load the sf package!"

```{r}
```

# Question 9

**1 point**

Inspect the `soybean` data using the `head()`
function.

```{r}
```

# Question 10

**2 points**

Plot a yield map using the `"yield_bu"` column of
the `soybean` dataset and create a histogram of
yield.

```{r}
```

# Question 11

**2 points**

Calculate the mean yield and assign it to the
variable `soybean_mu`. Calculate the standard
deviation of yield and assign it to the variable
`soybean_sigma`.

```{r}
```

# Question 12

**1 point**

Using the normal distribution, calculate the
probability of observing a yield greater than or
equal to 90 bu/ac.

```{r}
```

# Question 13

**1 point**

Using the normal distribution, calculate the
probability of observing a yield between 75 and
85 bu/ac.

```{r}
```

# Question 14

**1 point**

If the probability of observing a yield less than
65.8 bu/ac is 5%, about how many observations out
of 100 would you expect to be less than or equal
to 65.8 bu/ac?

```{r}
```

Write your brief interpretation here:

    (Your answer)

