---
title: "Exercise: Shapefiles"
output: html_document
---


## Setup

If you haven't already, install the required packages:

```{r, eval=FALSE}
install.packages("sf")
install.packages("apsimx")
install.packages("gstat")
install.packages("stars")
install.packages("soilDB")
install.packages("spData")
```

These are only needed once. After installation, you can comment out or delete this chunk.

# Introduction
If you download data from a tractor,
applicator, or combine, it will often be in
shapefile format.

A shapefile is usually bundled as multiple files
with the same name and different extensions (for
example `.dbf`, `.prj`, `.shp`, and `.shx`). The
`.prj` file is especially important because it
stores projection/CRS information.

# Case Study
We will work with another soil test shapefile from Minnesota.

```{r}
library(sf)

soil_test_data <- st_read("data/Burnholdt_grid_sample.shp", quiet = TRUE)
head(soil_test_data)
```

Always inspect geometry before analysis.

Some layers use `POINT` geometry (single location
per record), while others use `POLYGON` geometry.
That determines what kind of plot and operations
make sense.

Confirm geometry and CRS:

```{r}
unique(st_geometry_type(soil_test_data))
st_crs(soil_test_data)
```

Now make CRS actionable. If data are in
longitude/latitude, transform them to a projected
CRS before analysis.

```{r}
st_is_longlat(soil_test_data)

# Use a projected CRS for local field analysis.
soil_test_data <- st_transform(soil_test_data, 32615)
st_crs(soil_test_data)
```

In this exercise, we will map soil organic matter
(`attribute == "Om"`).

```{r}
unique(soil_test_data$attribute)
single_attribute_data <- subset(soil_test_data, attribute == 'Om')
```

Check a few rows:

```{r}
head(single_attribute_data)
```

Now create a base map using `plot()`.

```{r}
plot(st_geometry(single_attribute_data),
     pch = 16, cex = 1,
     col = "grey40", main = "Soil OM Sampling Points")
```

To color points by OM values, plot the `measure`
column directly. `sf` handles the color scale and
legend automatically.

```{r}
plot(single_attribute_data["measure"], 
     pch = 16, cex = 0.8,
     main = "Soil Organic Matter (OM)")
```



# Practice
Using the same workflow, map another soil
attribute.

Run this to see available attributes:

```{r}
unique(soil_test_data$attribute)
```

Choose one attribute (for example `P`, `K`, or
`Ph`) and complete the steps below:

1) Filter rows for your selected attribute.
2) Plot the `measure` column directly using `sf`'s
   default method.

```{r}
practice_data <- soil_test_data[soil_test_data$attribute == "Ph", ]

plot(practice_data["measure"], pch = 16, cex = 0.8,
     main = "Practice Map")
```

After creating your map, answer:

Where are the highest values located?
Do you see any spatial clustering?
If management was uniform, what might explain this pattern?