---
title: "Exercise: SSURGO"
output: html_document
---

# Introduction
In this lesson, we saw how SSURGO can add soil
context to yield maps.

This file is a guided demonstration. The goal is
to connect yield data to SSURGO soil polygons,
then ask: do soil types explain yield differences?

We start by loading the field yield shapefile.

```{r}
library(sf)

field <- st_read("data/yield shape.shp", quiet = TRUE)
head(field)
dim(field)
```

Before going further, check the CRS. Every spatial
analysis should start here. It determines whether
distances and areas are meaningful.

```{r}
st_crs(field)
st_is_longlat(field)
```

If `st_is_longlat()` returns `TRUE`, coordinates are
in degrees. We will need to reproject before any
distance-based operation.


## Quick Yield Map

First, make a simple yield map with base plotting.

```{r}
plot(field["Yld_Vol_Dr"], pch = 16, cex = 0.5,
     pal = function(n){hcl.colors(n, 'Temps', rev = TRUE)},
     breaks = 'quantile',
     main = "Yield Map")
```


## Load SSURGO Tables

Now pull SSURGO tables directly using
`apsimx::get_ssurgo_tables()`.

We create a field polygon from yield points and
use it as the area of interest.

```{r}
library(apsimx)

field_poly <- st_convex_hull(st_union(field))
plot(field_poly)
field_poly <- st_transform(field_poly, 4326)
field_poly <- st_as_sf(field_poly)

soil_tables <- get_ssurgo_tables(aoi = field_poly)
names(soil_tables)
```


## Explore the Spatial Layer

`get_ssurgo_tables()` returns a list. One element is
`mapunit.shp`, the actual spatial polygons for each
map unit. Let's look at it first before touching any
attributes.

```{r}
mapunit_shp <- soil_tables$mapunit.shp
class(mapunit_shp)
nrow(mapunit_shp)
head(mapunit_shp)
```

Before doing anything with the polygons, check
that they're valid. Invalid geometries (self-
intersecting rings, unclosed boundaries) will cause
silent failures in spatial operations.

```{r}
st_is_valid(mapunit_shp)

# If any are FALSE, repair them
mapunit_shp <- st_make_valid(mapunit_shp)
```

This might be a no-op here, but get in the habit.
you will hit invalid geometries in real data.

Each row is one map unit polygon. The only attribute
here is `mukey`, a unique ID that links polygons to
the tabular SSURGO tables.

Plot the raw polygons to see how the field is divided:

```{r}
# Reproject field boundary to match mapunit CRS for overlay
field_poly_match <- st_transform(field_poly, st_crs(mapunit_shp))

plot(st_geometry(mapunit_shp), border = "black", lwd = 1.5,
     main = "SSURGO Map Unit Polygons")

plot(st_geometry(field_poly_match), border = "firebrick",
     lwd = 2,
     add = TRUE)

legend("topright", legend = c("Map units", "Field boundary"),
       col = c("black", "firebrick"), lwd = c(1.5, 2), bty = "n")
```

How many distinct map units are there, and what are
they called?

```{r}
mapunit_names <- soil_tables$mapunit[, c("mukey", "muname")]
mapunit_names
```

Note the coarse resolution: a handful of polygons
covers the entire field. Yield monitor data may have
thousands of points at sub-meter density; SSURGO
rarely has more than 5–10 polygons for a typical
farm field.

Think about what this means: many yield points will
fall inside the same polygon and share identical
SSURGO values. All within-polygon variability in
yield is invisible to SSURGO.

## Add Map Unit Labels to the Map

We can label each polygon by name. First get centroids:

```{r}
centroids <- st_centroid(mapunit_shp)
coords <- st_coordinates(centroids)
labels <- mapunit_names$muname[match(mapunit_shp$mukey, mapunit_names$mukey)]

plot(st_geometry(mapunit_shp), border = "black", lwd = 1.5,
     main = "Map Units: Named")
text(coords[, 1], coords[, 2], labels = labels,
     cex = 0.6, col = "navy")
```

Note: a centroid can fall outside its own polygon
if the shape is irregular or concave. That's fine
for labels, but never use it instead of an actual
point-in-polygon test in analysis.

Each name typically includes the soil series and slope
class (e.g., "Tama silty clay loam, 2–5%").

## Build a Soil Spatial Layer

SSURGO stores data across three tables, so we need
to pick one representative value per map unit and
then merge it back to the spatial layer.

We do this in three steps.

**Step 1:** Keep only the dominant component per
map unit (there can be multiple soil types within one
map unit; we take the one with the largest `comppct.r`):

```{r}
component <- soil_tables$component[, c("mukey", "cokey", "comppct.r")]
names(component)[3] <- "comppct"
component
component <- component[order(component$mukey, -component$comppct), ]
component <- component[!duplicated(component$mukey), ]
component
nrow(component)
head(component)
```

**Step 2:** Keep topsoil horizon values (lowest
`hzdept.r` = shallowest depth) for each component:

```{r}
chorizon <- soil_tables$chorizon[, c("cokey", "hzdept.r",
                                     "om.r", "claytotal.r",
                                     "silttotal.r", "sandtotal.r",
                                     "cec7.r", "ph1to1h2o.r",
                                     "awc.r", "ksat.r",
                                     "dbthirdbar.r")]
names(chorizon)[2] <- "hzdept"
chorizon
chorizon <- chorizon[order(chorizon$cokey, chorizon$hzdept), ]
chorizon <- chorizon[!duplicated(chorizon$cokey), ]
nrow(chorizon)
head(chorizon)
```

**Step 3:** Merge everything and attach to the spatial
polygons:

```{r}
soil_data <- merge(mapunit_names, component, by = "mukey", all.x = TRUE)
soil_data <- merge(soil_data, chorizon, by = "cokey", all.x = TRUE)

complete_soil_data <- merge(mapunit_shp, soil_data, by = "mukey", all.x = TRUE)
complete_soil_data
complete_soil_data <- st_transform(complete_soil_data, 4326)

head(complete_soil_data[, c("muname", "comppct", "om.r", "claytotal.r")])
```


## Map SSURGO Variables

### Organic Matter

```{r}
plot(complete_soil_data["om.r"],
     border = "grey60",
     main = "SSURGO Organic Matter (%)")
```

### Clay Content

Use the same pattern as above, just swap the column:

```{r}
plot(complete_soil_data["claytotal.r"],
     border = "grey60",
     main = "SSURGO Clay (%)")
```


## Match Yield and Soil

Now we connect the two datasets. We use `st_within`
as the join predicate: each yield point is matched
to the polygon that completely contains it.

```{r}
field <- st_transform(field, st_crs(complete_soil_data))
yield_soil <- st_join(field, complete_soil_data, join = st_within)

head(yield_soil[, c("Yld_Vol_Dr", "mukey", "muname", "om.r", "claytotal.r")])
```

How many yield points fall in each map unit?

```{r}
table(yield_soil$mukey)
```

Do the counts look equal? Unequal point density
matters. Your average from 50 points is less
certain than your average from 5,000.

Each yield point now carries `mukey`, `muname`, and
soil properties alongside its yield value.


## Aggregate Yield by Soil Map Unit

SSURGO is coarse, so many yield points pile into
the same soil polygon. We can average the points
within each map unit. Use `mukey` (the unique ID)
instead of `muname` (the label). Names can
repeat across different surveys.

```{r}
yield_by_mu <- aggregate(Yld_Vol_Dr ~ mukey,
                         data = yield_soil,
                         FUN = function(x) mean(x, na.rm = TRUE))
yield_by_mu
```

Now stick those averages back on the soil map. Join
by `mukey`, so the legend still has the soil names:

```{r}
soil_yield <- merge(complete_soil_data, yield_by_mu,
                    by = "mukey", all.x = TRUE)

plot(soil_yield["Yld_Vol_Dr"],
     border = "grey60",
     main = "Average Yield by Soil Map Unit")
```

**Interpret the map.** Consider:

- Which map units have the highest average yield?
- Are the differences between map units large or
  small relative to total yield variability?
- Does soil type seem to explain most of the yield
  pattern, or are there within-polygon differences
  that SSURGO misses entirely?
- If SSURGO fully explained yield, what spatial
  pattern would you expect to see? Would yield be
  uniform within each polygon, with sharp changes
  at polygon boundaries?


## Within-Unit Variability

The average yield per soil map unit tells you
whether soil types differ on average. But SSURGO
is coarse. Means hide all the yield variability
*inside* each polygon. Add the standard deviation:

```{r}
yield_sd_by_mu <- aggregate(Yld_Vol_Dr ~ mukey,
                            data = yield_soil,
                            FUN = function(x) sd(x, na.rm = TRUE))
yield_sd_by_mu
```

Consider:

- Is within-unit variability large compared to the
  between-unit differences you just mapped?
- If so, what does that tell you about SSURGO's
  ability to explain field-level yield patterns?


## Practice

Use the same pattern as the clay plot above to map
one or more of the following. Choose the ones that
show the most variation across map units:

- `silttotal.r`: silt content (%)
- `sandtotal.r`: sand content (%)
- `cec7.r`: cation exchange capacity (cmol/kg)
- `ph1to1h2o.r`: soil pH
- `awc.r`: available water capacity (cm/cm)
- `ksat.r`: saturated hydraulic conductivity (µm/s)
- `dbthirdbar.r`: bulk density at field capacity (g/cm³)

Template:

```{r}
plot(complete_soil_data["silttotal.r"],
     border = "grey60",
     main = "Available Water Capacity")
```




