---
title: "Exercise: Rasters"
output: html_document
---

# Introduction
In the lecture, we saw that rasters are grids:
each cell holds one value. In this exercise,
you'll build a raster from point data using kriging,
then turn it into a management map.

# Case Study: Soil Test Data
You'll build a lime recommendation map from
data collected in Minnesota.


## Load and Filter Data
```{r}
library(sf)
library(gstat)
library(stars)


point_data <- st_read("data/Folie N & SE_grid_sample.shp", quiet = TRUE)
head(point_data)
```

Each row is one sampled location. They're
POINT geometries.

Our second shapefile describes the field boundary.
Without this boundary, we would not know the shape
or spatial extent of our raster.

```{r}
boundary <- st_read("data/Folie N &SE_boundary.shp", quiet = TRUE)

head(boundary)
```

The boundary is a `POLYGON` layer with one row.

Now pull out buffer pH (`Bph`):

```{r}
selected_data <- point_data[point_data$attribute == "Bph", ]
```

Before interpolating, spot-check where the points
are with a quick map.

```{r}
plot(st_geometry(boundary), col = "grey95", border = "grey70", reset = FALSE)
plot(st_geometry(selected_data), add = TRUE, pch = 16, cex = 0.8, col = "steelblue")
title("Buffer pH Sampling Points")
```


## Build a Grid
To make a raster, first create a grid template,
then clip it to the field boundary.

```{r}
boundary <- st_transform(boundary, 3857)
selected_data <- st_transform(selected_data, 3857)

st_bbox(boundary)
```

Build and crop the grid:

```{r}
grd <- st_as_stars(st_bbox(boundary),
                   dx = 10,
                   dy = 10)
plot(grd)
grd <- st_crop(grd, boundary)
plot(grd)
```

Plot the grid template.
```{r}
plot(st_geometry(boundary), col = "grey95", border = "black", reset = FALSE)
plot(st_geometry(st_as_sf(grd)), add = TRUE, border = "black", col = NA)
title("Grid Clipped to Field Boundary")
```
Cells are tiny, so you won't see them clearly, but
you can see the grid matches the field footprint.

## Build the Semivariogram
Three steps:

1) Calculate empirical semivariance (`variogram()`).
2) Fit a model (`fit.variogram()`).
3) Plot together to check the fit.

```{r}

v <- variogram(measure ~ 1, selected_data)
v

# "Sph" is the spherical variogram model.
m <- fit.variogram(v, vgm("Sph"))
plot(v, model = m)
```
Good if the model tracks the empirical points
before you move to kriging.

## Kriging
Kriging uses the variogram to estimate values for
the unsampled grid cells.

`krige()` needs four things:

1) formula (`measure ~ 1`)
2) point data (`selected_data`)
3) prediction grid (`grd`)
4) variogram model (`m`)


```{r}
kriged_data <- krige(formula = measure ~ 1,
                     selected_data, 
                     grd, 
                     model = m)
```

The predicted surface is the first layer.

```{r}
finished_data <- kriged_data[1]
```

## Mapping Kriged Data
Plot the interpolated buffer pH surface.

```{r}
plot(finished_data["var1.pred"], reset = FALSE,
    col = hcl.colors(12, "RdYlGn"), main = "Predicted Buffer pH (Kriging)")
plot(st_geometry(boundary), add = TRUE, border = "black", col = NA, lwd = 1)
plot(st_geometry(selected_data), add = TRUE, pch = 16, cex = 0.5, col = "grey20")
```

Warmer colors = higher predicted values. The
boundary and points let you compare predictions
against your sample pattern.

## Building the Recommendation

Make a lime rate from interpolated buffer pH.
University of Minnesota guidance:

$$rate = 37000 - bpH \times 5000$$

https://extension.umn.edu/liming/lime-needs-minnesota

Predicted values are in `var1.pred`.

```{r}
finished_data$rec <- (37000 - finished_data$var1.pred * 5000)
```

Set rates to zero where buffer pH is high (above
6.8, which means the recommendation falls below
3000 lb/ac):

```{r}
finished_data$rec <- ifelse(finished_data$rec < 3000, 0, finished_data$rec)
```

## Mapping the Recommendation
Plot the recommendation surface.

```{r}
plot(finished_data["rec"], reset = FALSE,
    col = hcl.colors(12, "RdYlGn"), main = "Recommended Lime Rate")
plot(st_geometry(boundary), add = TRUE, border = "black", col = NA, lwd = 1)
```

This converts predicted buffer pH into action:
higher zones get lime applied.

# Practice 1
Build a soil K raster from the same field.

1) Change to `K`.
```{r}
selected_data <- point_data[point_data$attribute == "K", ]

boundary <- st_transform(boundary, 3857)
selected_data <- st_transform(selected_data, 3857)
```

2) Create the raster framework.
```{r}
grd <- st_as_stars(st_bbox(boundary))
grd <- st_crop(grd, boundary)

```

3) Build and inspect the variogram.
```{r}
v <- variogram(measure ~ 1, selected_data)
m <- fit.variogram(v, vgm("Sph"))
plot(v, model = m)
```

4) Krige and extract predicted values.
```{r}
kriged_data <- gstat::krige(formula = measure ~ 1, selected_data, grd, model = m)

finished_data <- kriged_data[1]
```

5) Map soil K values.
```{r}
plot(finished_data["var1.pred"], reset = FALSE,
    col = hcl.colors(12, "RdYlGn"), main = "Predicted Soil K (Kriging)")
plot(st_geometry(boundary), add = TRUE, border = "black", col = NA, lwd = 1)
plot(st_geometry(selected_data), add = TRUE, pch = 16, cex = 0.5, col = "grey20")
```

Map reading tip: compare high and low prediction
zones against sample locations.

6) Calculate K recommendation rates for a 200 bu/ac
yield goal.

Use: K recommendation = $(1.12 - 0.0056 \times K) \times 200$

Set to zero where K > 200 ppm:

```{r}
finished_data$rec <- (1.12 - 0.0056 * finished_data$var1.pred) * 200
finished_data$rec <- ifelse(finished_data$var1.pred > 200, 0, finished_data$rec)
```

7) Plot recommended rates.
```{r}
plot(finished_data["rec"], reset = FALSE,
    col = hcl.colors(12, "RdYlGn"), main = "Recommended K Rate")
plot(st_geometry(boundary), add = TRUE, border = "black", col = NA, lwd = 1)
```

Map reading tip: this map shows where K
application is expected to be most justified.




# Practice 2
Build a pH raster. Same process, just change
`selected_data` to `Ph`. No recommendation
layer needed.

1) Filter for pH.
```{r}
selected_data <- point_data[point_data$attribute == "Ph", ]
selected_data <- st_transform(selected_data, 3857)
```

2) Build grid.
```{r}
grd <- st_as_stars(st_bbox(boundary))
grd <- st_crop(grd, boundary)
```

3) Variogram.
```{r}
v <- variogram(measure ~ 1, selected_data)
m <- fit.variogram(v, vgm("Sph"))
plot(v, model = m)
```

4) Krige and extract predicted values.
```{r}
kriged_data <- gstat::krige(formula = measure ~ 1, selected_data, grd, model = m)

finished_data <- kriged_data[1]
```

5) Map soil pH values.
```{r}
plot(finished_data["var1.pred"], reset = FALSE,
    col = hcl.colors(12, "RdYlGn"), main = "Predicted Soil pH (Kriging)")
plot(st_geometry(boundary), add = TRUE, border = "black", col = NA, lwd = 1)
plot(st_geometry(selected_data), add = TRUE, pch = 16, cex = 0.5, col = "grey20")
```

Map reading tip: this is a predicted surface, not
direct observations at every location.

