---
title: "Exercise: Rasters"
output: html_document
---

# Introduction

This exercise builds a raster-based interpolation
workflow from point samples using variograms and
kriging, then maps the interpolated surface.

# Case Study: Soil Test Mapping

## Load and Filter Data

```{r}
library(tidyverse)
library(gstat)
library(sp)
library(raster)
library(stars)
library(leaflet)
library(leafem)

point_data = st_read("data/Folie N & SE_grid_sample.shp", quiet=TRUE)
head(point_data)
```

```{r}
boundary = st_read("data/Folie N &SE_boundary.shp", quiet=TRUE)

head(boundary)
```

```{r}
selected_data = point_data %>%
  filter(attribute=="Bph")
```

```{r}
selected_data %>%
  leaflet() %>%
  addCircleMarkers() %>%
  addProviderTiles(provider = providers$Esri.WorldImagery)
```

## Build the Grid

```{r}
library(stars)

boundary = st_transform(boundary, 3857)
selected_data = st_transform(selected_data, 3857)

st_bbox(boundary)
```

```{r}
grd = st_bbox(boundary) %>%
  st_as_stars() %>%
  st_crop(boundary)

```

```{r}
grd %>%
  leaflet()  %>%
  addProviderTiles(provider = providers$Esri.WorldImagery) %>%
  addStarsImage()
```

## Build Semivariogram

```{r}
v = variogram(measure~1, selected_data)

m = fit.variogram(v, vgm("Sph"))
plot(v, model = m)
```

## Kriging

```{r}
library(gstat)

kriged_data = gstat::krige(formula = measure~1, selected_data, grd, model=m)


```

```{r}
finished_data = kriged_data[1]
```

## Map Kriged Output

```{r}
library(RColorBrewer)

finished_data %>%
  leaflet() %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addStarsImage(opacity = 0.5,
                colors="RdYlGn")
```

```{r}
finished_data %>%
  leaflet() %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addStarsImage(opacity = 0.5,
                colors="RdYlGn",
                group = "nutrients") %>%
  addImageQuery(finished_data,
                group = "nutrients")
```

## Build Recommendation Layer

```{r}
finished_data$rec = (37000 - finished_data$var1.pred*5000)
```

```{r}
finished_data$rec = if_else(finished_data$rec<3000, 0, finished_data$rec)
```

```{r}
leaflet(finished_data["rec"]) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addStarsImage(opacity = 0.5,
                colors = "RdYlGn")
```

```{r}
leaflet(finished_data["rec"]) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addStarsImage(opacity = 0.5,
                colors = "RdYlGn") %>%
  addImageQuery(finished_data["rec"],
                type = "click")
```

# Practice

Create a soil K raster using the same workflow.

1. Change selected attribute from `"Bph"` to `"K"`.

```{r}
selected_data = point_data %>%
  filter(attribute=="Bph")

boundary = st_transform(boundary, 3857)
selected_data = st_transform(selected_data, 3857)
```

2. Create raster framework.

```{r}
### make grid
grd = st_bbox(boundary) %>%
  st_as_stars() %>%
  st_crop(boundary)
```

3. Build and inspect variogram.

```{r}
v = variogram(measure~1, selected_data)
m = fit.variogram(v, vgm("Sph"))
plot(v, model = m)
```
