---
title: "Sample Statistics"
subtitle: "AGRON 5130"
author: "YOUR NAME HERE"
output: pdf_document
---


# Introduction

The goal of this assignment is to give you
hands-on practice working with **sample
statistics** and
**confidence intervals**, building directly on the
concepts introduced in this week’s videos. You
will work with real agronomic datasets, compute
sample-based summary statistics, quantify
uncertainty using the standard error, and
construct confidence intervals using the *t*
distribution.

Submit a knitted PDF or HTML generated from your
R Markdown file to Canvas.


# Question 1

**1 point**

Load the dataset
`"data/millet_uniformity.csv"`
and examine the first six rows using the `head()`
function.

Pearl millet yield was sampled at 160 locations
within a field in Kundewadi, Niphad, in the district
of Nasik, India.

```{r}
millet <- read.csv('./data/millet_uniformity.csv')
head(millet)
```


# Question 2

**1 point**

Calculate the **mean**, **standard deviation**, and
**number of observations** for millet yield.

```{r}
xbar <- mean(millet$yield)
s <- sd(millet$yield)
n <- length(millet$yield)
xbar
s
n
```


# Question 3

**1 point**

Calculate the **standard error of the mean** for
millet yield.

```{r}
se <- s / sqrt(n)
se
```


# Question 4

**2 points**

Calculate the *t*-value (using the `qt()` function)
needed to construct a **95% confidence interval**
for the mean yield.

A 95% confidence interval leaves 2.5% of the
probability in each tail, so use `0.975`. Remember
that the degrees of freedom are one less than the
number of observations.

```{r}
t.val <- qt(p = c(0.025, 0.975),
            df = n - 1 )

t.val
```


# Question 5

**2 points**

Calculate the **95% confidence interval** for millet
yield using the statistics computed above.

```{r}
ci <- xbar + (t.val * se)
ci
```


# Question 6

**1 point**

Load the dataset
`"data/safflower_uniformity.csv"`
and examine the first six rows using the `head()`
function.

Safflower yield was sampled 340 times within a
safflower field in Eloy, Arizona.

```{r}
saff <- read.csv('./data/safflower_uniformity.csv')
head(saff)
```


# Question 7

**1 point**

Calculate the **mean**, **standard deviation**, and
**number of observations** for safflower yield.

```{r}
xbar <- mean(saff$yield)
s <- sd(saff$yield)
n <- length(saff$yield)
xbar
s
n
```


# Question 8

**2 points**

Calculate the **standard error of the mean** for
safflower yield.

```{r}
se <- s / sqrt(n)
se
```


# Question 9

**2 points**

Calculate the *t*-value (using the `qt()` function)
needed to construct a **90% confidence interval**
for the mean safflower yield.

A 90% confidence interval leaves 5% of the
probability in each tail, so use `0.95`.

```{r}
t.val <- qt(p = c(0.05, 0.95),
            df = n - 1 )
t.val
```


# Question 10

**2 points**

Calculate the **90% confidence interval** for
safflower yield using the statistics computed above.

```{r}
ci <- xbar + (t.val * se)
ci
```
