---
title: "Messy and Missing Data"
subtitle: "AGRON 5130 - Module 9"
author: "Caio dos Santos"
date: "2026-03-11"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    smooth: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Introduction

Up to this point in the course, most datasets we
used were clean and well-behaved.

Real agronomic data are often not.

In this unit, we focus on what to do **before**
we trust a model:

- inspect distributions
- inspect variance patterns
- identify suspicious observations
- decide how to handle missing data

The goal is not to make data look perfect. The
goal is to understand whether our inferences are
credible.

# Why This Matters

Most tests we have used rely on assumptions:

- residuals are approximately normal
- variances are comparable across treatments
- errors are independent

When these assumptions fail, p-values and
confidence intervals can be misleading.

# Inspecting Data for Normal Distributions

## Histograms

A histogram is the fastest first check. A roughly
normal dataset should be symmetric around its
center.

```{r}
set.seed(20260311)
normal_yield <- rnorm(500, mean = 180, sd = 4)
skewed_biomass <- rgamma(1000, shape = 2.5, scale = 2.2)

par(mfrow = c(1, 2))
hist(normal_yield,
     main = "Corn Yield (Approx. Normal)",
     xlab = "Yield",
     col = "grey80",
     border = "white")
hist(skewed_biomass,
     main = "Weed Biomass (Right-Skewed)",
     xlab = "Biomass",
     col = "grey80",
     border = "white")
par(mfrow = c(1, 1))
```

A skewed histogram is not automatically wrong, but
it tells us we may need a transformation or a
different model family.

## Rank Percentile Plots

Rank-percentile plots help us see tail behavior.
Strong curvature in upper or lower tails is a red
flag for non-normality.

```{r}
rank_plot <- function(x, title_text) {
  p <- (rank(x, ties.method = "average") - 0.5) / length(x)
  plot(p, x,
       pch = 21,
       bg = "grey60",
       xlab = "Percentile Rank",
       ylab = "Observed Value",
       main = title_text)
  abline(lm(x ~ p), col = "tomato", lwd = 2)
}

par(mfrow = c(1, 2))
rank_plot(normal_yield, "Rank Plot: Normal")
rank_plot(skewed_biomass, "Rank Plot: Skewed")
par(mfrow = c(1, 1))
```

## Box Plots

Box plots are useful for quickly spotting extreme
values and comparing spread among groups.

```{r}
trt <- rep(c("A", "B", "C"), each = 40)
y <- c(rnorm(40, 170, 5), rnorm(40, 172, 5), c(rnorm(39, 171, 5), 205))
boxplot(y ~ trt,
        col = c("grey85", "grey75", "grey65"),
        ylab = "Yield",
        xlab = "Treatment",
        main = "Box Plot by Treatment")
```

# Inspecting Equal Variance

## Mean-Variance Pattern

When variance increases with the mean, we have
heteroscedasticity. This is common in count and
biomass data.

```{r}
set.seed(20260312)
groups <- factor(rep(1:8, each = 30))
g_means <- seq(20, 90, length.out = 8)
z <- unlist(lapply(g_means, function(m) rnorm(30, m, m * 0.12)))

mv <- aggregate(z, list(groups), function(v) c(mean = mean(v), var = var(v)))
mv <- do.call(data.frame, mv)
colnames(mv) <- c("group", "mean", "var")

plot(mv$mean, mv$var,
     pch = 19,
     xlab = "Group Mean",
     ylab = "Group Variance",
     main = "Mean-Variance Plot")
abline(lm(var ~ mean, data = mv), col = "tomato", lwd = 2)
```

## Homogeneity Tests

Formal tests (e.g., Bartlett or Levene) can help,
but should never replace visual diagnostics.

```{r}
bartlett.test(z ~ groups)
```

# Dealing with Messy Data

## Outliers

Outliers can be:

- true biological extremes
- measurement or data-entry errors
- symptoms of model misspecification

Outliers should not be dropped automatically.
Document your reasoning and, when possible, report
analyses with and without influential points.

## Non-normality and Unequal Variance

Common options include:

- transformation (`log`, `sqrt`, Box-Cox)
- variance-aware models
- robust methods

```{r}
log_biomass <- log(skewed_biomass + 1)
par(mfrow = c(1, 2))
hist(skewed_biomass, main = "Original", xlab = "Biomass", col = "grey80", border = "white")
hist(log_biomass, main = "log(x+1)", xlab = "Transformed", col = "grey80", border = "white")
par(mfrow = c(1, 1))
```

# Dealing with Missing Data

Missing data are common in field trials:

- plot destroyed by flooding
- machine or sensor failure
- sampling error

The key question is: **why is data missing?**

- MCAR: missing completely at random
- MAR: missing at random conditional on observed data
- MNAR: missing not at random

Simple deletion may be acceptable for small,
random losses. Larger or structured missingness
usually requires explicit handling.

```{r}
set.seed(20260313)
d <- data.frame(
  trt = rep(c("A", "B"), each = 15),
  yield = c(rnorm(15, 175, 5), rnorm(15, 182, 5))
)
missing_rows <- sample(1:nrow(d), 5)
d$yield[missing_rows] <- NA

summary(d)

# Complete-case mean by treatment
aggregate(yield ~ trt, data = d, FUN = function(v) mean(v, na.rm = TRUE))
```

# Conclusion

Data cleaning is not separate from statistics. It
**is** statistics.

Before modeling, always ask:

- Are distributions reasonable?
- Are variances comparable?
- Are extreme values plausible?
- Is missingness ignorable?

Careful diagnostics protect you from false
confidence and lead to stronger agronomic
inference.
