---
title: "Confidence Interval for Sample Mean"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Introduction

In the last part of this lesson, we learned how
the **standard error** and the **t-distribution**
work together to define a **confidence interval**.
A confidence interval is a range of values
centered on the sample mean that, in a given
percentage of repeated samples, is expected to
contain the true population mean.

The confidence interval is defined by an **upper**
and **lower confidence limit**, which are
determined by the product of the appropriate
*t-value* and the *standard error of the mean*.


# Calculating the Confidence Interval

Once we know how to calculate the standard error
and obtain the appropriate t-value, constructing a
confidence interval is straightforward:

1. Multiply the standard error by the t-value  
2. Subtract this product from the sample mean to
obtain the **lower confidence limit**
3. Add this product to the sample mean to obtain
the **upper confidence limit**
4. Report the confidence interval in parentheses
as *(lower confidence limit, upper confidence limit)*


# Case Study: Peanut Sample 1

In the `data` folder, there are nine sample
datasets from the peanut uniformity trial, each
with a different sample size. Let’s start by
loading the first sample.

```{r}
peanut <- read.csv("data/peanut_sample_1.csv")
head(peanut)
```

Let’s define a new variable, `yield`, using the
yield column from the dataset.

```{r}
yield <- peanut$yield
```

First, we calculate the sample mean and sample
standard deviation.

```{r}
yield_mean <- mean(yield)
yield_sd <- sd(yield)
```

Next, we calculate the **standard error of the
mean**. Recall that the standard error is the
standard deviation divided by the square root of
the number of samples.

In this example, we know there are only two
observations, but in practice it is safer to let R
determine the sample size for us. We can do that
using the `length()` function.

```{r}
no_samples <- length(yield)
```

Now we can calculate the standard error.

```{r}
yield_se <- yield_sd / sqrt(no_samples)
```

To construct the confidence interval, we now need
the appropriate **t-value**. For this example, we
want a **90% confidence interval**. This means
that 5% of the distribution lies in each tail, so
the upper tail begins at 95%.

We can obtain the t-value using the `qt()`
function.

```{r}
t_value <- qt(0.95, 1)
```

Where does the `1` come from? This is the
**degrees of freedom**, which is equal to the
number of samples minus one. Since we have two
samples, we have one degree of freedom.

At this point, we know:

* the standard error is approximately 0.46
* the t-value is approximately 6.31

The final step is to subtract and add the product
of the standard error and t-value from the sample
mean.

```{r}
lower_limit <- yield_mean - (yield_se * t_value)
upper_limit <- yield_mean + (yield_se * t_value)

lower_limit
upper_limit
```

The resulting confidence interval is **(−1.06,
4.74)**.

Clearly, yield cannot be negative. However,
because the sample mean is close to zero and the
sample size is very small, the confidence interval
is extremely wide. This example highlights an
important point in statistics and data science:
**statistical results must always be interpreted
in the context of domain knowledge**.

Finally, whenever you report a confidence
interval, you should also report its **confidence
level** and **degrees of freedom**. In this case,
we would report:


$$CI(0.90, 1) = (-1.06,\ 4.74)$$


# Case Study: Peanut Sample 2

Let’s go through the second sample a bit more quickly.

```{r}
peanut_2 <- read.csv("data/peanut_sample_2.csv")
head(peanut_2)
```

First, we calculate the sample mean, standard
deviation, and standard error.

```{r}
yield_2 <- peanut_2$yield
yield_mean_2 <- mean(yield_2)
yield_sd_2 <- sd(yield_2)
no_samples_2 <- length(yield_2)
yield_se_2 <- yield_sd_2 / sqrt(no_samples_2)
```

This time, we want a **95% confidence interval**,
which places 2.5% of the distribution in each
tail. This means the upper tail begins at 97.5%.

We also have five samples, so the degrees of
freedom are (5 - 1 = 4).

```{r}
t_value_2 <- qt(0.975, 4)
```

Now we compute the confidence interval.

```{r}
lower_limit_2 <- yield_mean_2 - (yield_se_2 * t_value_2)
upper_limit_2 <- yield_mean_2 + (yield_se_2 * t_value_2)

lower_limit_2
upper_limit_2
```

The confidence interval for this sample is:

$$CI(0.95, 4) = (1.50,\ 2.30)$$

---

# Practice

The datasets `peanut_sample_3.csv` through `peanut_sample_9.csv` are available for you to practice constructing confidence intervals. For reference, the correct **95% confidence intervals** are listed below:

* `peanut_sample_3.csv`: (1.95, 2.41)
* `peanut_sample_4.csv`: (1.98, 2.34)
* `peanut_sample_5.csv`: (1.86, 2.65)
* `peanut_sample_6.csv`: (1.98, 2.57)
* `peanut_sample_7.csv`: (2.09, 2.42)
* `peanut_sample_8.csv`: (2.12, 2.49)
* `peanut_sample_9.csv`: (1.42, 2.15)


  