---
title: "T-Distribution"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Introduction

This week, we introduce the **t-distribution**.
The t-distribution describes the distribution of
*sample means* around their mean value. This
contrasts with the **Z-distribution**, which
describes the distribution of *individual
observations* around the population mean.

The most important difference between the
t-distribution and the Z-distribution is that the
**shape of the t-distribution depends on sample
size**. As the number of samples used to calculate
the mean increases, the t-distribution becomes
narrower and taller, gradually approaching the
standard normal distribution.


# Plotting the Distribution

In the lecture, we once again used the
`shadeDist()` function from the **fastGraph**
package. If you did not install this package in
the previous unit, you can do so using (just
remove the #):

```{r}
#install.packages("fastGraph")
```

Then, load the package into your session:

```{r}
library(fastGraph)
```

We can use `shadeDist()` to visualize how the
shape of the t-distribution changes as the number
of samples increases.

Let’s start by looking at the t-distribution for a
sample size of **2**. Since degrees of freedom are
equal to *n − 1*, this corresponds to **1 degree
of freedom**. We will shade the area between *t =
−1* and *t = 1* for reference.

```{r}
shadeDist(c(-1, 1), "dt", 1, lower.tail = FALSE)
```


Remember that `shadeDist()` requires four arguments:

1. `c(-1, 1)` specifies the range of t-values to
shade
2. `"dt"` tells R to use the t-distribution
3. The third argument is the **degrees of
freedom**, which is always one less than the
number of samples
4. `lower.tail = FALSE` tells R to shade the
*middle* of the distribution rather than the tails

Now let’s increase the sample size to **4** (3
degrees of freedom).

```{r}
shadeDist(c(-1, 1), "dt", 3, lower.tail = FALSE)
```

You should notice there's a greater probability
concentrated under the same bounds, indicating
that the distribution itself has become narrower.

If we increase the sample size to **6**, the shape
does not change as dramatically, but the
proportion of the curve between *t = −1* and *t =
1* increases.

```{r}
shadeDist(c(-1, 1), "dt", 5, lower.tail = FALSE)
```

Finally, if we increase the sample size to **30**,
the curve continues to narrow and looks much
closer to a standard normal distribution.

```{r}
shadeDist(c(-1, 1), "dt", 29, lower.tail = FALSE)
```


# Calculating t-values

So far, we have supplied `shadeDist()` with
t-values and degrees of freedom, and R has
calculated the corresponding probabilities.

We can reverse this process and calculate
**t-values directly** using the `qt()` function.
This function returns the t-value associated with
a given probability and number of degrees of
freedom.

Suppose we want to find the t-values that bound
the **middle 50%** of potential sample means when
each mean is based on **four samples** (3 degrees
of freedom).

If 50% of the distribution is in the middle, then
25% of the distribution lies below this range and
25% lies above it. We can calculate the lower
bound as follows:

```{r}
qt(0.25, 3)
```

This tells R to compute the t-value below which
25% of the distribution falls, assuming 3 degrees
of freedom.

To find the upper bound, we calculate the t-value
below which 75% of the distribution falls:

```{r}
qt(0.75, 3)
```

As expected, the lower and upper limits are
symmetric around zero—they differ only in sign.

We can verify this result using `shadeDist()` by
plugging in the t-values we just calculated.

```{r}
shadeDist(c(-0.7649, 0.7649), "dt", 3, lower.tail = FALSE)
```

The shaded region matches the target proportion.


What if we want to find the t-values that bound
the **middle 68%** of potential sample means? In
this case, 32% of the distribution lies outside
the range, with 16% in each tail.

The lower bound would be:

```{r}
qt(0.16, 3)
```

And the upper bound:

```{r}
qt(0.84, 3)
```


How about the t-values that define the range
containing **90%** of potential sample means?

```{r}
qt(c(0.05, 0.95), 3)
```

And finally, the range containing **95%** of
potential sample means:

```{r}
qt(c(0.025, 0.975), 3)
```


# Practice

Calculate the t-values that define the **middle
60%** of potential sample means if there were **10
samples**.

The lower limit is:

```{r}
qt(0.2, 9)
```

What is the upper limit?


**Q:** What is the range of t-values associated
with **70%** of potential sample means if the
number of samples is 12?
**A:** (−1.09, 1.09)

**Q:** What is the range of t-values associated
with **80%** of potential sample means if the
number of samples is 14?
**A:** (−1.35, 1.35)

**Q:** What is the range of t-values associated
with **90%** of potential sample means if the
number of samples is 17?
**A:** (−1.75, 1.75)

**Q:** What is the range of t-values associated
with **95%** of potential sample means if the
number of samples is 20?
**A:** (−2.09, 2.09)

**Q:** What is the range of t-values associated
with **99%** of potential sample means if the
number of samples is 20?
**A:** (−2.86, 2.86)


