---
title: "Standard Error"
output:
  pdf_document
---

# Introduction

In the same way that the standard deviation
describes how individual observations vary around
a population mean, the **standard error of the
mean** describes how *sample means* vary around
their mean. In other words, the standard error is
 the standard deviation of sample means.

This concept becomes especially important when we
cannot measure an entire population and instead
rely on samples to estimate population-level
statistics.


# Case Study: Tomatoes

Once again, we will work with some familiar
datasets, including tomato, barley, cotton, and
peanut yield data. We will start with the tomato
dataset.

Let’s load the data and take a look at the first
six rows.

```{r}
tomato <- read.csv("data/tomato_uniformity.csv")
head(tomato)
```

# Calculating the Standard Error

The standard error of the mean is calculated as
the population standard deviation divided by the
square root of the number of samples. To compute
it, we first need the standard deviation of the
population.

Let’s define the yield column and calculate its
standard deviation using the `sd()` function.

```{r}
yield <- tomato$yield 
tomato_sd <- sd(yield)
tomato_sd
```

Now we can divide the standard deviation by the
square root of the sample size to obtain the
standard error of the mean.

What would the standard error of the sample mean
be if we took **four samples**?

```{r}
tomato_se <- tomato_sd / sqrt(4)
tomato_se
```


We will talk more about the *t* distribution in
the next exercise, but for now we can make a
useful approximation. Roughly 95% of sample means
should fall within about **two standard errors**
of the mean of their sampling distribution.

In this case, two standard errors is approximately
10.6. That means we would expect about 95% of
sample means to fall within 50.3 ± 10.6, or
roughly between **39.7 and 60.9**.

You do not need to fully understand the code
below. Just run it and focus on the resulting
plot.

In the example below, we simulate 1,000 sample
means, each based on four randomly selected
observations from the tomato yield population. The
histogram shows the distribution of those sample
means, with the population mean shown in red.

```{r}
set.seed(20260130)
sample_list <- list()

for(i in c(1:1000)){
  samples <- mean(sample(yield, 4))
  samples <- as.data.frame(samples)
  sample_list[[i]] <- samples
}

sample_list_df <- do.call(rbind.data.frame, sample_list)
names(sample_list_df) <- "yield"

pop_mean <- mean(yield)

hist(sample_list_df$yield, breaks = 30,
     main = 'Histogram of Yield')
abline(v = pop_mean, col = 'red')
abline(v = c(39.7, 60.9), col = 'blue')

```

Most of the sample means fall within the interval
we discussed earlier (shown in blue), with only a
small proportion appearing in the tails.

What would the standard error of the sample mean
be if we took **seven samples** instead?

```{r}
tomato_se <- tomato_sd / sqrt(7)
tomato_se
```


# Practice: Barley

Now let’s practice calculating the standard error
using a different dataset.

What would the standard error of the sample mean
be if we took **eight samples** from the barley
yield population?

First, load the `"barley_uniformity.csv"` dataset
and inspect the first six rows.

```{r}
barley <- read.csv("data/barley_uniformity.csv")
head(barley)
```

Next, define the column of interest and create a
histogram of its distribution.

```{r}
barley_yield <- barley$yield
hist(barley_yield)
```

To calculate the standard error, we first need the
standard deviation of the population.

```{r}
barley_sd <- sd(barley_yield)
barley_sd
```

What would the standard error of the sample mean
be if we took **two samples**?

```{r}
barley_se <- barley_sd / sqrt(2)
barley_se
```

The standard error should be approximately
**22.0**.

Try calculating the standard error for additional
sample sizes. You should obtain values close to
the following:

* 4 samples: SE ≈ 15.6
```{r}
barley_sd / sqrt(4)
```
* 5 samples: SE ≈ 13.9
* 7 samples: SE ≈ 11.8


# Practice: Cotton

Finally, repeat this process using the
`"cotton_uniformity.csv"` dataset.

Calculate the standard errors of the sample mean
for sample sizes of **3**, **4**, and **6**. You
should obtain values close to:

* 3 samples: SE ≈ 0.13
* 4 samples: SE ≈ 0.11
* 6 samples: SE ≈ 0.09

